### 2023-06-14 Fifteen-minute-neighborhoods/Fifteen-minute-neighborhoods-title-page.Rmd
Initial commit

---
title: 'Reality or wishful thinking: Examining the feasibility of 15-minute neighborhoods from the perspective of network attributes' 
subtitle: 
titlerunning: 'Examinining fifteen minute neighborhoods from the perspective of network attributes'
authorrunning: 
thanks: | 
    This project was supported by a McMaster Institute for Research on Aging Catalyst Grant.

authors: 
- name: Antonio Paez
  address: School of Earth, Environment and Society, McMaster University
  email: paezha@mcmaster.ca

- name: James R. Dunn
  address: Department of Health, Aging and Society, McMaster University
  email: jim.dunn@mcmaster.ca

- name: Josh Arbess
  address: Mechanical Engineering and Society, McMaster University
  email: arbessj@mcmaster.ca

keywords:
- Walking
- Accessibility
- 15-minute neighborhood

#PACS: 
#- PAC1
#- superPAC

#MSC:
#- MSC code 1
#- MSC code 2

abstract: |
  Walking has been the primary form of personal transportation for as long as humans have been humans. From prehistoric migration to the dawn of populated settlements, the space-time constraints of walking have fundamentally determined how far interactions with the environment and other humans were feasible. It took tens of thousands of years for humans to walk out from their original African niche to eventually settle every nook of the planet.

  The advent of permanent settlements imposed stronger space-time constraints on people’s movement than had hitherto been the case. Having to return to a fixed abode or to go to a designated place to drudge or trade meant that people become more local of necessity. Despite the invention of the wheel and the use of animal power for transportation purposes, humanity remained for the most part pedestrian for millennia [@roberts1998short]. The physical effort involved in walking limited the geographical extent of settlements and in turn settlements evolved to facilitate walking.

  The dominance of travel by foot faded within the span of a few short decades. Several revolutions contributed to this. Technological innovations in the 19th and 20th centuries led to the internal combustion engine, smooth paving surfaces, and systems to move fuel. Concurrently, socio-technological innovations (e.g., Fordism) created the basis for mass production and consumption. The pioneer in this respect was of course the auto industry; many decades later, whole economic systems are still dominated by this sector (Jane Jacobs famously quipped that “[t]he purpose of life is to produce and consume automobiles.”) Early in the automotive era cars were seen as an ideal solution to many urban ills [@brown2009planning], which contributes to explain their enthusiastic reception. In the second part of the 20th century, motorized mobility rose to become the dominant form of transportation in cities and regions around the world. The automobile replaced walking as the key determinant of how far settlements could grow, and cities grew to accommodate this form of transportation, often with unfortunate single-mindedness.

  The results of this process have been calamitous.

  Reliance on motorized mobility contributes to pollution, climate change, the erosion of social capital and sense of community, road fatalities, and poor health due to sedentary lifestyles. For years now, work has aimed to grow a consensus about the importance of communities that better serve all their residents, and not only their vehicles. The 15-minute neighborhood [15MN; @pozoukidou2021fifteen] is among a handful of ideas that strive to emphasize movement at a human scale, in environments that accommodate a wide range of capabilities throughout the lifespan, and with the aim of improving livability and health in ways that automobility can no longer promise, let alone deliver.

  A challenge faced by 15MNs is the legacy of decades of auto-centric planning. Streetscapes are key parts of the hardware of cities, not only for what is evident at surface level (e.g., sidewalks, pavements), but also due to other accessory yet hidden infrastructure, both physical (e.g., water, sewage, power) and social (e.g., property rights, right of ways). Suburban developments in North America are often implicitly or explicitly designed to discourage through traffic. This is done by creating predominantly single use landscapes with meandering, poorly connected roads. Alas, this form of development cuts both ways, since the high built-in cost of navigation does not discriminate between outsiders and residents.

  A relevant concern is identifying types of streetscapes that can plausibly support 15MNs. To address this question, we investigate the current accessibility situation in parts of Canada’s largest metropolitan region. The analysis consists of two parts, with positive and normative aspects 15-minute walking neighborhoods are studied, and their accessibility levels assessed (positive analysis). Maximal opportunity landscapes are then used to simulate equivalent opportunity landscapes throughout the region. Accessibility is then reanalyzed from the normative perspective of the provision of opportunities. The results of this analysis are finally correlated to neighborhood network attributes, including connectivity, centrality, and clustering. The results of this investigation provide valuable information about neighborhoods, their morphology and potential to support the aspirational goal of providing opportunities within 15-minute walks for their residents. This information can help to identify target neighborhoods for planning interventions, as well as neighborhoods for whom the 15-minute ideal could be little more than wishful thinking.

bibliography: bibliography.bib
biblio-style: spphys
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

<!--
@knight2018walkable
@liu2022toward
@pozoukidou2021fifteen
@weng2019fifteen

 Walking has naturally, at an evolutionary level, been the primary form of personal transportation for as long as humans have been humans. From every prehistoric migration to the dawn of populated settlements, the space-time constraints of walking have fundamentally determined how far interactions with the environment and other humans were feasible. It took tens of thousands of years for humans to walk out from their original niche in the African savannah to eventually colonize, and settle in, every nook of the planet.
    The advent of permanent human settlements at some point in the Neolithic period imposed stronger space-time constraints on people’s movement than had hitherto been the case. The need to return to a fixed abode or to go to a designated place to drudge or trade meant that there was only so much space that a person could trade for time, while still being able to toil for life, before even enjoying leisure and recreation. Despite the invention of the wheel by the Babylonians, and the use of animal power for transportation purposes, humanity remained for the most part pedestrian for millennia. Animal power was a rarity—an expensive and therefore exclusive rarity: by some estimates, horse ownership in the US in the 18th century was dominated by white landowners. The physical effort involved in walking necessarily limited the geographical extent of settlements. In a very real way, walking determined how far settlements could grow, and in turn settlements evolved that facilitated walking. 
    From a historical perspective, the dominance of movement by foot came to a screeching halt all within the span of a few decades. Several revolutions contributed to this change, starting in the 19th century, but accelerating and possibly reaching their peak in the late 20th or early 21st centuries. Among these were technological innovations in materials and energy that led to the internal combustion engine, smooth paving surfaces, and systems to transport fuels across vast distances to enable many other forms of transportation. At the same time, there were important socio-technological innovations such as Fordism, which created the basis for modern industrial and labor-economic systems to sustain mass production and consumption, initially of automotive vehicles and still, many decades later, still dominated by the auto industry supply chain complex (Jane Jacobs famously quipped that “[t]he purpose of life is to produce and consume automobiles.”) Early in the automotive era cars were seen as an ideal solution to many urban ills (e.g., congestion, manure); in the second part of the 20th century, motorized mobility rose to become in many places the dominant form of transportation. The automobile replaced walking as the determinant of how far settlements could grow, and as a matter of policy cities decided to grow to accommodate this form of transportation, often with an unfortunate single-minded focus.
    The results of this process have been nothing short of disastrous.
    The impacts of reliance on motorized mobility are anywhere we look, from pollution, the contributions of the transportation sector to climate change, the erosion of social capital and sense of community in the places where we live, road fatalities, and the public health implications of sedentary lifestyles. For at least two decades now, planning, transportation, and public health professionals and scholars alike have been working towards a consensus on the importance of creating and fostering communities that better serve all their residents, and not only their vehicles. A handful of ideas have been proposed that aim to meet this objective, including transit-oriented development, transit villages, aging-friendly cities (alternatively neighborhoods), and more recently the 15- (or sometimes 5-, 20- or 30-) minute city (or neighborhood). These concepts all share some commonalities: movement happens at a more human scale, in a way that accommodates capacities throughout the lifespan, and that contributes to quality of life and health in ways that automobility can no longer promise, let alone deliver.
     A challenge faced by proposals such as the 15-minute city is the legacy of decades of auto-centric urban planning. Streetscapes are a firm element of the hardware of cities, and not only for what is evident at surface level (e.g., sidewalks, pavements), but also due to all the accessory but hidden infrastructure, both physical (e.g., water, sewage, power) and social (e.g., property rights, right of ways). In North America, suburban developments (particularly in the post-war period), were implicitly or explicitly designed to discourage through-traffic. This was done by creating predominantly single use suburbs with meandering, poorly connected roads. Alas, this form of development cuts both ways as the built-in high cost of navigation does not discriminate between outsiders and residents. A relevant question therefore concerns the kinds of streetscapes that can in principle support 15-minute neighborhoods.
    To address this question, we investigate the current accessibility situation in parts of Canada’s major metropolitan region. The analysis consists of two parts, with positive and normative characters. 15-minute walking neighborhoods are studied, and their accessibility levels assessed (positive analysis). Optimal opportunity landscapes are then used to simulate equivalent opportunity landscapes throughout the region. Accessibility is then reanalyzed from the normative perspective of the provision of opportunities. The results of this analysis are finally correlated to neighborhood network attributes, including connectivity, centrality, and clustering.
    The results of this investigation provide valuable information about neighborhoods, their morphology and potential to support the aspirational goal of providing opportunities within 15-minute walks for their residents. This information can help to identify target neighborhoods for planning interventions, as well as neighborhoods for whom the 15-minute ideal could be little more than wishful thinking.

-->

## References

### 2023-06-14 Fifteen-minute-neighborhoods/Fifteen-minute-neighborhoods-title-page.tex
Initial commit

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\begin{document}

\title{Reality or wishful thinking: Examining the feasibility of
15-minute neighborhoods from the perspective of network
attributes \thanks{This project was supported by a McMaster Institute
for Research on Aging Catalyst Grant.} }

    \titlerunning{Examinining fifteen minute neighborhoods from the
perspective of network attributes}

\author{  Antonio Paez \and  James R. Dunn \and  Josh Arbess \and  }

\institute{
        Antonio Paez \at
     School of Earth, Environment and Society, McMaster University \\
     \email{\href{mailto:paezha@mcmaster.ca}{\nolinkurl{paezha@mcmaster.ca}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        James R. Dunn \at
     Department of Health, Aging and Society, McMaster University \\
     \email{\href{mailto:jim.dunn@mcmaster.ca}{\nolinkurl{jim.dunn@mcmaster.ca}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Josh Arbess \at
     Mechanical Engineering and Society, McMaster University \\
     \email{\href{mailto:arbessj@mcmaster.ca}{\nolinkurl{arbessj@mcmaster.ca}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Walking has been the primary form of personal transportation for as long
as humans have been humans. From prehistoric migration to the dawn of
populated settlements, the space-time constraints of walking have
fundamentally determined how far interactions with the environment and
other humans were feasible. It took tens of thousands of years for
humans to walk out from their original African niche to eventually
settle every nook of the planet.

The advent of permanent settlements imposed stronger space-time
constraints on people's movement than had hitherto been the case. The
need to return to a fixed abode or to go to a designated place to drudge
or trade meant that people become more local of necessity. Despite the
invention of the wheel and the use of animal power for transportation
purposes, humanity remained for the most part pedestrian for millennia
(Roberts 1998). The physical effort involved in walking limited the
geographical extent of settlements and in turn settlements evolved to
facilitate walking.

The dominance of travel by foot faded within the span of a few decades.
Several revolutions contributed to this. Technological innovations in
the 19th and 20th centuries led to the internal combustion engine,
smooth paving surfaces, and systems to transport. Concurrently,
socio-technological innovations (e.g., Fordism) created the basis for
mass production and consumption. The pioneer in this respect was of
course the auto industry; many decades later, whole economic systems are
still dominated by this sector (Jane Jacobs famously quipped that
``{[}t{]}he purpose of life is to produce and consume automobiles.'')
Early in the automotive era cars were seen as an ideal solution to many
urban ills (Brown, Morris, and Taylor 2009), which contributes to
explain their enthusiastic reception. In the second part of the 20th
century, motorized mobility rose to become the dominant form of
transportation in cities and regions around the world. The automobile
replaced walking as the key determinant of how far settlements could
grow, and cities grew to accommodate this form of transportation, often
with unfortunate single-mindedness.

The results of this process have been calamitous.

Reliance on motorized mobility contributes to pollution, climate change,
the erosion of social capital and sense of community, road fatalities,
and poor health due to sedentary lifestyles. For years now, work has
aimed to grow a consensus about the importance of communities that
better serve all their residents, and not only their vehicles. The
15-minute neighborhood {[}15MN; Pozoukidou and Chatziyiannaki (2021){]}
is among a handful of ideas that strive to emphasize movement at a human
scale, in environments that accommodate a wide range of capabilities
throughout the lifespan, with the aim of improving livability and health
in ways that automobility can no longer promise, let alone deliver.

A challenge faced by 15MNs is the legacy of decades of auto-centric
planning. Streetscapes are key parts of the hardware of cities, not only
for what is evident at surface level (e.g., sidewalks, pavements), but
also due to other accessory but hidden infrastructure, both physical
(e.g., water, sewage, power) and social (e.g., property rights, right of
ways). Suburban developments in North America are often implicitly or
explicitly designed to discourage through-traffic. This is done by
creating predominantly single use landscapes with meandering, poorly
connected roads. Alas, this form of development cuts both ways, since
the high built-in cost of navigation does not discriminate between
outsiders and residents.

A relevant question concerns the kinds of streetscapes that can support
15MNs. To address this question, we investigate the current
accessibility situation in parts of Canada's major metropolitan region.
The analysis consists of two parts, with positive and normative
characters. 15-minute walking neighborhoods are studied, and their
accessibility levels assessed (positive analysis). Optimal opportunity
landscapes are then used to simulate equivalent opportunity landscapes
throughout the region. Accessibility is then reanalyzed from the
normative perspective of the provision of opportunities. The results of
this analysis are finally correlated to neighborhood network attributes,
including connectivity, centrality, and clustering. The results of this
investigation provide valuable information about neighborhoods, their
morphology and potential to support the aspirational goal of providing
opportunities within 15-minute walks for their residents. This
information can help to identify target neighborhoods for planning
interventions, as well as neighborhoods for whom the 15-minute ideal
could be little more than wishful thinking.
\\
\keywords{
        Walking \and
        Accessibility \and
        15-minute neighborhood \and
    }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-brown2009planning}{}}%
Brown, J. R., E. A. Morris, and B. D. Taylor. 2009. {``Planning for Cars
in Cities: Planners, Engineers, and Freeways in the 20th Century.''}
Journal Article. \emph{Journal of the American Planning Association} 75
(2): 161--77. \url{https://doi.org/10.1080/01944360802640016}.

\leavevmode\vadjust pre{\hypertarget{ref-pozoukidou2021fifteen}{}}%
Pozoukidou, Georgia, and Zoi Chatziyiannaki. 2021. {``15-Minute City:
Decomposing the New Urban Planning Eutopia.''} Journal Article.
\emph{Sustainability} 13 (2): 928.
\url{https://doi.org/10.3390/su13020928}.

\leavevmode\vadjust pre{\hypertarget{ref-roberts1998short}{}}%
Roberts, Ian. 1998. {``A Short History of Walking.''} Journal Article.
\emph{Nature Medicine} 4 (3): 263--64.
\url{https://doi.org/10.1038/nm0398-263}.

\end{CSLReferences}

\bibliographystyle{spphys}
\bibliography{bibliography.bib}

\end{document}

### 2023-06-14 README.Rmd
Initial commit

--- 2023-06-14 Fifteen-minute-neighborhoods/Fifteen-minute-neighborhoods-title-page.Rmd
@@
 ---
-title: 'Reality or wishful thinking: Examining the feasibility of 15-minute neighborhoods from the perspective of network attributes' 
-subtitle: 
-titlerunning: 'Examinining fifteen minute neighborhoods from the perspective of network attributes'
-authorrunning: 
-thanks: | 
-    This project was supported by a McMaster Institute for Research on Aging Catalyst Grant.
+output: github_document
+bibliography: [fifteen-minute-neighborhoods/bibliography.bib]
+---
 
-authors: 
-- name: Antonio Paez
-  address: School of Earth, Environment and Society, McMaster University
-  email: paezha@mcmaster.ca
+<!-- README.md is generated from README.Rmd. Please edit that file -->
 
-- name: James R. Dunn
-  address: Department of Health, Aging and Society, McMaster University
-  email: jim.dunn@mcmaster.ca
+```{r, include = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
+```
 
-- name: Josh Arbess
-  address: Mechanical Engineering and Society, McMaster University
-  email: arbessj@mcmaster.ca
+# Reality or wishful thinking: Examining the feasibility of 15-minute neighborhoods from the perspective of network attributes
 
-keywords:
-- Walking
-- Accessibility
-- 15-minute neighborhood
+<!-- badges: start -->
+<!-- badges: end -->
 
-#PACS: 
-#- PAC1
-#- superPAC
+- Antonio Paez, School of Earth, Environment and Society, McMaster University  
+- Belén Martín Ramos, ETSI de Montes, Forestal y del Medio Natural, Universidad Politécnica de Madrid  
+- Josh Arbess, Mechanical Engineering and Society, McMaster University  
+- James R. Dunn, Department of Health, Aging and Society, McMaster University
 
-#MSC:
-#- MSC code 1
-#- MSC code 2
+## Extended Abstract
 
-abstract: |
 Walking has been the primary form of personal transportation for as long as humans have been humans.
@@
 To address this question, we investigate the current accessibility situation in parts of Canada’s largest metropolitan region.
-The analysis consists of two parts, with positive and normative aspects 15-minute walking neighborhoods are studied, and their accessibility levels assessed (positive analysis).
+The analysis consists of two parts, with positive and normative aspects.
+15-minute walking neighborhoods are studied, and their accessibility levels assessed (positive analysis).
 Maximal opportunity landscapes are then used to simulate equivalent opportunity landscapes throughout the region.
@@
 
-bibliography: bibliography.bib
-biblio-style: spphys
-# bibstyle options spbasic(default), spphys, spmpsci
-output: rticles::springer_article
----
-
-<!--
-@knight2018walkable
-@liu2022toward
-@pozoukidou2021fifteen
-@weng2019fifteen
-
- Walking has naturally, at an evolutionary level, been the primary form of personal transportation for as long as humans have been humans.
-From every prehistoric migration to the dawn of populated settlements, the space-time constraints of walking have fundamentally determined how far interactions with the environment and other humans were feasible.
-It took tens of thousands of years for humans to walk out from their original niche in the African savannah to eventually colonize, and settle in, every nook of the planet.
-    The advent of permanent human settlements at some point in the Neolithic period imposed stronger space-time constraints on people’s movement than had hitherto been the case.
-The need to return to a fixed abode or to go to a designated place to drudge or trade meant that there was only so much space that a person could trade for time, while still being able to toil for life, before even enjoying leisure and recreation.
-Despite the invention of the wheel by the Babylonians, and the use of animal power for transportation purposes, humanity remained for the most part pedestrian for millennia.
-Animal power was a rarity—an expensive and therefore exclusive rarity: by some estimates, horse ownership in the US in the 18th century was dominated by white landowners.
-The physical effort involved in walking necessarily limited the geographical extent of settlements.
-In a very real way, walking determined how far settlements could grow, and in turn settlements evolved that facilitated walking.
-    From a historical perspective, the dominance of movement by foot came to a screeching halt all within the span of a few decades.
-Several revolutions contributed to this change, starting in the 19th century, but accelerating and possibly reaching their peak in the late 20th or early 21st centuries.
-Among these were technological innovations in materials and energy that led to the internal combustion engine, smooth paving surfaces, and systems to transport fuels across vast distances to enable many other forms of transportation.
-At the same time, there were important socio-technological innovations such as Fordism, which created the basis for modern industrial and labor-economic systems to sustain mass production and consumption, initially of automotive vehicles and still, many decades later, still dominated by the auto industry supply chain complex (Jane Jacobs famously quipped that “[t]he purpose of life is to produce and consume automobiles.”) Early in the automotive era cars were seen as an ideal solution to many urban ills (e.g., congestion, manure); in the second part of the 20th century, motorized mobility rose to become in many places the dominant form of transportation.
-The automobile replaced walking as the determinant of how far settlements could grow, and as a matter of policy cities decided to grow to accommodate this form of transportation, often with an unfortunate single-minded focus.
-    The results of this process have been nothing short of disastrous.
-    The impacts of reliance on motorized mobility are anywhere we look, from pollution, the contributions of the transportation sector to climate change, the erosion of social capital and sense of community in the places where we live, road fatalities, and the public health implications of sedentary lifestyles.
-For at least two decades now, planning, transportation, and public health professionals and scholars alike have been working towards a consensus on the importance of creating and fostering communities that better serve all their residents, and not only their vehicles.
-A handful of ideas have been proposed that aim to meet this objective, including transit-oriented development, transit villages, aging-friendly cities (alternatively neighborhoods), and more recently the 15- (or sometimes 5-, 20- or 30-) minute city (or neighborhood).
-These concepts all share some commonalities: movement happens at a more human scale, in a way that accommodates capacities throughout the lifespan, and that contributes to quality of life and health in ways that automobility can no longer promise, let alone deliver.
-     A challenge faced by proposals such as the 15-minute city is the legacy of decades of auto-centric urban planning.
-Streetscapes are a firm element of the hardware of cities, and not only for what is evident at surface level (e.g., sidewalks, pavements), but also due to all the accessory but hidden infrastructure, both physical (e.g., water, sewage, power) and social (e.g., property rights, right of ways).
-In North America, suburban developments (particularly in the post-war period), were implicitly or explicitly designed to discourage through-traffic.
-This was done by creating predominantly single use suburbs with meandering, poorly connected roads.
-Alas, this form of development cuts both ways as the built-in high cost of navigation does not discriminate between outsiders and residents.
-A relevant question therefore concerns the kinds of streetscapes that can in principle support 15-minute neighborhoods.
-    To address this question, we investigate the current accessibility situation in parts of Canada’s major metropolitan region.
-The analysis consists of two parts, with positive and normative characters.
-15-minute walking neighborhoods are studied, and their accessibility levels assessed (positive analysis).
-Optimal opportunity landscapes are then used to simulate equivalent opportunity landscapes throughout the region.
-Accessibility is then reanalyzed from the normative perspective of the provision of opportunities.
-The results of this analysis are finally correlated to neighborhood network attributes, including connectivity, centrality, and clustering.
-    The results of this investigation provide valuable information about neighborhoods, their morphology and potential to support the aspirational goal of providing opportunities within 15-minute walks for their residents.
-This information can help to identify target neighborhoods for planning interventions, as well as neighborhoods for whom the 15-minute ideal could be little more than wishful thinking.
-
--->
-
 ## References

### 2023-06-14 data-raw/01-OSM-Network-and-Walksheds.Rmd
Initial commit

---
title: "Network-Preprocessing"
author: "Antonio Paez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document I collect and preprocess the network data for this research.

# Preliminaries

Load packages:
```{r}
# Allocate memory
options(java.parameters = "-Xmx20G")

library(disk.frame) # Larger-than-RAM Disk-Based Data Manipulation Framework
library(dplyr) # A Grammar of Data Manipulation
library(ggforce) # Accelerating 'ggplot2'
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(h3) # R Bindings for H3
library(purrr) # Functional Programming Tools
library(r5r) # Rapid Realistic Routing with 'R5'
library(sf) # Simple Features for R
library(skimr) # Compact and Flexible Summaries of Data
library(units) # Measurement Units for R Vectors
library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

Load data objects from package {vaccHamilton}. Urban types is the way the City of Hamilton classifies different parts of the city as "Urban", "Suburban", and "Rural". The second object (`data_da_2016`) includes the geometry of Dissemination Areas and some demographic data for 2016:
```{r}
data("urban_types")
data("data_da_2016")
```

Use only urban and suburban Hamilton:
```{r}
urban_hammer <- urban_types |>
  filter(Type != "Rural")
```

Find the DAs that are urban or suburban:
```{r}
urban_hammer_da <- data_da_2016 |>
  st_transform(crs = st_crs(urban_types)) |>
  st_intersection(urban_hammer)
```

Plot urban/suburban DAs:
```{r}
ggplot() + 
  geom_sf(data = urban_hammer_da) + 
  geom_sf(data = urban_hammer,
          aes(color = Type), 
          fill = NA, 
          linewidth = 1)
```

Notice that some DAs appear twice after intersecting with urban fabric types:
```{r}
urban_hammer_da |> 
  left_join(
    urban_hammer_da |> 
      st_drop_geometry() |>
      group_by(GeoUID) |>
      summarize(n = n()),
    by = "GeoUID") |>
  ggplot() +
  geom_sf(aes(fill = factor(n)))

```

This is not a problem. In most cases, some DAs will get two walksheds instead of one.

Coordinates of centroids of the DAs that intersect urban/suburban types (in lat-long). These will be the origins of trips, that is, our proxies for "neighborhoods":
```{r}
origins <- st_centroid(urban_hammer_da) |>
  st_transform(crs = st_crs(4326))
```

# Geohashes

The simplest way to proceed, other than trying to calculate the travel times to every single amenity (and considering that we will at some point simulate the locations of amenities) is to geohash the region of interest. That way, we can count the number of amenities for polygons with precalculated travel times.

Retrieve h3 geohashing at a fairly high resolution to identify Hamilton; the coordinates are for the City of Hamilton:
```{r}
urban_hammer_h3 <- geo_to_h3(c(43.2501, -79.8496),
                             res = 4) |>
  k_ring(radius = 1)
```

Plot the geohash retrieved in this way:
```{r}
ggplot() + 
  geom_sf(data = h3_to_geo_boundary_sf(urban_hammer_h3),
          aes(fill = h3_index)) +
  geom_sf(data = urban_types,
          fill = NA)
```

Keep only the h3 indices that overlap with Hamilton:
```{r}
urban_hammer_h3 = c("842b9b7ffffffff", "842b9b5ffffffff", "842ab4bffffffff")
```

Plot kept indices:
```{r}
ggplot() + 
  geom_sf(data = h3_to_geo_boundary_sf(urban_hammer_h3),
          aes(fill = h3_index)) +
  geom_sf(data = urban_types,
          fill = NA)
```

Obtain children at a higher resolution:
```{r}
urban_hammer_h3 <- urban_hammer_h3 |> 
  purrr::map(\(x) h3_to_children(x, res = 8)) |>
  unlist()
```

Plot current geohashes:
```{r}
ggplot() + 
  geom_sf(data = h3_to_geo_boundary_sf(urban_hammer_h3)) +
  geom_sf(data = urban_types,
          fill = NA)
```

Choose only hexagons that touch urban/suburban Hamilton:
```{r}
urban_hammer_h3 <- h3_to_geo_boundary_sf(urban_hammer_h3) |>
  st_transform(crs = st_crs(urban_types)) |>
  st_intersection(urban_types |> 
                    filter(Type != "Rural"))

urban_hammer_h3 <- urban_hammer_h3 |> 
  pull(h3_index)
```

Plot:
```{r}
ggplot() + 
  geom_sf(data = h3_to_geo_boundary_sf(urban_hammer_h3)) +
  geom_sf(data = urban_types,
          fill = NA)
```

Currently this is resolution 8:
```{r}
h3_get_resolution(urban_hammer_h3)
```

Obtain children at resolution 13 (average area is 43.87 $m^2$):
```{r}
urban_hammer_h3 <- urban_hammer_h3 |> 
  purrr::map(\(x) h3_to_children(x, res = 13)) |>
  unlist()
```

```{r}
destinations_h3 <- h3_to_geo(urban_hammer_h3)
```

## Network routing

I used [BBBike](https://download.bbbike.org/osm/bbbike/) to extract OSM data for Hamilton on July 7, 2023. The name of the file is `planet_-80.279,43.048_-79.318,43.459.osm.pbf`. I copied this file to folder `data-raw/r5_graph`.

Set Up R5 Routing. First define the path to where the graph is located:
```{r set up r5 path, include=FALSE}
r5_path <- file.path("./r5_graph")
```

Download and import GTFS (Hamilton transit data)
```{r eval=FALSE}
download.file(url = "https://transitfeeds.com/p/hamilton-street-railway/31/latest/download", 
              destfile = file.path(r5_path, 
                                   "HSR_transit.zip"), 
              mode = "wb")
```

Download and import GTFS (Hamilton transit data)
```{r eval=FALSE}
download.file(url = "https://transitfeeds.com/p/burlington-transit/294/latest/download", 
              destfile = file.path(r5_path, 
                                   "Burlington_transit.zip"), 
              mode = "wb")
```

Build the graph:
```{r build-graph, include = FALSE}
r5_hamilton_cma <- setup_r5(data_path = r5_path, 
                            verbose = FALSE)
```

Prepare Input Data for `r5r`. The origins are the coordinates of the DAs and the destinations the coordinates of the h3 geohashes:
```{r prepare-inputs}
# save origins in format expected by R5R (id, lon, lat)
origins_i <- data.frame(ID = origins$GeoUID, 
                        origins |>
                          st_transform(crs = 4326) |>
                          st_coordinates()) |>
  rename(lon = X, lat = Y, id = ID) |>
  dplyr::select(id, lon, lat)

# now vaccination sites
destinations_j <- data.frame(id = urban_hammer_h3, 
                             lat = destinations_h3[,1],
                             lon = destinations_h3[,2])
```

Calculate OD Matrix for walking:
```{r calculate walk od matrix}

# set up batching according to how many origin rows to process at one time
chunksize = 4000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
num_chunks = ceiling(nrow(origins_i)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(origins_i,
                                outdir = "./df/origins_i",
                                nchunks = num_chunks,
                                overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){ 
  origins_i_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5_hamilton_cma,
                                  origins = origins_i_chunk,
                                  destinations = destinations_j,
                                  mode = c("WALK"),
                                  departure_datetime = as.POSIXct(strptime("2021-04-05 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                                  max_walk_time = 15, # minutes
                                  max_trip_duration = 15)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "./df/output_ttm_walk",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
ttm_walk_time <- end.time - start.time
ttm_walk_time
```

Retrieve network as a simple features object:
```{r}
hamilton_net <- street_network_to_sf(r5_hamilton_cma)
```

Save network as sf:
```{r}
save(hamilton_net, file = "hamilton_net.rda", compress = "xz")
```

## Extract travel time matrix

```{r load od matrix for walk, include=FALSE}
# connect to the walking travel time matrix disk frame
ttm_walk.disk.frame <- disk.frame("./df/output_ttm_walk")
#opportunities_j <- as.data.frame(nyc_cb_point) |> select(GEOID10, total_emp) |> rename(toId = GEOID10, o_j = total_emp)
```

Convert disk.frame to data frame:
```{r}
ttm_walk <- as.data.frame(ttm_walk.disk.frame) |>
  transmute(GeoUID = from_id, h3_index = to_id, travel_time = travel_time_p50)
```

## Retrieve h3 geometries of walksheds by origin

Retrieve the h3 areas as multipolygons:
```{r}
walksheds <- ttm_walk |> 
  group_by(GeoUID) |>
  group_map(~ h3_set_to_multi_polygon(.x$h3_index),
            .keep = TRUE)
```

Obtain the GeoUIDs of the walksheds:
```{r}
walkshed_GeoUIDs <- ttm_walk |>
  group_by(GeoUID) |>
  summarize(GeoUID = first(GeoUID),
            .groups = "drop")
```

Convert walksheds with GeoUIDs of DAs to a data frame:
```{r}
walksheds <- data.frame(GeoUID = walkshed_GeoUIDs,
                         geometry = purrr::map_df(walksheds,
                            tibble::as_tibble)) |>
  st_as_sf(crs = 4326)

```

Verify the catchments:
```{r}
test_unit = sample.int(nrow(walksheds), 1)

ggplot() +
  geom_sf(data = data_da_2016 |>
            filter(GeoUID == walksheds$GeoUID[test_unit]),
          color = "black",
          fill= NA) + 
  geom_sf(data = walksheds[test_unit,],
          color = "blue",
          fill = NA,
          linewidth = 1)
```

Join the urban type to DAs in the walksheds object:
```{r}
walksheds <- walksheds |>
  left_join(origins |>
              st_drop_geometry() |>
              transmute(GeoUID, Type = Type.1),
            by = "GeoUID")
```

Convert to factor:
```{r}
walksheds <- walksheds |>
  mutate(Type = factor(Type))
```

Summarize the walksheds:
```{r}
summary(walksheds)
```

The DAs and the catchments match. Now save the data objects:
```{r}
save(urban_hammer_da, file = "urban_hammer_da.rda", compress = "xz")
save(ttm_walk, file = "ttm_walk.rda", compress = "xz")
save(walksheds, file = "walksheds.rda", compress = "xz")
```

Calculate area of walksheds:
```{r}
walksheds <- walksheds |>
  mutate(area = st_area(geometry))
```

Summarize by urban type:
```{r}
walksheds |>
  st_drop_geometry() |>
  mutate(area = units::drop_units(area)) |>
  group_by(Type) |>
  skim()
```

```{r}
walksheds |>
  st_drop_geometry() |>
  mutate(area = log(units::drop_units(set_units(area, km2)))) |>
  lm(area ~ Type, data = _) |>
  summary()
```

```{r}
ggplot(data = walksheds |>
  st_drop_geometry() |>
  mutate(area = units::drop_units(set_units(area, km2))),
  aes(x = Type, y = area)) +
  geom_violin()
```

Urban walksheds close to the mean:
```{r}
walksheds |>
  st_drop_geometry() |>
  mutate(area = units::drop_units(area)) |>
  filter(Type == "Urban", area >= 1372092 * 0.999 & area <= 1372092 * 1.001)
```

Suburban walksheds close to the mean area of suburban walksheds:
```{r}
walksheds |>
  st_drop_geometry() |>
  mutate(area = units::drop_units(area)) |>
  filter(Type == "Suburban", area >= 1100484 * 0.99 & area <= 1100484 * 1.01)
```

Compare the areas of walksheds using comparable areas:
```{r}
mock_areas <- data.frame(Type = c("Suburban",
                          "Urban"), 
                 area = c(1100484,
                          1372092),
                 x0 = c(0,0),
                 y0 = c(0,0)) |>
  mutate(r = sqrt(area/pi))
```

```{r}
ggplot() +
  geom_circle(data = mock_areas,
              aes(x0 = x0,
                  y0 = y0,
                  r = r,
                  color = Type))+
  coord_equal()
```

### 2023-06-14 data-raw/02-OSM-Amenities.Rmd
Initial commit

---
title: "Network-Preprocessing"
author: "Antonio Paez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document I collect and preprocess the OSM data with amenities.

# Preliminaries

Load packages:
```{r}
library(dplyr) # A Grammar of Data Manipulation
library(forcats) # Tools for Working with Categorical Variables (Factors)
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(h3) # R Bindings for H3
library(osmdata) # Import 'OpenStreetMap' Data as Simple Features or Spatial Objects
library(purrr) # Functional Programming Tools
library(sf) # Simple Features for R
library(tidyr) # Tidy Messy Data
library(units) # Measurement Units for R Vectors
library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

Load data objects from package {vaccHamilton}. Urban types is the way the City of Hamilton classifies different parts of the city as "Urban", "Suburban", and "Rural":
```{r}
data("urban_types")
#data("data_da_2016")
```

Use only urban and suburban Hamilton:
```{r}
urban_hammer <- urban_types |>
  filter(Type != "Rural")
```

# Obtain amenities

Set the bounding box to the extent of urban/suburban Hamilton:
```{r}
b <- st_bbox(urban_hammer |>
               st_transform(crs = 4326))
```

Initialize the query:
```{r}
amenities <- opq(bbox = b) |>
  add_osm_feature(key = "amenity") |>
  osmdata_sf()
```

The number of amenities retrieved is:
```{r}
nrow(amenities$osm_points)
```

See OSM documentation for relevant amenities: https://wiki.openstreetmap.org/wiki/Key:amenity

```{r}
summary(factor(amenities$osm_points$amenity))
```
Notice the large number of NAs.

Refine search by segmenting the categories of amenities.

Sustenance:
```{r}
sustenance <- opq(bbox = b) |>
  add_osm_features(c("\"amenity\"=\"bar\"",
                     "\"amenity\"=\"biergarten\"",
                     "\"amenity\"=\"cafe\"",
                     "\"amenity\"=\"fast_food\"",
                     "\"amenity\"=\"food_court\"",
                     "\"amenity\"=\"ice_cream\"",
                     "\"amenity\"=\"pub\"",
                     "\"amenity\"=\"restaurant\"")) |>
  osmdata_sf()
```

Check the output:
```{r}
summary(factor(sustenance$osm_points$amenity))
```

Extract the points and add a column with the category "Sustenance":
```{r}
sustenance_points <- sustenance$osm_points |>
  mutate(Category = "Sustenance")
```

Education:
```{r}
education <- opq(bbox = b) |>
  add_osm_features(c("\"amenity\"=\"library\"",
                     "\"amenity\"=\"school\"")) |>
  osmdata_sf()
```

Check the output:
```{r}
summary(factor(education$osm_points$amenity))
```

We don't like this...what is 1,938 NAs in schools and libraries? Libraries is a reasonable number (there are about 22 public libraries, and some are rural). For this reason extract only libraries.

Extract the points, remove the schools and NAs and add a column with the category "Library":
```{r}
library_points <- education$osm_points |>
  filter(amenity == "library") |>
  mutate(Category = "Library")
```

Financial:
```{r}
financial <- opq(bbox = b) |>
  add_osm_features(c("\"amenity\"=\"atm\"",
                     "\"amenity\"=\"bank\"",
                     "\"amenity\"=\"bureau_de_change\"")) |>
  osmdata_sf()
```

Check the output:
```{r}
summary(factor(financial$osm_points$amenity))
```

Possibly 200+ banks and atms not labeled?

Extract the points and add a column with the category "Library":
```{r}
financial_points <- education$osm_points |>
  mutate(Category = "Financial")
```

Health services:
```{r}
healthcare <- opq(bbox = b) |>
  add_osm_features(c("\"amenity\"=\"clinic\"",
                     "\"amenity\"=\"dentist\"",
                     "\"amenity\"=\"doctors\"",
                     "\"amenity\"=\"pharmacy\"")) |>
  osmdata_sf()
```

Check the output:
```{r}
summary(factor(healthcare$osm_points$amenity))
```

Extract the points and add a column with the category "Library":
```{r}
healthcare_points <- healthcare$osm_points |>
  mutate(Category = "Healthcare")
```

Bind all amenity categories as points:
```{r}
amenities_points <- rbind(sustenance_points |>
                            select(osm_id, name, amenity, Category),
                          library_points |>
                            select(osm_id, name, amenity, Category),
                          financial_points |>
                            select(osm_id, name, amenity, Category),
                          healthcare_points |>
                            select(osm_id, name, amenity, Category))
```

Plot:
```{r}
ggplot() +
  geom_sf(data = urban_types,
          aes(fill = Type)) +
  geom_sf(data = amenities_points,
          aes(color = Category))
```

Obtain the coordinates of the points:
```{r}
amenities_coords <- amenities_points |>
  st_coordinates()

amenities_points <- amenities_points |>
  mutate(lon = amenities_coords[,1], lat = amenities_coords[,2])
```

Use the coordinates to obtain the h3 indices at level 13:
```{r}
amenities_points <- amenities_points |>
  mutate(h3_index = geo_to_h3(st_coordinates(geometry),
                              13))
```

Save the data objects:
```{r}
save(sustenance, file = "sustenance.rda", compress = "xz")
save(education, file = "education.rda", compress = "xz")
save(financial, file = "financial.rda", compress = "xz")
save(healthcare, file = "healthcare.rda", compress = "xz")
save(amenities_points, file = "amenities_points.rda", compress = "xz")
```

Append type of urban fabric to amenities:
```{r}
amenities_by_type <- amenities_points |>
  select(Category) |>
  st_transform(crs = st_crs(urban_types)) |>
  st_intersection(urban_types)
```

Summarize number of amenities by category by urban type:
```{r}
amenities_by_type <- amenities_by_type |>
  st_drop_geometry() |>
  group_by(Type) |>
  group_map(~table(.x$Category))

amenities_by_type
```

Notice how the number of amenities is lowest for "Rural" but this is because our search did not include all of rural Hamilton. Still, there is a difference between suburban and urban Hamilton, with urban Hamilton counting more amenties of almost every category. This is despite the difference in areas between the two regions: suburban Hamilton covers considerably more area than urban Hamilton.
```{r}
area_by_type <- urban_hammer |>
  mutate(area = st_area(geometry) |>
           units::set_units(km2))

area_by_type
```

What is the density of amenities in each of these two urban fabrics?
```{r}
suburban_density <- amenities_by_type[[2]]/drop_units(area_by_type$area[1])
urban_density <- amenities_by_type[[3]]/drop_units(area_by_type$area[2])

data.frame(suburban = suburban_density,
           urban = urban_density) |>
  mutate(ratio = urban.Freq/suburban.Freq) |>
  transmute(Category = suburban.Var1, 
            suburban_density = set_units(suburban.Freq, 1/km2), 
            urban_density = set_units(urban.Freq, 1/km2), 
            ratio)
```

### 2023-06-14 data-raw/03-Network-Attributes-by-Walkshed.Rmd
Initial commit

---
title: "Network-Preprocessing"
author: "Antonio Paez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document I aim to partition the network based on walksheds.

# Preliminaries

Load packages:
```{r}
#library(disk.frame) # Larger-than-RAM Disk-Based Data Manipulation Framework
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(igraph)
#library(h3) # R Bindings for H3
#ibrary(osmdata)
#library(purrr) # Functional Programming Tools
#library(r5r) # Rapid Realistic Routing with 'R5'
library(sf) # Simple Features for R
library(sfnetworks)
library(tidyr)
#library(skimr) # Compact and Flexible Summaries of Data
#library(units)
#library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

```{r}
load(file = "hamilton_net.rda")
load(file = "urban_hammer_da.rda")
load(file = "walksheds.rda")
```

Extract the edges of the network:
```{r}
hamilton_edges <- hamilton_net$edges
```

Calculate network attributes by walkshed:
```{r}
# Time algorithm
start.time <- Sys.time()

# Initialize data frame
walksheds_net_attributes <- data.frame(GeoUID = walksheds$GeoUID,
                                       n_nodes = numeric(889), 
                                       transitivity = numeric(889),
                                       motifs = numeric(889),
                                       n_edges = numeric(889),
                                       diameter = numeric(889), 
                                       radius = numeric(889),
                                       girth = numeric(889), 
                                       global_efficiency = numeric(889),
                                       edge_density = numeric(889),
                                       edge_connectivity = numeric(889),
                                       mean_distance = numeric(889),
                                       min_cut = numeric(889))

for(i in 1:nrow(walksheds)){
  wsnet <- st_intersection(hamilton_edges, 
                           walksheds[i,]) |> 
    st_cast(to = "LINESTRING")
  wsnet <- as_sfnetwork(wsnet)
  E(wsnet)$weight <- E(wsnet)$length
  walksheds_net_attributes$n_nodes[i] <- gorder(wsnet)
  walksheds_net_attributes$transitivity[i] <- transitivity(wsnet)
  walksheds_net_attributes$motifs[i] <- count_motifs(wsnet)
  walksheds_net_attributes$n_edges[i] <- gsize(wsnet)
  walksheds_net_attributes$diameter[i] <- diameter(wsnet)
  walksheds_net_attributes$radius[i] <- radius(wsnet)
  walksheds_net_attributes$girth[i] <- girth(wsnet)$girth
  walksheds_net_attributes$global_efficiency[i] <- global_efficiency(wsnet)
  walksheds_net_attributes$edge_density[i] <- edge_density(wsnet)
  walksheds_net_attributes$edge_connectivity[i] <- edge_connectivity(wsnet)
  walksheds_net_attributes$mean_distance[i] <- mean_distance(wsnet)
  walksheds_net_attributes$min_cut[i] <- min_cut(wsnet)
  print(i)
}

end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

Join the type of urban fabric:
```{r}
walksheds_net_attributes <- walksheds_net_attributes |>
  mutate(Type = walksheds$Type) |>
  select(GeoUID, Type, n_nodes:min_cut)
```

Save the data object:
```{r}
save(walksheds_net_attributes, file = "walksheds_net_attributes.rda", compress = "xz")
```

```{r}
walksheds_net_attributes |>
  group_by(Type) |>
  skim()
```

Edge connectivity is not super useful: there is little variation - it is a measure of network vulnerability, minimum number of edges to remove to disconnect two nodes (there is no longer a path between them.)

Transitivity: it is the probability that the neighbors of a node are neighbors between them. In a Manhattan-style grid this number will be low because there will be lots and lots of squares.

Motifs (calculated for size 3): It indicates the presence of particular subgraphs that repeat themselves. In other words, a graph with a large number of motifs will tend to have few unique elements (some elements happen again, again, and again). A graph with many unique elements will have few motifs (repetitive patterns).

Motfis (from igraph help pages)
```{r}
g <- sample_pa(100)
motifs(g, 3)
count_motifs(g, 3)
sample_motifs(g, 3)

```

```{r}
pattern <- graph.full(3)
my.graph <- grg.game(100, 0.2) 
```

### 2023-06-14 data-raw/04-Accessibility-Calculations.Rmd
Initial commit

---
title: "Network-Preprocessing"
author: "Antonio Paez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document I aim to partition the network based on walksheds.

# Preliminaries

Load packages:
```{r}
#library(disk.frame) # Larger-than-RAM Disk-Based Data Manipulation Framework
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
#library(igraph)
#library(h3) # R Bindings for H3
#ibrary(osmdata)
#library(purrr) # Functional Programming Tools
#library(r5r) # Rapid Realistic Routing with 'R5'
library(sf) # Simple Features for R
#library(sfnetworks)
library(tidyr)
library(skimr) # Compact and Flexible Summaries of Data
#library(units)
library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

```{r}
#load(file = "hamilton_net.rda")
load(file = "amenities_points.rda")
load(file = "urban_hammer_da.rda")
data(urban_types)
load(file = "walksheds.rda")
```

Intersect amenities with walksheds to obtain the cumulative opportunities:
```{r}
amenities_walksheds <- amenities_points |>
  select(amenity, Category) |>
  st_intersection(walksheds)
```

Save amenities by walkshed:
```{r}
save(amenities_walksheds, file = "amenities_walkshed.rda", compress = "xz")
```

Summary of amenity categories by DA and Type:
```{r}
amenities_walksheds_summary <- amenities_walksheds |>
  drop_na() |>
  st_drop_geometry() |>
  group_by(GeoUID,
           Type,
           Category) |>
  summarize(n = n(),
            .groups = "drop") |>
  pivot_wider(id_cols = c(GeoUID, Type),
              names_from = Category,
              values_from = n,
              values_fill = 0)
```

Map accessibility by category of amenity:
```{r}
urban_hammer_da |>
  select(GeoUID, Type.1) |>
  rename(Type = Type.1) |>
  left_join(amenities_walksheds_summary,
            by = c("GeoUID", "Type")) |>
  mutate(across(Sustenance:Library, ~ replace_na(.x, 0))) |>
  pivot_longer(cols = -c(GeoUID, Type, geometry),
               names_to = "Amenity_Category",
               values_to = "Number_of_Amenities") |>
  #mutate(Number_of_Amenities = ifelse(Number_of_Amenities == 0, NA, Number_of_Amenities)) |>
  mutate(Number_of_Amenities = case_when(Number_of_Amenities == 0 ~ "0",
                                         Number_of_Amenities > 0 & Number_of_Amenities <= 3 ~ "1-3",
                                         Number_of_Amenities > 3 & Number_of_Amenities <= 6 ~ "4-6",
                                         Number_of_Amenities > 6 & Number_of_Amenities <= 9 ~ "7-9",
                                         Number_of_Amenities > 9 ~ "10 or more"),
         Number_of_Amenities = factor(Number_of_Amenities,
                                      levels = c("0", "1-3", "4-6", "7-9", "10 or more"),
                                      ordered = TRUE)) |>
  ggplot() +
  geom_sf(aes(fill = Number_of_Amenities)) +
  geom_sf(data = urban_types |>
            filter(Type != "Rural"),
          aes(color = Type),
          fill = NA) +
  #scale_fill_fermenter(direction = 1, palette = "Greens") +
  facet_wrap(~ Amenity_Category) + 
  theme_void()
```

Find the most 15-minutes DA:
```{r}
amenities_walksheds_summary |>
  filter(Financial == max(Financial) |
         Sustenance == max(Sustenance) |
         Healthcare == max(Healthcare) |
         Library == max(Library))
```

Looks like GeoUID == "35250420" has a really nice combination: max in banking, max in sustenance, pretty high in health, and one library.

How many DAs have zero amenities?
```{r}
zero_amenity_walksheds <- urban_hammer_da |>
  select(GeoUID, Type.1) |>
  rename(Type = Type.1) |>
  left_join(amenities_walksheds_summary,
            by = c("GeoUID", "Type")) |>
  mutate(across(Sustenance:Library, ~ replace_na(.x, 0))) |>
  pivot_longer(cols = -c(GeoUID, Type, geometry),
               names_to = "Amenity_Category",
               values_to = "Number_of_Amenities") |>
  group_by(GeoUID, Type) |>
  summarize(Number_of_Amenities = sum(Number_of_Amenities),
            .groups = "drop") |>
  filter(Number_of_Amenities == 0)
```

```{r}
zero_amenity_walksheds |>
  pull(Type) |>
  table()
```

Check to see if there are any DAs in this list that have both an Urban and a Suburban part:
```{r}
zero_amenity_walksheds |>
  group_by(GeoUID, Type) |>
  summarize(n = n()) |>
  filter(n > 1)
```

Nope, there are none.

What is the area of DAs with zero amenities? What is their population?
```{r}
zero_amenity_walksheds_summary <- zero_amenity_walksheds |>
              st_drop_geometry() |>
  left_join(data_da_2016,
            by = c("GeoUID")) |>
  rename(Type.1 = Type.x,
         Type = Type.y) |>
  group_by(Type.1) |>
  summarize(Population = sum(Population),
            area = sum(`Shape Area`),
            .groups = "drop") |>
  mutate(Prop_Population = Population/sum(data_da_2016$Population), 
         Prop_Tot_Area = area/sum(data_da_2016$`Shape Area`),
         walksheds = "Zero Amenities")

zero_amenity_walksheds_summary
```

Find all minimal 15-minutes DAs, meaning those that have at least one of each amenity:
```{r}
minimal_amenities_walksheds <- amenities_walksheds_summary |>
  filter(Financial >= 1,
         Sustenance >= 1,
         Healthcare >= 1,
         Library >= 1)
```

How many minimal 15 min DAs are urban/suburban:
```{r}
minimal_amenities_walksheds |>
  pull(Type) |>
  table()
```

What is the area of DAs with minimal 15-min walk sheds? What is their population?
```{r}
minimal_amenities_walksheds_summary <- minimal_amenities_walksheds |>
              st_drop_geometry() |>
  left_join(data_da_2016,
            by = c("GeoUID")) |>
  rename(Type.1 = Type.x,
         Type = Type.y) |>
  group_by(Type.1) |>
  summarize(Population = sum(Population),
            area = sum(`Shape Area`),
            .groups = "drop") |>
  mutate(Prop_Population = Population/sum(data_da_2016$Population), 
         Prop_Area = area/sum(data_da_2016$`Shape Area`),
         walksheds = "Minimal Amenities")

minimal_amenities_walksheds_summary
```

Populations by urban fabric type:
```{r}
city_wide_summary <- urban_hammer_da |>
  st_drop_geometry() |>
  group_by(Type.1) |>
  summarize(Population = sum(Population),
            area = sum(Shape.Area),
            .groups = "drop") |>
  mutate(Prop_Population = Population/sum(data_da_2016$Population),
         Prop_Area = area/sum(data_da_2016$`Shape Area`),
         walksheds = "City Wide")

city_wide_summary
```

Proportion of suburban and urban population in zero amenity sheds:
```{r}
zero_amenity_walksheds_summary$Population/city_wide_summary$Population
```

Almost 50% of suburban population (which represent about 23% of the total population in Hamilton) live in DAs with zero amenities within the 15-min walk shed.

About 18% of urban population (representing about 5% of the total population of the city) live in DAs with zero amenities within the 15-min walk shed.

Proportion of suburban and urban population in minimal 15-min sheds:
```{r}
minimal_amenities_walksheds_summary$Population/city_wide_summary$Population
```

Barely 2.5% of suburban residents live in DAs with minimal 15-min walk sheds. They represent approximately half a percentage point of the total population of the city.

More than 15% of urban population live in DAs with minimal 15-min walk sheds, and account for approximately 2.6% of the total population of the city.

Compare amenities by category and urban fabric type:
```{r}
amenities_walksheds_summary |>
  mutate(across(Sustenance:Library, ~ replace_na(.x, 0))) |>
  pivot_longer(cols = -c(GeoUID, Type),
               names_to = "Amenity_Category",
               values_to = "Number_of_Amenities") |>
  #mutate(Number_of_Amenities = ifelse(Number_of_Amenities == 0, NA, Number_of_Amenities)) |>
  mutate(Number_of_Amenities = case_when(Number_of_Amenities == 0 ~ "0",
                                         Number_of_Amenities > 0 & Number_of_Amenities <= 3 ~ "1-3",
                                         Number_of_Amenities > 3 & Number_of_Amenities <= 6 ~ "4-6",
                                         Number_of_Amenities > 6 & Number_of_Amenities <= 9 ~ "7-9",
                                         Number_of_Amenities > 9 ~ "10 or more"),
         Number_of_Amenities = factor(Number_of_Amenities,
                                      levels = c("0", "1-3", "4-6", "7-9", "10 or more"),
                                      ordered = TRUE)) |>
  group_by(Amenity_Category,
           Type) |>
  skim()
```

Sample two walksheds with at least one amenity, one from urban fabric type and one suburban, and plot:
```{r}
seed <- sample.int(1000000, 1)
set.seed(seed)
seed

# 409029, 534918

sample_walksheds <- amenities_walksheds |>
  group_by(Type) |>
  slice_sample(n = 1)

ggplot()  + 
  # geom_sf(data = urban_types |>
  #               filter(Type != "Rural"),
  #         aes(fill = Type)) +
  geom_sf(data = walksheds |>
            filter((GeoUID == sample_walksheds$GeoUID[1] & Type == sample_walksheds$Type[1]) | 
                     (GeoUID == sample_walksheds$GeoUID[2] & Type == sample_walksheds$Type[2])),
          aes(fill = Type),
          color = "black")+
  geom_sf(data = amenities_walksheds |>
            filter((GeoUID == sample_walksheds$GeoUID[1] & Type == sample_walksheds$Type[1]) | 
                     (GeoUID == sample_walksheds$GeoUID[2] & Type == sample_walksheds$Type[2])),
          #aes(color = Category)
          )
```

### 2023-06-14 data-raw/05-Walkshed-Profiles.Rmd
Initial commit

---
title: "Profile Walksheds"
author: "Antonio Paez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we profile the ped sheds based on their network attributes.

# Preliminaries

Load packages:
```{r}
library(dplyr) # A Grammar of Data Manipulation
library(evtree) # Evolutionary Learning of Globally Optimal Trees
library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(glue) # Interpreted String Literals
library(patchwork) # The Composer of Plots
library(tidyr) # Tidy Messy Data
library(tree) # Classification and Regression Trees
library(sf) # Simple Features for R
library(skimr) # Compact and Flexible Summaries of Data
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(units) # Measurement Units for R Vectors
library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

Load data:
```{r}
load(file = "amenities_walkshed.rda")
load(file = "hamilton_net.rda")
load(file = "urban_hammer_da.rda")
data("urban_types")
load(file = "walksheds.rda")
load(file = "walksheds_net_attributes.rda")
```

## Classification tree

The aim is to train a classification tree to profile urban and suburban walksheds based on the attributes of the network. We need to drop 1 observation that has an NA in `transitivity`:
```{r}
d <- walksheds_net_attributes |>
                  drop_na()
```

We use a technique with evolutionary trees - an alternative to classification trees that is less affecte by the greediness of the algorithm. Since evolutionary trees have some randomness (for the initial values), we can try several to find whether there are patterns. In this case we try four different random seed.

Tree 1 with seed 1 (912571):
```{r}
#seed1 <- sample.int(1000000, 1)
seed1 <- 912571
set.seed(seed1)
tree1 <- evtree(Type ~ ., 
                data = d |> select(-GeoUID))
summary(tree1)
```

Tree 2 with seed 2 (94766):
```{r}
#seed2 <- sample.int(1000000, 1)
seed2 <- 94766
set.seed(seed2)
tree2 <- evtree(Type ~ ., 
                data = d |> select(-GeoUID))
summary(tree2)
```

Tree 3 with seed 3 (46125):
```{r}
#seed3 <- sample.int(1000000, 1)
seed3 <- 46125
set.seed(seed3)
tree3 <- evtree(Type ~ ., 
                data = d |> select(-GeoUID))
summary(tree3)
```

Tree 4 with seed 4 (556090)
```{r}
#seed4 <- sample.int(1000000, 1)
seed4 <- 556090
set.seed(seed4)
tree4 <- evtree(Type ~ ., 
                data = d |> select(-GeoUID))
summary(tree4)
```

```{r}
#seed4 <- sample.int(1000000, 1)
seed4 <- 556090
set.seed(seed4)
tree5 <- evtree(Type ~ ., 
                data = d |> select(-GeoUID))
summary(tree5)
```

Create a function to estimate the missclassification rate and the evaluation function:
```{r}
mc <- function(obj) 1 - mean(predict(obj) == d$Type)

evalfun <- function(obj) 2 * nrow(d) * mc(obj) + width(obj) * log(nrow(d))

#trees <- list("evtree" = tree1, "rpart" = rp, "ctree" = ct, "rpart2" = rp2, "ctree2" = ct2)
trees <- list("evtree 1" = tree1, "evtree 2" = tree2, "evtree 3" = tree3, "evtree 4" = tree4)

round(sapply(trees, function(obj) c("misclassification" = mc(obj), "evaluation function" = evalfun(obj))), digits = 7)

```

All trees have similar fit.

Plot the trees:
```{r}
tree1_plot <- ggparty(tree1) +
  geom_edge() +
  geom_edge_label(aes(label = breaks_label)) +
  geom_node_label(aes(label = splitvar), ids = "inner") +
  # identical to  geom_node_splitvar() +
  geom_node_label(aes(label = info), ids = "terminal")

tree2_plot <- ggparty(tree2) +
  geom_edge() +
  geom_edge_label(aes(label = breaks_label)) +
  geom_node_label(aes(label = splitvar), ids = "inner") +
  # identical to  geom_node_splitvar() +
  geom_node_label(aes(label = info), ids = "terminal")

tree3_plot <- ggparty(tree3) +
  geom_edge() +
  geom_edge_label(aes(label = breaks_label)) +
  geom_node_label(aes(label = splitvar), ids = "inner") +
  # identical to  geom_node_splitvar() +
  geom_node_label(aes(label = info), ids = "terminal")

tree4_plot <- ggparty(tree4) +
  geom_edge() +
  geom_edge_label(aes(label = breaks_label)) +
  geom_node_label(aes(label = splitvar), ids = "inner") +
  # identical to  geom_node_splitvar() +
  geom_node_label(aes(label = info), ids = "terminal")

(tree1_plot + tree2_plot) / (tree3_plot + tree4_plot)
```

Notice that transitivity and motifs are the dominant attributes in the classification trees. Trees 1, 3, and 4 have the same general structure, and in fact trees 3 and 4 are identical. Based on this we choose to use `tree3`.

Check the break labels of `tree3` to round:
```{r}
ggparty(tree3)$data$breaks_label
```

Plot the tree with the rounded break labels:
```{r}
ggparty(tree3) +
  geom_edge() +
  geom_edge_label(aes(label = str_replace(breaks_label, 
                                          "0.0207311636201516", 
                                          "0.021"))) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.0,
                     ids = c(1, 2, 5)) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.02,
                     ids = c(3, 4, 6, 7)) +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", 
                                            fill = Type),
                                        position = position_fill()),
                               xlab("Type"),
                               theme_minimal()),
                 # draw only one label for each axis
                 shared_axis_labels = TRUE,
                 # draw line between tree and legend
                 legend_separator = FALSE
                 )
```

Find examples of ped sheds in each of the leafs and plot the network for illustration.

Leaf 1:
```{r}

seed <- sample.int(100000, 1)
set.seed(seed)
seed
#69717

st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity < 0.0207311636201516,
                             motifs < 378, 
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf()
```

Leaf 2 (urban):
```{r}
seed <- sample.int(100000, 1)
set.seed(seed)
seed
#84451, 44785

st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(motifs >= 378, 
                             transitivity < 0.0207311636201516,
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf()
```

Leaf 3 (suburban):
```{r}
seed <- sample.int(100000, 1)
set.seed(seed)
seed
#81461, 84451

st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(motifs < 1024, 
                             transitivity >= 0.0207311636201516,
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf()
```

Leaf 4 (urban):
```{r}
seed <- sample.int(100000, 1)
set.seed(seed)
seed
#52534

st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(motifs >= 1024, 
                             transitivity >= 0.0207311636201516,
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf()
```

Add leaf labels to data:
```{r}
d <- d |>
  mutate(leaf = case_when(transitivity < 0.0207311636201516 & motifs < 378 ~ "Leaf 1",
                          transitivity < 0.0207311636201516 & motifs >= 378 ~ "Leaf 2",
                          transitivity >= 0.0207311636201516 & motifs < 1024 ~ "Leaf 3",
                          transitivity >= 0.0207311636201516 & motifs >= 1024 ~ "Leaf 4"),
         leaf = factor(leaf))
```

```{r}
skim(d)
```

```{r}
d |>
  left_join(walksheds |>
              select(-Type) |>
              mutate(area = st_area(geometry),
                     area = units::set_units(area, km^2) |> 
                       drop_units(),
                     area = log(area)) |>
              st_drop_geometry(),
            by = "GeoUID") |>
  lm(area ~ leaf, data = _) |>
  summary()
```

But what about accessibility??

Summary of amenity categories by DA and Type:
```{r}
amenities_walksheds_summary <- amenities_walksheds |>
  drop_na() |>
  st_drop_geometry() |>
  group_by(GeoUID,
           Type,
           Category) |>
  summarize(n = n(),
            .groups = "drop") |>
  pivot_wider(id_cols = c(GeoUID, Type),
              names_from = Category,
              values_from = n,
              values_fill = 0)
```

To map accessibility by category of amenity we first categorize the number of amenities:
```{r}
amentities_da <-  urban_hammer_da |>
  select(GeoUID, Type.1) |>
  rename(Type = Type.1) |>
  left_join(amenities_walksheds_summary,
            by = c("GeoUID", "Type")) |>
  mutate(across(Sustenance:Library, ~ replace_na(.x, 0))) |>
  pivot_longer(cols = -c(GeoUID, Type, geometry),
               names_to = "Amenity_Category",
               values_to = "Number_of_Amenities") |>
  #mutate(Number_of_Amenities = ifelse(Number_of_Amenities == 0, NA, Number_of_Amenities)) |>
  mutate(Number_of_Amenities = case_when(Number_of_Amenities == 0 ~ "0",
                                         Number_of_Amenities > 0 & Number_of_Amenities <= 3 ~ "1-3",
                                         Number_of_Amenities > 3 & Number_of_Amenities <= 6 ~ "4-6",
                                         Number_of_Amenities > 6 & Number_of_Amenities <= 9 ~ "7-9",
                                         Number_of_Amenities > 9 ~ "10 or more"),
         Number_of_Amenities = factor(Number_of_Amenities,
                                      levels = c("0", "1-3", "4-6", "7-9", "10 or more"),
                                      ordered = TRUE))
```

Plot:
```{r}
amentities_da |>
  ggplot() +
  geom_sf(aes(fill = Number_of_Amenities)) +
  geom_sf(data = urban_types |>
            filter(Type != "Rural"),
          aes(color = Type),
          fill = NA) +
  #scale_fill_fermenter(direction = 1, palette = "Greens") +
  facet_wrap(~ Amenity_Category) + 
  theme_void()
```

Pivot wider:
```{r}
junk <- amentities_da |>
  pivot_wider(names_from = "Amenity_Category",
              values_from = "Number_of_Amenities")
```

```{r}
junk2 <- d |>
  left_join(junk |>
              #select(-Type) |>
              st_drop_geometry(),
            by = c("GeoUID", "Type")) |>
  mutate(across(Sustenance:Library, ~forcats::fct_drop(.x)))
```

```{r}
summary(junk2 |> select(Sustenance:Library))
```

## Clusters: try self-organizing maps (SOM)

```{r}
wsn.som <- trainSOM(x.data = walksheds_net_attributes |>
                       select(-c(GeoUID, Type)),
                     dimension = c(5,5), 
                     verbose = TRUE, 
                     nb.save = 5, 
                     topo = "hexagonal")
```

```{r}
wsn.som
```

```{r}
plot(wsn.som, what="energy")
```

```{r}
wsn.som$clustering
```

```{r}
wsn.som$clustering |> table()
```

```{r}
plot(wsn.som, what="obs", type="hitmap")
```

```{r}
summary(wsn.som)
```

```{r}
par(mfrow = c(2,2))
plot(wsn.som, what = "obs", type = "color", variable = 1)
plot(wsn.som, what = "obs", type = "color", variable = 2)
```

```{r}
plot(wsn.som, what = "add", type = "pie", variable = walksheds_net_attributes$Type) +
  scale_fill_brewer(type = "qual") + 
  guides(fill = guide_legend(title = "Type"))
```

```{r}
quality(wsn.som)
```

```{r}
plot(superClass(wsn.som))
```

```{r}
wsn.som <- trainSOM(x.data = walksheds_net_attributes |>
                       select(-c(GeoUID, Type)),
                     dimension = c(4, 4), #2,3
                     verbose = TRUE, 
                     nb.save = 5,
                     topo = "square")
```

```{r}
plot(wsn.som, what="energy")
```

```{r}
quality(wsn.som)
```

### 2023-06-14 presentations/2023-colloque-montreal.qmd
Initial commit

---
title: |
    | Profiling 
    | the 15-min *ped shed* 
    | from the perspective of 
    | network attributes
author: "**Antonio Páez**, Belén Martín, Josh Arbess, James R. Dunn"
format: revealjs
execute: 
  cache: true
---

```{r load-packages}
#| include: false
#| cache: false

library(broom) # Convert Statistical Objects into Tidy Tibbles
library(dplyr) # A Grammar of Data Manipulation
library(evtree) # Evolutionary Learning of Globally Optimal Trees
library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(glue) # Interpreted String Literals
library(gt) # Easily Create Presentation-Ready Display Tables
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(tidyr) # Tidy Messy Data
library(tree) # Classification and Regression Trees
library(sf) # Simple Features for R
library(skimr) # Compact and Flexible Summaries of Data
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(units) # Measurement Units for R Vectors
library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
```

```{r load-data}
#| include: false

load(file = glue(here(), "/data-raw/amenities_walkshed.rda"))
load(file = glue(here(), "/data-raw/hamilton_net.rda"))
load(file = glue(here(), "/data-raw/urban_hammer_da.rda"))
data("urban_types")
load(file = glue(here(), "/data-raw/walksheds.rda"))
load(file = glue(here(), "/data-raw/walksheds_net_attributes.rda"))
```

```{r clean-net-attributes-data}
#| include: false
d <- walksheds_net_attributes |>
                  drop_na()
```

# The 15-min neighborhood

##

- A form of chronourbanism that emphasizes movement at a human scale;
- in environments that accommodate a wide range of capabilities throughout the lifespan;
- and with the aim of improving livability and health.

##

The 15-min neighborhood^[Sometimes mislabeled as the 15-min city.] is powered by:

::: {.r-fit-text style="margin-top: 50px; font-size: 3em; color: black;"}
The 15-min ped shed
:::

A pedestrian shed $\equiv$ a 15-min isochrone.

##

When combined with the environment, the 15-min ped shed is:

1. A normative...
2. ...cumulative opportunities...
3. ...accessibility index.

$$
A_i = \sum_{j \in \text{15-min ped shed}} O_j
$$

Keeping the environment constant, bigger ped sheds result in higher accessibility.

##

The maximum ped shed area^[For a given walking speed] would be in a homogeneous, isotropic surface (i.e., a featureless plain)

![](plain-978359-pxhere.com.jpg){.absolute top=150 left=200 height="450"}

::: {.notes}
image source: https://pxhere.com/en/photo/978359
:::

##

Streetscapes limit where movement can happen, which inevitably leads to smaller ped sheds.

##

And furthermore, not all streetscapes are created equal.

![](suburban-603627-pxhere.com.jpg){.absolute top=200 left=-150 height="400"}

![](urban-995263-pxhere.com.jpg){.absolute top=200 right=-150 height="400"}

::: {.notes}
image source: urban streetscape https://pxhere.com/en/photo/995263
image source: suburban streetscape https://pxhere.com/en/photo/603627
:::

# Research questions

How much do 15-min ped sheds vary in a typical Canadian city (Hamilton)?

What are their characteristics?

# Materials

##

Open source data

- Street network (OSM).
- Amenities (OSM).
- Dissemination Areas (DAs: Census).
- Urban types (City of Hamilton).

##

$n = 888$ urban and suburban zones in Hamilton (rural Hamilton excluded).

```{r study-area}
#| echo: false

ggplot() +
  geom_sf(data = st_union(urban_types)) +
  geom_sf(data = urban_types |>
            filter(Type != "Rural"),
          aes(fill = Type)) +
  geom_sf(data = urban_hammer_da,
          fill = NA) +
  scale_fill_manual(values = c("Urban" = "#619CFF", "Suburban" = "#F8766D")) +
  theme_void()
```

##

Open source software (`R`)

- Data collection and processing.
- Routing.
- Spatial analysis.
- Network analysis.

##

Open, reproducible research

[https://github.com/paezha/15-minute-neighborhood](https://github.com/paezha/15-minute-neighborhood)

# Ped sheds

##

Ped sheds calculated from zonal centroids using [`r5r`](https://ipeagit.github.io/r5r/) assuming a walking speed of 3.6 $km/h$.

```{r example-of-network}
#| echo: false
#| warning: false

max_shed_r <- (set_units(3.6, km/h) |> set_units(m/min)) * set_units(15, min)

sample_da <- data_da_2016 |>
  filter(GeoUID == "35250102")

sample_da_coords <- data.frame(geometry = st_geometry(sample_da |>
                                                        st_centroid() |>
                                                        st_coordinates() |>
                                                        st_point())) |>
  st_as_sf(crs = 4326)

sample_net <- hamilton_net$edges |>
  st_intersection(st_buffer(sample_da_coords, max_shed_r))

ggplot() +
  geom_sf(data = sample_net,
          color = "lightgray") +
  geom_sf(data = sample_da,
          fill = NA,
          linewidth = 1,
          color = "#619CFF") +
  geom_sf(data = sample_da_coords,
          color = "#619CFF",
          size = 3) +
  annotation_scale() +

  theme_void()
```

##

Summary of ped shed areas (max. ped shed area is `r round(pi * max_shed_r^2 |> units::set_units(km^2), 2)` $km^2$).

Notice the large variability within the city and between suburban and urban Hamilton.

```{r ped-shed-area-summary}
#| echo: false

global <- walksheds |>
  mutate(area = st_area(geometry),
         area = units::set_units(area, km^2) |> 
           drop_units()) |>
  st_drop_geometry() |>
  select(-c(GeoUID, Type)) |>
  skim()

by_type <- walksheds |>
  mutate(area = st_area(geometry),
         area = units::set_units(area, km^2) |> 
           drop_units()) |>
  st_drop_geometry() |>
  select(-c(GeoUID)) |>
  group_by(Type) |>
  skim()

rbind(global |> 
        select(numeric.mean:numeric.hist) |> 
        mutate(across(!ends_with("hist"), 
                      ~round(.x, 2))), 
      by_type|> 
        select(numeric.mean:numeric.hist) |> 
        mutate(across(!ends_with("hist"), 
                      ~round(.x, 2)))) |>
  mutate(Type = c("All", "Suburban", "Urban"),
         n = c(d |> nrow(),
               d |> filter(Type == "Suburban") |> nrow(),
               d |> filter(Type == "Urban") |> nrow())) |>
  select(Type, n, starts_with("numeric")) |>
  rename_with(~ str_replace(.x, "numeric.", ""), starts_with("numeric")) |>
  gt() |>
  cols_width(everything() ~ px(100)) |>
  tab_options(table.font.size = 30)
```

# Network attributes

## {.smaller .scrollable}

Global descriptors of the network.

- Transitivity (probability that the friends of my friends are my friends: minimal circuits).
- Motifs (recurrence of graph elements: repetition of patterns).
- Diameter (longest shortest path in network: how distant are the two most distant nodes).
- Radius (shortest shortest path in network: how close are the two closest nodes).
- Girth (length of shortest circle: shortest path for a node to reach itself).
- Global efficiency (harmonic mean of distances between all nodes: how easy it is on average to reach others).
- Edge density (ratio of actual to maximum edges: how connected is the network).
- Edge connectivity (How many edges need to be removed to eliminate all paths between two nodes).
- Mean distance (mean of length of all shortest paths).

##

Profiling the ped sheds based on their network attributes.

```{r train-classification-tree}
#| echo: false

# This is Tree 3 with seed 3 (46125) in 05-Walkshed-Profiles.Rmd
seed3 <- 46125
set.seed(seed3)
tree3 <- evtree(Type ~ ., 
                data = d |> select(-GeoUID))
```

```{r plot-classification-tree}
#| echo: false

# Plot the tree with rounded break labels:

ggparty(tree3) +
  geom_edge() +
  geom_edge_label(aes(label = str_replace_all(breaks_label, 
                                          "0.0207311636201516", 
                                          "0.021"))) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.0,
                     ids = c(1, 2, 5)) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.02,
                     ids = c(3, 4, 6, 7)) +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", 
                                            fill = Type),
                                        position = position_fill()),
                               xlab("Type"),
                               theme_minimal()),
                 # draw only one label for each axis
                 shared_axis_labels = TRUE,
                 # draw line between tree and legend
                 legend_separator = FALSE)
```

##

Typically suburban networks contain many unique patterns (few motifs).

```{r plot-leaf-1}
#| include: false
#| warning: false

set.seed(69717)

leaf_1_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity < 0.0207311636201516,
                             motifs < 378, 
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity < 0.021, motifs < 378") +
  theme_void()
```

```{r plot-leaf-3}
#| include: false
#| warning: false

set.seed(84451)

leaf_3_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity >= 0.0207311636201516,
                             motifs < 1024, 
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity >= 0.021, motifs < 1,024") +
  theme_void()
```

```{r display-suburban-networks}
#| echo: false

leaf_1_plot + plot_spacer() + leaf_3_plot
```

##

Typically urban networks contain isomorphic patterns that repeat (large number of motifs).

```{r plot-leaf-2}
#| include: false
#| warning: false

set.seed(44785)

leaf_2_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity < 0.0207311636201516,
                             motifs >= 378, 
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity < 0.021, motifs >= 378") +
  theme_void()
```

```{r plot-leaf-4}
#| include: false
#| warning: false

set.seed(81461)

leaf_4_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity >= 0.0207311636201516,
                             motifs >= 1024, 
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity >= 0.021, motifs >= 1,024") +
  theme_void()
```

```{r display-urban-networks}
#| echo: false

leaf_2_plot + plot_spacer() + leaf_4_plot
```

# Network attributes of ped sheds

##

Re-examining the areas of the ped sheds.

```{r add-classification-to-data}
#| include: false

d <- d |>
  mutate(leaf = case_when(transitivity < 0.0207311636201516 & motifs < 378 ~ "Leaf 1",
                          transitivity < 0.0207311636201516 & motifs >= 378 ~ "Leaf 2",
                          transitivity >= 0.0207311636201516 & motifs < 1024 ~ "Leaf 3",
                          transitivity >= 0.0207311636201516 & motifs >= 1024 ~ "Leaf 4"),
         leaf = factor(leaf))
```

```{r join-area-of-ped-sheds}
#| include: false

d <- d |>
  left_join(walksheds |>
              mutate(area = st_area(geometry),
                     area = units::set_units(area, km^2) |> 
                       drop_units()) |>
              st_drop_geometry(),
            by = c("GeoUID", "Type")) 
```

```{r area-of-ped-sheds-by-classification}
#| echo: false

d |>
  select(area, leaf) |>
  group_by(leaf) |>
  skim() |>
  mutate(across(numeric.mean:numeric.p100, 
                      ~round(.x, 2)),
         Type = c("Typically suburban", "Typically urban", "Typically suburban", "Typically urban")) |>
  select(leaf, Type, starts_with("numeric")) |>
  rename_with(~ str_replace(.x, "numeric.", ""), starts_with("numeric")) |>
  gt() |>
  cols_width(everything() ~ px(100)) |>
  tab_options(table.font.size = 30)
```

##

Ped shed area as a function of network attributes.

```{r estimate-model-area}
#| include: false
mdl <- d |>
  mutate(area = log(area),
         transitivity_2 = transitivity^2,
         motifs = motifs/1000,
         motifs_2 = (motifs)^2) |>
  lm(area ~ transitivity + transitivity_2 + motifs + motifs_2, data = _)
```

```{r tidy-results-of-model-area}
#| include: false

mdl_tidy <- mdl |>
  tidy() |>
  mutate(term = c("intercept", "transitivity", "transitivity^2", "motifs", "motifs^2"))
```

```{r plot-effects-transitivity}
#| include: false

b2 <- mdl_tidy |> filter(term == "transitivity") |> pull(estimate)
b3 <- mdl_tidy |> filter(term == "transitivity^2") |> pull(estimate)
f_transitivity <- function(x) exp(b1 + b2 * x +  b3 * x^2)

effect_transitivity <- ggplot() +
  geom_function(fun = f_transitivity,
                xlim = c(0, 0.12)) +
    ylim(c(0.6, 1.6)) +
  ggtitle("Expected area as a function of transitivity") + 
  ylab("area (sq.km)") +
  xlab("transitivity") +
  theme_minimal()
```

```{r plot-effect-motifs}
#| include: false

b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
b4 <- mdl_tidy |> filter(term == "motifs") |> pull(estimate)
b5 <- mdl_tidy |> filter(term == "motifs^2") |> pull(estimate)
f_motifs <- function(x) exp(b1 + b4 * x +  b5 * x^2)

effect_motifs <- ggplot() +
  geom_function(fun = f_motifs,
                xlim = c(0, 6.824)) +
  ylim(c(0.6, 1.6)) +
  ggtitle("Expected area as a function of motifs (in 1,000s)") + 
  ylab("area (sq.km)") +
  xlab("motifs (1,000s)") +
  theme_minimal()
```

```{r display-model-effects}
#| echo: false

effect_transitivity + effect_motifs
```

# In summary...

## Conclusions

- Ped shed area varies greatly within Hamilton.
- The size of ped sheds varies significantly with the attributes of the network.
- The effect is non-linear.
- Results suggest an optimal combination of attributes to maximize ped shed areas.

## Implications

- Typically suburban networks result in relatively small ped sheds.
- Turning suburbs into 15-min neighborhoods may require major transportation infrastructure changes.
- Alternatively, major increases in density.

## Future work

- Expand the analysis to include the environment, i.e., the opportunity landscape.

### 2023-06-14 presentations/2023-colloque-montreal.qmd
Revisions to colloque presentation after comments from BMR

@@
     | Profiling 
-    | the 15-min *ped shed* 
+    | the 15-min *pedshed* 
     | from the perspective of 
@@
 
-- A form of chronourbanism that emphasizes movement at a human scale;
-- in environments that accommodate a wide range of capabilities throughout the lifespan;
-- and with the aim of improving livability and health.
+The 15-min neighborhood^[Sometimes mislabeled as the 15-min city.].
+
+- A form of chronourbanism that emphasizes movement at a human scale...
+- ...in environments that accommodate a wide range of capabilities throughout the lifespan...
+- ...with the aim of improving livability and health.
 
@@
 
-The 15-min neighborhood^[Sometimes mislabeled as the 15-min city.] is powered by:
+The 15-min neighborhood is powered by:
 
 ::: {.r-fit-text style="margin-top: 50px; font-size: 3em; color: black;"}
-The 15-min ped shed
+The 15-min pedshed
 :::
@@
 
-When combined with the environment, the 15-min ped shed is:
+When combined with the environment, the 15-min pedshed is:
 
@@
 $$
-A_i = \sum_{j \in \text{15-min ped shed}} O_j
+A_i = \sum_{j \in \text{15-min pedshed}} O_j
 $$
 
-Keeping the environment constant, bigger ped sheds result in higher accessibility.
+Keeping the environment constant, bigger pedsheds result in higher accessibility.
 
@@
 
-The maximum ped shed area^[For a given walking speed] would be in a homogeneous, isotropic surface (i.e., a featureless plain)
+The maximum pedshed area^[For a given walking speed] would be in a homogeneous, isotropic surface (i.e., a featureless plain).
 
@@
 
-Streetscapes limit where movement can happen, which inevitably leads to smaller ped sheds.
+Streetscapes limit where movement can happen, which inevitably leads to smaller pedsheds.
 
@@
 
-How much do 15-min ped sheds vary in a typical Canadian city (Hamilton)?
+How much do 15-min pedsheds vary in a typical Canadian city (Hamilton)?
 
@@
 
-Open source data
+Open source data.
 
@@
 
-Open source software (`R`)
+Open source software (`R`).
 
@@
 
-Open, reproducible research
+Open, reproducible research.
 
@@
 
-# Ped sheds
+# pedsheds
 
@@
 
-Ped sheds calculated from zonal centroids using [`r5r`](https://ipeagit.github.io/r5r/) assuming a walking speed of 3.6 $km/h$.
+pedsheds calculated from zonal centroids using [`r5r`](https://ipeagit.github.io/r5r/) assuming a walking speed of 3.6 $km/h$.
 
@@
           size = 3) +
+  geom_sf(data = sample_da_coords |>
+            st_buffer(dist = 900),
+          fill = NA,
+          color = "black") +
+  labs(caption = "Blue polygon is DA boundary; blue dot is centroid") +
   annotation_scale() +
-
   theme_void()
@@
 
-Summary of ped shed areas (max. ped shed area is `r round(pi * max_shed_r^2 |> units::set_units(km^2), 2)` $km^2$).
+Summary of pedshed areas (max. pedshed area is `r round(pi * max_shed_r^2 |> units::set_units(km^2), 2)` $km^2$).
 
@@
 
-Profiling the ped sheds based on their network attributes.
+Profiling the pedsheds based on their network attributes.
 
@@
 
-# Network attributes of ped sheds
+# Network attributes of pedsheds
 
@@
 
-Re-examining the areas of the ped sheds.
+Re-examining the areas of the pedsheds.
 
@@
 
-Ped shed area as a function of network attributes.
+pedshed area as a function of network attributes.
 
@@
 
-- Ped shed area varies greatly within Hamilton.
-- The size of ped sheds varies significantly with the attributes of the network.
+- pedshed area varies greatly within Hamilton.
+- The size of pedsheds varies significantly with the attributes of the network.
 - The effect is non-linear.
-- Results suggest an optimal combination of attributes to maximize ped shed areas.
+- Results suggest an optimal combination of attributes to maximize pedshed areas.
 
@@
 
-- Typically suburban networks result in relatively small ped sheds.
+- Typically suburban networks result in relatively small pedsheds.
 - Turning suburbs into 15-min neighborhoods may require major transportation infrastructure changes.
 - Alternatively, major increases in density.
+- Very unlikely that the 15-min city will be achieved, but many 15-min neighborhoods may be possible.
 
@@
 - Expand the analysis to include the environment, i.e., the opportunity landscape.
+- Try other cities.

### 2023-06-15 data-raw/03-Network-Attributes-by-Walkshed.Rmd
Data objects created

@@
 ---
-title: "Network-Preprocessing"
+title: "Analysis of Network Attributes"
 author: "Antonio Paez"
 date: "`r Sys.Date()`"
-output: pdf_document
+output: html_notebook
 ---
@@
 Load packages:
-```{r}
-#library(disk.frame) # Larger-than-RAM Disk-Based Data Manipulation Framework
+```{r warning=FALSE, message=FALSE}
 library(dplyr) # A Grammar of Data Manipulation
@@
 library(igraph)
-#library(h3) # R Bindings for H3
-#ibrary(osmdata)
-#library(purrr) # Functional Programming Tools
-#library(r5r) # Rapid Realistic Routing with 'R5'
+library(skimr) # Compact and Flexible Summaries of Data
 library(sf) # Simple Features for R
@@
 library(tidyr)
-#library(skimr) # Compact and Flexible Summaries of Data
-#library(units)
-#library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
 ```
@@
 Calculate network attributes by walkshed:
-```{r}
+```{r warning=FALSE}
 # Time algorithm
@@
   walksheds_net_attributes$min_cut[i] <- min_cut(wsnet)
-  print(i)
 }

### 2023-06-15 data-raw/04-Accessibility-Calculations.Rmd
Data objects created

@@
 ---
-title: "Network-Preprocessing"
+title: "Accessibility Calculations"
 author: "Antonio Paez"
@@
 
-In this document I aim to partition the network based on walksheds.
+In this document we calculate the amenities at the level of pedshed to get cumulative accessibility scores.
 
@@
 Load packages:
-```{r}
-#library(disk.frame) # Larger-than-RAM Disk-Based Data Manipulation Framework
+```{r warning=FALSE, message=FALSE}
 library(dplyr) # A Grammar of Data Manipulation
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
-#library(igraph)
-#library(h3) # R Bindings for H3
-#ibrary(osmdata)
-#library(purrr) # Functional Programming Tools
-#library(r5r) # Rapid Realistic Routing with 'R5'
 library(sf) # Simple Features for R
-#library(sfnetworks)
-library(tidyr)
 library(skimr) # Compact and Flexible Summaries of Data
-#library(units)
+library(tidyr) # Tidy Messy Data
 library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
@@
 Intersect amenities with walksheds to obtain the cumulative opportunities:
-```{r}
+```{r warning=FALSE}
+# Time algorithm
+start.time <- Sys.time()
+
 amenities_walksheds <- amenities_points |>
@@
   st_intersection(walksheds)
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
 ```

### 2023-06-15 data-raw/05-Walkshed-Profiles.Rmd
Data objects created

@@
 ---
-title: "Profile Walksheds"
+title: "Profiling Walksheds"
 author: "Antonio Paez"
 date: "`r Sys.Date()`"
-output: pdf_document
+output: html_notebook
 ---
@@
 Load packages:
-```{r}
+```{r warning=FALSE, message=FALSE}
 library(dplyr) # A Grammar of Data Manipulation

### 2023-06-15 data-raw/01-OSM-Network-and-Routing.Rmd
Tracking noebooks and cache files

--- 2023-06-14 data-raw/01-OSM-Network-and-Walksheds.Rmd
@@
 ---
-title: "Network-Preprocessing"
+title: "Network and Routing"
 author: "Antonio Paez"
 date: "`r Sys.Date()`"
-output: pdf_document
+output: html_notebook
 ---
@@
 ```{r}
-# Allocate memory
+# Allocate memory for routing
 options(java.parameters = "-Xmx20G")
@@
 }
+
 end.time <- Sys.time()
-ttm_walk_time <- end.time - start.time
-ttm_walk_time
+time.taken <- end.time - start.time
+time.taken
 ```

### 2023-06-16 data-raw/01-OSM-Network-and-Routing.Rmd
Recreated routing and walksheds after cleaning the zoning system and adding validation points

@@
 Load packages:
-```{r}
+```{r warning=FALSE, message=FALSE}
 # Allocate memory for routing
@@
 
+Search for duplicates after the intersection:
+```{r}
+duplicate_DAs <- urban_hammer_da |> 
+  st_drop_geometry() |>
+  group_by(GeoUID) |> 
+  summarize(n = n()) |>
+  filter(n > 1)
+```
+
+Duplicates are caused by small inconsistencies between the DA boundaries and the urban/suburban boundaries, and the partitions result in slivers of DA:
+```{r}
+urban_hammer_da |>
+  filter(GeoUID %in% duplicate_DAs$GeoUID[sample.int(50, 1)]) |>
+  ggplot() +
+  geom_sf(aes(fill = GeoUID))
+```
+
+This leads to some detritus:
+```{r}
+urban_hammer_da |>
+  filter(GeoUID == "35250031") |>
+  ggplot() +
+  geom_sf(aes(fill = GeoUID)) +
+  geom_sf(data = data_da_2016,
+          fill = NA)
+```
+
+Summary of areas of the areas of DAs after the intersection. `r` is the area of the zone after the intersection divided by the area of the original DA:
+```{r}
+urban_hammer_da <- urban_hammer_da |>
+  mutate(area = st_area(geometry) |> set_units(km^2)) |>
+    mutate(r = area/Shape.Area)
+
+urban_hammer_da |>
+  filter(GeoUID %in% duplicate_DAs$GeoUID) |>
+  select(GeoUID, Shape.Area, Area..sq.km., area, r) |>
+  summary()
+```
+
+This is the distribution of the areas of all duplicated DAs
+```{r}
+urban_hammer_da |>
+  filter(GeoUID %in% duplicate_DAs$GeoUID) |>
+  ggplot(aes(x = r)) +
+  geom_histogram()
+```
+
+These are them:
+```{r}
+ggplot() +
+  geom_sf(data = urban_hammer_da |>
+            filter(GeoUID %in% duplicate_DAs$GeoUID),
+          aes(fill = drop_units(r))) +
+  geom_sf(data = data_da_2016,
+          fill = NA)
+```
+
+Split the table into DAs that had no duplicates and DAs that did:
+```{r}
+urban_hammer_dup <- urban_hammer_da |>
+  st_transform(crs = st_crs(data_da_2016)) |>
+  filter(GeoUID %in% duplicate_DAs$GeoUID)
+
+urban_hammer_nondup <- urban_hammer_da |>
+  st_transform(crs = st_crs(data_da_2016)) |>
+  filter(!(GeoUID %in% duplicate_DAs$GeoUID))
+```
+
+Choose areas to keep that are not detritus (areas that are more than 50% of the area of the original DA):
+```{r}
+urban_hammer_dup <- urban_hammer_dup |>
+  filter(r > set_units(0.50, km^2))
+```
+
+Replace the geometry by its original geometry:
+```{r}
+urban_hammer_dup <- urban_hammer_dup |>
+  st_drop_geometry() |>
+  left_join(data_da_2016 |>
+              select(GeoUID),
+            by = "GeoUID") |>
+  select(-c(area, r)) |>st_as_sf()
+```
+
+Bind with non-duplicated DAs (who get to keep the geometry they got after intersecting with the urban types):
+```{r}
+urban_hammer_da <- rbind(urban_hammer_nondup |>
+                           select(-c(area, r)),
+                         urban_hammer_dup)
+```
+
 Plot urban/suburban DAs:
@@
 
-Notice that some DAs appear twice after intersecting with urban fabric types:
+Coordinates of centroids of the DAs that intersect urban/suburban types (in lat-long). These will be the origins of trips, that is, our proxies for "neighborhoods":
 ```{r}
-urban_hammer_da |> 
-  left_join(
-    urban_hammer_da |> 
-      st_drop_geometry() |>
-      group_by(GeoUID) |>
-      summarize(n = n()),
-    by = "GeoUID") |>
-  ggplot() +
-  geom_sf(aes(fill = factor(n)))
-
+origins_da <- st_centroid(urban_hammer_da) |>
+  st_transform(crs = st_crs(4326))
 ```
 
-This is not a problem. In most cases, some DAs will get two walksheds instead of one.
+Also, sample in urban and suburban zones to replicate the number of points for DAs. We will this as a validation dataset.
+```{r}
+# For reproducibility
+seed <- 22331264
+set.seed(seed)
 
-Coordinates of centroids of the DAs that intersect urban/suburban types (in lat-long). These will be the origins of trips, that is, our proxies for "neighborhoods":
+origins_validation_suburban <- st_sample(urban_hammer_da |>
+            filter(Type.1 == "Suburban"),
+          size = rep(1, nrow(urban_hammer_da |>
+            filter(Type.1 == "Suburban")))) |>
+  st_as_sf() |>
+  mutate(Type = "Suburban")
+
+origins_validation_urban <- st_sample(urban_hammer_da |>
+            filter(Type.1 == "Urban"),
+          size = rep(1, nrow(urban_hammer_da |>
+            filter(Type.1 == "Urban")))) |>
+  st_as_sf() |>
+  mutate(Type = "Urban")
+```
+
+Bind and plot:
 ```{r}
-origins <- st_centroid(urban_hammer_da) |>
-  st_transform(crs = st_crs(4326))
+origins_validation <- rbind(origins_validation_suburban,
+                          origins_validation_urban) |>
+  st_join(data_da_2016 |>
+            select(GeoUID)) |>
+  mutate(geometry = x)
+
+ggplot() +
+  geom_sf(data = urban_types) + 
+  geom_sf(data = origins_validation,
+          aes(color = Type))
 ```
@@
 
+Convert to coordinates and store to use as destinations:
 ```{r}
@@
 
-Prepare Input Data for `r5r`. The origins are the coordinates of the DAs and the destinations the coordinates of the h3 geohashes:
+Retrieve network as a simple features object:
+```{r}
+hamilton_net <- street_network_to_sf(r5_hamilton_cma)
+```
+
+Save network as sf:
+```{r}
+save(hamilton_net, file = "hamilton_net.rda", compress = "xz")
+```
+
+Prepare Input Data for `r5r` for the DAs.
+
+The origins are the coordinates of the DAs and the destinations the coordinates of the h3 geohashes:
 ```{r prepare-inputs}
 # save origins in format expected by R5R (id, lon, lat)
-origins_i <- data.frame(ID = origins$GeoUID, 
-                        origins |>
+origins_i_da <- data.frame(ID = origins_da$GeoUID, 
+                        origins_da |>
                           st_transform(crs = 4326) |>
@@
 
-# now vaccination sites
+# save origins in format expected by R5R (id, lon, lat)
+origins_i_validation <- data.frame(ID = origins_validation$GeoUID, 
+                        origins_validation |>
+                          st_transform(crs = 4326) |>
+                          st_coordinates()) |>
+  rename(lon = X, lat = Y, id = ID) |>
+  dplyr::select(id, lon, lat)
+
+# now destinations sites
 destinations_j <- data.frame(id = urban_hammer_h3, 
@@
 
-Calculate OD Matrix for walking:
-```{r calculate walk od matrix}
+Calculate OD Matrix for walking (DAs):
+```{r calculate walk od matrix DAs}
 
@@
 chunksize = 4000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
-num_chunks = ceiling(nrow(origins_i)/chunksize)
+num_chunks = ceiling(nrow(origins_i_da)/chunksize)
 
 # create origin-destination pairs
-origins_chunks <- as.disk.frame(origins_i,
-                                outdir = "./df/origins_i",
+origins_chunks <- as.disk.frame(origins_i_da,
+                                outdir = "./df/origins_i_da",
                                 nchunks = num_chunks,
@@
                                             nchunks = 1,
-                                            outdir = "./df/output_ttm_walk",
+                                            outdir = "./df/output_ttm_walk_da",
                                             compress = 50,
@@
 
-Retrieve network as a simple features object:
-```{r}
-hamilton_net <- street_network_to_sf(r5_hamilton_cma)
+Calculate OD Matrix for walking (validation points):
+```{r calculate walk od matrix validation}
+
+# set up batching according to how many origin rows to process at one time
+chunksize = 4000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
+num_chunks = ceiling(nrow(origins_i_validation)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(origins_i_validation,
+                                outdir = "./df/origins_i_validation",
+                                nchunks = num_chunks,
+                                overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){ 
+  origins_i_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5_hamilton_cma,
+                                  origins = origins_i_chunk,
+                                  destinations = destinations_j,
+                                  mode = c("WALK"),
+                                  departure_datetime = as.POSIXct(strptime("2021-04-05 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                                  max_walk_time = 15, # minutes
+                                  max_trip_duration = 15)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "./df/output_ttm_walk_validation",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
 ```
 
-Save network as sf:
-```{r}
-save(hamilton_net, file = "hamilton_net.rda", compress = "xz")
-```
+## Extract travel time matrices
 
-## Extract travel time matrix
-
-```{r load od matrix for walk, include=FALSE}
+For DAs:
+```{r load od matrix for walk DAs, include=FALSE}
 # connect to the walking travel time matrix disk frame
-ttm_walk.disk.frame <- disk.frame("./df/output_ttm_walk")
+ttm_walk.disk.frame <- disk.frame("./df/output_ttm_walk_da")
 #opportunities_j <- as.data.frame(nyc_cb_point) |> select(GEOID10, total_emp) |> rename(toId = GEOID10, o_j = total_emp)
@@
 ```{r}
-ttm_walk <- as.data.frame(ttm_walk.disk.frame) |>
+ttm_walk_da <- as.data.frame(ttm_walk.disk.frame) |>
+  transmute(GeoUID = from_id, h3_index = to_id, travel_time = travel_time_p50)
+```
+
+For validation points:
+```{r load od matrix for walk validation points, include=FALSE}
+# connect to the walking travel time matrix disk frame
+ttm_walk.disk.frame <- disk.frame("./df/output_ttm_walk_validation")
+#opportunities_j <- as.data.frame(nyc_cb_point) |> select(GEOID10, total_emp) |> rename(toId = GEOID10, o_j = total_emp)
+```
+
+Convert disk.frame to data frame:
+```{r}
+ttm_walk_validation <- as.data.frame(ttm_walk.disk.frame) |>
   transmute(GeoUID = from_id, h3_index = to_id, travel_time = travel_time_p50)
@@
 
-Retrieve the h3 areas as multipolygons:
+For DAs: retrieve the h3 areas as multipolygons:
 ```{r}
-walksheds <- ttm_walk |> 
+start.time <- Sys.time()
+
+walksheds_da <- ttm_walk_da |> 
   group_by(GeoUID) |>
@@
             .keep = TRUE)
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
 ```
@@
 ```{r}
-walkshed_GeoUIDs <- ttm_walk |>
+walkshed_GeoUIDs_da <- ttm_walk_da |>
   group_by(GeoUID) |>
@@
 ```{r}
-walksheds <- data.frame(GeoUID = walkshed_GeoUIDs,
-                         geometry = purrr::map_df(walksheds,
+start.time <- Sys.time()
+
+walksheds_da <- data.frame(GeoUID = walkshed_GeoUIDs_da,
+                         geometry = purrr::map_df(walksheds_da,
                             tibble::as_tibble)) |>
@@
 
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
+```
+
+For validation points: retrieve the h3 areas as multipolygons:
+```{r}
+start.time <- Sys.time()
+
+walksheds_validation <- ttm_walk_validation |> 
+  group_by(GeoUID) |>
+  group_map(~ h3_set_to_multi_polygon(.x$h3_index),
+            .keep = TRUE)
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
+```
+
+Obtain the GeoUIDs of the walksheds:
+```{r}
+walkshed_GeoUIDs_validation <- ttm_walk_validation |>
+  group_by(GeoUID) |>
+  summarize(GeoUID = first(GeoUID),
+            .groups = "drop")
+```
+
+Convert walksheds with GeoUIDs of DAs to a data frame:
+```{r}
+start.time <- Sys.time()
+
+walksheds_validation <- data.frame(GeoUID = walkshed_GeoUIDs_validation,
+                         geometry = purrr::map_df(walksheds_validation,
+                            tibble::as_tibble)) |>
+  st_as_sf(crs = 4326)
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
 ```
@@
 ```{r}
-test_unit = sample.int(nrow(walksheds), 1)
+test_unit = sample.int(nrow(walksheds_da), 1)
 
@@
   geom_sf(data = data_da_2016 |>
-            filter(GeoUID == walksheds$GeoUID[test_unit]),
+            filter(GeoUID == walksheds_da$GeoUID[test_unit]),
           color = "black",
           fill= NA) + 
-  geom_sf(data = walksheds[test_unit,],
+  geom_sf(data = walksheds_da[test_unit,],
           color = "blue",
@@
 
-Join the urban type to DAs in the walksheds object:
 ```{r}
-walksheds <- walksheds |>
-  left_join(origins |>
+test_unit = sample.int(nrow(walksheds_validation), 1)
+
+ggplot() +
+  geom_sf(data = data_da_2016 |>
+            filter(GeoUID == walksheds_validation$GeoUID[test_unit]),
+          color = "black",
+          fill= NA) + 
+  geom_sf(data = walksheds_validation[test_unit,],
+          color = "blue",
+          fill = NA,
+          linewidth = 1)
+```
+
+Join the urban type to DAs in the walksheds_da object:
+```{r}
+walksheds_da <- walksheds_da |>
+  left_join(origins_da |>
               st_drop_geometry() |>
@@
 
+Join the urban type to DAs in the walksheds_validation object:
+```{r}
+walksheds_validation <- walksheds_validation |>
+  left_join(origins_validation |>
+              st_drop_geometry() |>
+              transmute(GeoUID, Type = Type),
+            by = "GeoUID")
+```
+
 Convert to factor:
 ```{r}
-walksheds <- walksheds |>
+walksheds_da <- walksheds_da |>
+  mutate(Type = factor(Type))
+
+walksheds_validation <- walksheds_validation |>
   mutate(Type = factor(Type))
@@
 ```{r}
-summary(walksheds)
+summary(walksheds_da)
+
+summary(walksheds_validation)
 ```
@@
 save(urban_hammer_da, file = "urban_hammer_da.rda", compress = "xz")
-save(ttm_walk, file = "ttm_walk.rda", compress = "xz")
-save(walksheds, file = "walksheds.rda", compress = "xz")
+save(ttm_walk_da, file = "ttm_walk_da.rda", compress = "xz")
+save(walksheds_da, file = "walksheds_da.rda", compress = "xz")
+save(ttm_walk_validation, file = "ttm_walk_validation.rda", compress = "xz")
+save(walksheds_validation, file = "walksheds_validation.rda", compress = "xz")
 ```
@@
 ```{r}
-walksheds <- walksheds |>
+walksheds_da <- walksheds_da |>
+  mutate(area = st_area(geometry))
+
+walksheds_validation <- walksheds_validation |>
   mutate(area = st_area(geometry))
@@
 
-Summarize by urban type:
+Summarize DA walksheds by urban type:
 ```{r}
-walksheds |>
+walksheds_da |>
   st_drop_geometry() |>
@@
 
+Summarize validation walksheds by urban type:
 ```{r}
-walksheds |>
+walksheds_validation |>
+  st_drop_geometry() |>
+  mutate(area = units::drop_units(area)) |>
+  group_by(Type) |>
+  skim()
+```
+
+Linear model of walksheds area as a function of urban type:
+```{r}
+walksheds_da |>
+  st_drop_geometry() |>
+  mutate(area = log(units::drop_units(set_units(area, km2)))) |>
+  lm(area ~ Type, data = _) |>
+  summary()
+```
+
+Linear model of walksheds area as a function of urban type:
+```{r}
+walksheds_validation |>
   st_drop_geometry() |>
@@
 ```{r}
-ggplot(data = walksheds |>
+ggplot(data = rbind(walksheds_da |>
+                      mutate(source = "DAs"),
+                    walksheds_validation |>
+                      mutate(source = "validation")) |>
   st_drop_geometry() |>
@@
   aes(x = Type, y = area)) +
-  geom_violin()
+  geom_violin() +
+  facet_wrap(~ source)
 ```
-
-Urban walksheds close to the mean:
-```{r}
-walksheds |>
-  st_drop_geometry() |>
-  mutate(area = units::drop_units(area)) |>
-  filter(Type == "Urban", area >= 1372092 * 0.999 & area <= 1372092 * 1.001)
-```
-
-Suburban walksheds close to the mean area of suburban walksheds:
-```{r}
-walksheds |>
-  st_drop_geometry() |>
-  mutate(area = units::drop_units(area)) |>
-  filter(Type == "Suburban", area >= 1100484 * 0.99 & area <= 1100484 * 1.01)
-```
-
-Compare the areas of walksheds using comparable areas:
-```{r}
-mock_areas <- data.frame(Type = c("Suburban",
-                          "Urban"), 
-                 area = c(1100484,
-                          1372092),
-                 x0 = c(0,0),
-                 y0 = c(0,0)) |>
-  mutate(r = sqrt(area/pi))
-```
-
-```{r}
-ggplot() +
-  geom_circle(data = mock_areas,
-              aes(x0 = x0,
-                  y0 = y0,
-                  r = r,
-                  color = Type))+
-  coord_equal()
-```

### 2023-06-16 data-raw/03-Network-Attributes-by-Walkshed.Rmd
Recreated routing and walksheds after cleaning the zoning system and adding validation points

@@
 ---
-title: "Analysis of Network Attributes"
+title: "Network Attributes by Walkshed"
 author: "Antonio Paez"
@@
 load(file = "urban_hammer_da.rda")
-load(file = "walksheds.rda")
+load(file = "walksheds_da.rda")
+load(file = "walksheds_validation.rda")
 ```
@@
 # Initialize data frame
-walksheds_net_attributes <- data.frame(GeoUID = walksheds$GeoUID,
-                                       n_nodes = numeric(889), 
-                                       transitivity = numeric(889),
-                                       motifs = numeric(889),
-                                       n_edges = numeric(889),
-                                       diameter = numeric(889), 
-                                       radius = numeric(889),
-                                       girth = numeric(889), 
-                                       global_efficiency = numeric(889),
-                                       edge_density = numeric(889),
-                                       edge_connectivity = numeric(889),
-                                       mean_distance = numeric(889),
-                                       min_cut = numeric(889))
+walksheds_da_net_vars <- data.frame(GeoUID = walksheds_da$GeoUID,
+                                       n_nodes = numeric(nrow(walksheds_da)), 
+                                       transitivity = numeric(nrow(walksheds_da)),
+                                       motifs_3 = numeric(nrow(walksheds_da)),
+                                       motifs_4 = numeric(nrow(walksheds_da)),
+                                       n_edges = numeric(nrow(walksheds_da)),
+                                       diameter = numeric(nrow(walksheds_da)), 
+                                       radius = numeric(nrow(walksheds_da)),
+                                       girth = numeric(nrow(walksheds_da)), 
+                                       global_efficiency = numeric(nrow(walksheds_da)),
+                                       edge_density = numeric(nrow(walksheds_da)),
+                                       edge_connectivity = numeric(nrow(walksheds_da)),
+                                       mean_distance = numeric(nrow(walksheds_da)),
+                                       min_cut = numeric(nrow(walksheds_da)))
 
-for(i in 1:nrow(walksheds)){
+for(i in 1:nrow(walksheds_da)){
   wsnet <- st_intersection(hamilton_edges, 
-                           walksheds[i,]) |> 
+                           walksheds_da[i,]) |> 
     st_cast(to = "LINESTRING")
@@
   E(wsnet)$weight <- E(wsnet)$length
-  walksheds_net_attributes$n_nodes[i] <- gorder(wsnet)
-  walksheds_net_attributes$transitivity[i] <- transitivity(wsnet)
-  walksheds_net_attributes$motifs[i] <- count_motifs(wsnet)
-  walksheds_net_attributes$n_edges[i] <- gsize(wsnet)
-  walksheds_net_attributes$diameter[i] <- diameter(wsnet)
-  walksheds_net_attributes$radius[i] <- radius(wsnet)
-  walksheds_net_attributes$girth[i] <- girth(wsnet)$girth
-  walksheds_net_attributes$global_efficiency[i] <- global_efficiency(wsnet)
-  walksheds_net_attributes$edge_density[i] <- edge_density(wsnet)
-  walksheds_net_attributes$edge_connectivity[i] <- edge_connectivity(wsnet)
-  walksheds_net_attributes$mean_distance[i] <- mean_distance(wsnet)
-  walksheds_net_attributes$min_cut[i] <- min_cut(wsnet)
+  walksheds_da_net_vars$n_nodes[i] <- gorder(wsnet)
+  walksheds_da_net_vars$transitivity[i] <- transitivity(wsnet)
+  walksheds_da_net_vars$motifs_3[i] <- count_motifs(wsnet, size = 3)
+  walksheds_da_net_vars$motifs_4[i] <- count_motifs(wsnet, size = 4)
+  walksheds_da_net_vars$n_edges[i] <- gsize(wsnet)
+  walksheds_da_net_vars$diameter[i] <- diameter(wsnet)
+  walksheds_da_net_vars$radius[i] <- radius(wsnet)
+  walksheds_da_net_vars$girth[i] <- girth(wsnet)$girth
+  walksheds_da_net_vars$global_efficiency[i] <- global_efficiency(wsnet)
+  walksheds_da_net_vars$edge_density[i] <- edge_density(wsnet)
+  walksheds_da_net_vars$edge_connectivity[i] <- edge_connectivity(wsnet)
+  walksheds_da_net_vars$mean_distance[i] <- mean_distance(wsnet)
+  walksheds_da_net_vars$min_cut[i] <- min_cut(wsnet)
 }
@@
 ```{r}
-walksheds_net_attributes <- walksheds_net_attributes |>
-  mutate(Type = walksheds$Type) |>
+walksheds_da_net_vars <- walksheds_da_net_vars |>
+  mutate(Type = walksheds_da$Type) |>
   select(GeoUID, Type, n_nodes:min_cut)
@@
 ```{r}
-save(walksheds_net_attributes, file = "walksheds_net_attributes.rda", compress = "xz")
+save(walksheds_da_net_vars, file = "walksheds_da_net_vars.rda", compress = "xz")
+```
+
+Calculate network attributes by walkshed for the validation set:
+```{r warning=FALSE}
+# Time algorithm
+start.time <- Sys.time()
+
+# Initialize data frame
+walksheds_validation_net_vars <- data.frame(GeoUID = walksheds_validation$GeoUID,
+                                       n_nodes = numeric(nrow(walksheds_validation)), 
+                                       transitivity = numeric(nrow(walksheds_validation)),
+                                       motifs_3 = numeric(nrow(walksheds_validation)),
+                                       motifs_4 = numeric(nrow(walksheds_validation)),
+                                       n_edges = numeric(nrow(walksheds_validation)),
+                                       diameter = numeric(nrow(walksheds_validation)), 
+                                       radius = numeric(nrow(walksheds_validation)),
+                                       girth = numeric(nrow(walksheds_validation)), 
+                                       global_efficiency = numeric(nrow(walksheds_validation)),
+                                       edge_density = numeric(nrow(walksheds_validation)),
+                                       edge_connectivity = numeric(nrow(walksheds_validation)),
+                                       mean_distance = numeric(nrow(walksheds_validation)),
+                                       min_cut = numeric(nrow(walksheds_validation)))
+
+for(i in 1:nrow(walksheds_validation)){
+  wsnet <- st_intersection(hamilton_edges, 
+                           walksheds_validation[i,]) |> 
+    st_cast(to = "LINESTRING")
+  wsnet <- as_sfnetwork(wsnet)
+  E(wsnet)$weight <- E(wsnet)$length
+  walksheds_validation_net_vars$n_nodes[i] <- gorder(wsnet)
+  walksheds_validation_net_vars$transitivity[i] <- transitivity(wsnet)
+  walksheds_validation_net_vars$motifs_3[i] <- count_motifs(wsnet, size = 3)
+  walksheds_validation_net_vars$motifs_4[i] <- count_motifs(wsnet, size = 4)
+  walksheds_validation_net_vars$n_edges[i] <- gsize(wsnet)
+  walksheds_validation_net_vars$diameter[i] <- diameter(wsnet)
+  walksheds_validation_net_vars$radius[i] <- radius(wsnet)
+  walksheds_validation_net_vars$girth[i] <- girth(wsnet)$girth
+  walksheds_validation_net_vars$global_efficiency[i] <- global_efficiency(wsnet)
+  walksheds_validation_net_vars$edge_density[i] <- edge_density(wsnet)
+  walksheds_validation_net_vars$edge_connectivity[i] <- edge_connectivity(wsnet)
+  walksheds_validation_net_vars$mean_distance[i] <- mean_distance(wsnet)
+  walksheds_validation_net_vars$min_cut[i] <- min_cut(wsnet)
+}
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
+```
+
+Join the type of urban fabric:
+```{r}
+walksheds_validation_net_vars <- walksheds_validation_net_vars |>
+  mutate(Type = walksheds_validation$Type) |>
+  select(GeoUID, Type, n_nodes:min_cut)
+```
+
+Save the data object:
+```{r}
+save(walksheds_validation_net_vars, file = "walksheds_validation_net_vars.rda", compress = "xz")
 ```
@@
 ```{r}
-walksheds_net_attributes |>
+walksheds_da_net_vars |>
+  group_by(Type) |>
+  skim()
+```
+
+```{r}
+walksheds_validation_net_vars |>
   group_by(Type) |>
@@
 
-Motifs (calculated for size 3): It indicates the presence of particular subgraphs that repeat themselves.
+Motifs (calculated for size 3 and 4): It indicates the presence of particular subgraphs that repeat themselves.
 In other words, a graph with a large number of motifs will tend to have few unique elements (some elements happen again, again, and again).

### 2023-06-21 data-raw/02-OSM-Amenities.Rmd
Data objects created
NECTAR presentation

@@
 ---
-title: "Network-Preprocessing"
+title: "OSM Extracting Amenities"
 author: "Antonio Paez"
 date: "`r Sys.Date()`"
-output: pdf_document
+output: html_notebook
 ---
+
+**NOTE:** Notebook takes minutes to run.
 
@@
 
-Obtain the coordinates of the points:
+Obtain the h3 indices at level 13 from the simple features object:
 ```{r}
-amenities_coords <- amenities_points |>
-  st_coordinates()
-
-amenities_points <- amenities_points |>
-  mutate(lon = amenities_coords[,1], lat = amenities_coords[,2])
+amenities_h3 <- geo_to_h3(amenities_points, res = 13)
 ```
 
-Use the coordinates to obtain the h3 indices at level 13:
+Add h3 indices to amenities_points data object:
 ```{r}
-amenities_points <- amenities_points |>
-  mutate(h3_index = geo_to_h3(st_coordinates(geometry),
-                              13))
+amenities_points$h3_index <- amenities_h3
 ```

### 2023-06-21 data-raw/03-Network-Attributes-by-Walkshed.Rmd
NECTAR presentation

@@
 
+**NOTE:** Notebook takes about 7 hours to run.
+
 ```{r setup, include=FALSE}
@@
 
-In this document I aim to partition the network based on walksheds.
+In this document we aim to partition the network based on walksheds.
+
+We tried smoothing the pseudo-nodes, but this process messed up with the geometry of the network which affects some of the network analysis, so we decided not to proceed.
+Instead, we use the network as is.
+Intersecting the network mutilates all edges that cross the boundary of the pedshed.
+To avoid this we keep track of the indices of the edges that intersect the pedshed and then retrieve them from the network.
 
@@
 for(i in 1:nrow(walksheds_da)){
-  wsnet <- st_intersection(hamilton_edges, 
-                           walksheds_da[i,]) |> 
-    st_cast(to = "LINESTRING")
+  # Find the intersection between the edges and the walksheds. The edges will be truncated, but the indices can be used to retrieve the full edges.
+  wsnet <- hamilton_edges |>
+    st_filter(
+      walksheds_da[i,],
+      .predicate = st_intersects)
+
+  # Extract the full edges that were found to intersect the walkshed
+  wsnet <- hamilton_edges |> 
+    filter(edge_index %in% wsnet$edge_index)
+
+  # Convert to sfnetwork
   wsnet <- as_sfnetwork(wsnet)
+
   E(wsnet)$weight <- E(wsnet)$length
@@
 for(i in 1:nrow(walksheds_validation)){
-  wsnet <- st_intersection(hamilton_edges, 
-                           walksheds_validation[i,]) |> 
-    st_cast(to = "LINESTRING")
+  # Find the intersection between the edges and the walksheds. The edges will be truncated, but the indices can be used to retrieve the full edges.
+  wsnet <- hamilton_edges |>
+    st_filter(
+      walksheds_validation[i,],
+      .predicate = st_intersects)
+
+  # Extract the full edges that were found to intersect the walkshed
+  wsnet <- hamilton_edges |> 
+    filter(edge_index %in% wsnet$edge_index)
+
+  # Convert to sfnetwork
   wsnet <- as_sfnetwork(wsnet)
+
   E(wsnet)$weight <- E(wsnet)$length
@@
 ```
+
+Example of pedshed with network:
+```{r}
+sample_walkshed <- sample.int(nrow(walksheds_da), 1)
+
+wsnet <- hamilton_edges |>
+  st_filter(
+    walksheds_da[sample_walkshed,],
+    .predicate = st_intersects)
+
+# Extract the full edges that were found to intersect the walkshed
+wsnet <- hamilton_edges |> 
+  filter(edge_index %in% wsnet$edge_index)
+
+ggplot() +
+  geom_sf(data = walksheds_da[sample_walkshed,]) +
+  geom_sf(data = wsnet) +
+  theme_void()
+
+```

### 2023-06-21 data-raw/04-Accessibility-Calculations.Rmd
NECTAR presentation

@@
 ---
+
+**NOTE:** Notebook takes about 10 minutes to run.
 
@@
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(h3)
 library(sf) # Simple Features for R
@@
 data(urban_types)
-load(file = "walksheds.rda")
+load(file = "walksheds_da.rda")
+load(file = "walksheds_validation.rda")
 ```
@@
 
-amenities_walksheds <- amenities_points |>
+amenities_walksheds_da <- amenities_points |>
   select(amenity, Category) |>
-  st_intersection(walksheds)
+  st_intersection(walksheds_da)
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
+```
+
+Validation walksheds:
+```{r warning=FALSE}
+# Time algorithm
+start.time <- Sys.time()
+
+amenities_walksheds_validation <- amenities_points |>
+  select(amenity, Category) |>
+  st_intersection(walksheds_validation)
 
@@
 ```{r}
-save(amenities_walksheds, file = "amenities_walkshed.rda", compress = "xz")
+save(amenities_walksheds_da, file = "amenities_walkshed_da.rda", compress = "xz")
+save(amenities_walksheds_validation, file = "amenities_walkshed_validation.rda", compress = "xz")
 ```
@@
 ```{r}
-amenities_walksheds_summary <- amenities_walksheds |>
+amenities_walksheds_summary <- amenities_walksheds_da |>
   drop_na() |>
@@
 
-Almost 50% of suburban population (which represent about 23% of the total population in Hamilton) live in DAs with zero amenities within the 15-min walk shed.
+About 50% of suburban population (which represent about 22% of the total population in Hamilton) live in DAs with zero amenities within the 15-min walk shed.
 
-About 18% of urban population (representing about 5% of the total population of the city) live in DAs with zero amenities within the 15-min walk shed.
+About 15% of urban population (representing about 3.7% of the total population of the city) live in DAs with zero amenities within the 15-min walk shed.
 
@@
 
-Barely 2.5% of suburban residents live in DAs with minimal 15-min walk sheds. They represent approximately half a percentage point of the total population of the city.
+Barely 2.3% of suburban residents live in DAs with minimal 15-min walk sheds. They represent approximately 1% of the total population of the city.
 
-More than 15% of urban population live in DAs with minimal 15-min walk sheds, and account for approximately 2.6% of the total population of the city.
+More than 15% of urban population live in DAs with minimal 15-min walk sheds, and account for approximately 3.8% of the total population of the city.
 
@@
 
-# 409029, 534918
+# 609054, 144675, 741580
 
-sample_walksheds <- amenities_walksheds |>
+sample_walksheds <- amenities_walksheds_da |>
   group_by(Type) |>
@@
   #         aes(fill = Type)) +
-  geom_sf(data = walksheds |>
+  geom_sf(data = walksheds_da |>
             filter((GeoUID == sample_walksheds$GeoUID[1] & Type == sample_walksheds$Type[1]) | 
@@
           color = "black")+
-  geom_sf(data = amenities_walksheds |>
+  geom_sf(data = amenities_walksheds_da |>
             filter((GeoUID == sample_walksheds$GeoUID[1] & Type == sample_walksheds$Type[1]) | 
@@
 ```
+
+Distance to amenities:
+```{r}
+load("ttm_walk_da.rda")
+```
+
+```{r}
+amenities_by_h3 <- amenities_points |>
+  st_drop_geometry() |>
+  group_by(h3_index) |>
+  summarize(n_amenities = n(),
+            .groups = "drop")
+```
+
+Join the amenities to the travel time matrix by means of the h3 indices:
+```{r}
+ttm_walk_to_amenities <- ttm_walk_da |>
+  left_join(amenities_by_h3,
+            by = "h3_index") |>
+  drop_na(n_amenities)
+```
+
+Also join the urban fabric type:
+```{r}
+ttm_walk_to_amenities <- ttm_walk_to_amenities |>
+  left_join(urban_hammer_da |>
+              st_drop_geometry() |>
+              transmute(GeoUID, Type = Type.1),
+            by = "GeoUID")
+```
+
+Summarize travel times to amenities from DA centroids:
+```{r}
+junk <- ttm_walk_to_amenities |>
+  select(GeoUID, Type, travel_time) |>
+  group_by(GeoUID, Type) |>
+  summarize(min_travel_time = min(travel_time),
+            mean_travel_time = mean(travel_time),
+            .groups = "drop")
+```
+
+Distribution of mean travel times by urban fabric:
+```{r}
+junk |>
+  ggplot(aes(x = mean_travel_time,
+             color = Type)) +
+  geom_density()
+```
+
+Distribution of min travel times by urban fabric:
+```{r}
+junk |>
+  ggplot(aes(x = min_travel_time,
+             color = Type)) +
+  geom_density()
+```

### 2023-06-21 data-raw/05-Walkshed-Profiles.Rmd
NECTAR presentation

@@
 ```{r warning=FALSE, message=FALSE}
+library(broom) # Convert Statistical Objects into Tidy Tibbles
 library(dplyr) # A Grammar of Data Manipulation
@@
 library(glue) # Interpreted String Literals
+library(gt) # Easily Create Presentation-Ready Display Tables
 library(patchwork) # The Composer of Plots
@@
 ```{r}
-load(file = "amenities_walkshed.rda")
+load(file = "amenities_walkshed_da.rda")
 load(file = "hamilton_net.rda")
@@
 data("urban_types")
-load(file = "walksheds.rda")
-load(file = "walksheds_net_attributes.rda")
+load(file = "walksheds_da.rda")
+load(file = "walksheds_da_net_vars.rda")
 ```
@@
 ```{r}
-d <- walksheds_net_attributes |>
-                  drop_na()
+d <- walksheds_da_net_vars |>
+                  drop_na() |>
+  mutate(normalized_motifs_3 = motifs_3/n_edges,
+         normalized_motifs_4 = motifs_4/n_edges)
 ```
@@
 ```{r}
-#seed1 <- sample.int(1000000, 1)
-seed1 <- 912571
-set.seed(seed1)
+seed1 <- sample.int(1000000, 1)
+#seed1 <- 908544
+set.seed(908544)
 tree1 <- evtree(Type ~ ., 
-                data = d |> select(-GeoUID))
-summary(tree1)
+                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
+plot(tree1)
 ```
@@
 ```{r}
-#seed2 <- sample.int(1000000, 1)
-seed2 <- 94766
+seed2 <- sample.int(1000000, 1)
+#seed2 <- 669402
+seed2
 set.seed(seed2)
 tree2 <- evtree(Type ~ ., 
-                data = d |> select(-GeoUID))
-summary(tree2)
+                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
+plot(tree2)
 ```
@@
 ```{r}
-#seed3 <- sample.int(1000000, 1)
-seed3 <- 46125
+seed3 <- sample.int(1000000, 1)
+#seed3 <- 853251
+seed3
 set.seed(seed3)
 tree3 <- evtree(Type ~ ., 
-                data = d |> select(-GeoUID))
-summary(tree3)
+                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
+plot(tree3)
 ```
@@
 ```{r}
-#seed4 <- sample.int(1000000, 1)
-seed4 <- 556090
+seed4 <- sample.int(1000000, 1)
+#seed4 <- 707633
+seed4
 set.seed(seed4)
 tree4 <- evtree(Type ~ ., 
-                data = d |> select(-GeoUID))
-summary(tree4)
-```
-
-```{r}
-#seed4 <- sample.int(1000000, 1)
-seed4 <- 556090
-set.seed(seed4)
-tree5 <- evtree(Type ~ ., 
-                data = d |> select(-GeoUID))
-summary(tree5)
+                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
+plot(tree4)
 ```
@@
 
-Notice that transitivity and motifs are the dominant attributes in the classification trees.
+Notice that edge_density and normalized_motifs_3 are the dominant attributes in the classification trees.
 Trees 1, 3, and 4 have the same general structure, and in fact trees 3 and 4 are identical.
@@
 
+Check the break labels of `tree3` to round:
+```{r}
+ggparty(tree3)$data$breaks_label
+```
+
+```{r add-classification-to-data}
+#| include: false
+
+d <- d |>
+  mutate(leaf = case_when(edge_density < 0.00591859212548868 & normalized_motifs_3 < 0.86734693877551 ~ "Leaf 1",
+                          edge_density < 0.00591859212548868 & normalized_motifs_3 >= 0.86734693877551 ~ "Leaf 2",
+                          edge_density >= 0.00591859212548868 & normalized_motifs_3 < 0.967741935483871 ~ "Leaf 3",
+                          edge_density >= 0.00591859212548868 & normalized_motifs_3 >= 0.967741935483871 ~ "Leaf 4"),
+         leaf = factor(leaf))
+```
+
+```{r join-area-of-ped-sheds}
+#| include: false
+
+d <- d |>
+  left_join(walksheds |>
+              mutate(area = st_area(geometry),
+                     area = units::set_units(area, km^2) |> 
+                       drop_units()) |>
+              st_drop_geometry(),
+            by = c("GeoUID", "Type")) 
+```
+
+```{r area-of-ped-sheds-by-classification}
+#| echo: false
+
+d |>
+  select(area, leaf) |>
+  group_by(leaf) |>
+  skim() |>
+  mutate(across(numeric.mean:numeric.p100, 
+                      ~round(.x, 2)),
+         Type = c("Typically suburban", "Typically urban", "Typically suburban", "Typically urban")) |>
+  select(leaf, Type, starts_with("numeric")) |>
+  rename_with(~ str_replace(.x, "numeric.", ""), starts_with("numeric")) |>
+  gt() |>
+  cols_width(everything() ~ px(100)) |>
+  tab_options(table.font.size = 30)
+```
+
+Typically suburban networks contain many unique patterns (few motifs).
+
+0.00591859212548868 & normalized_motifs_3 < 0.86734693877551 ~ "Leaf 1",
+                          edge_density < 0.00591859212548868 & normalized_motifs_3 >= 0.86734693877551 ~ "Leaf 2",
+                          edge_density >= 0.00591859212548868 & normalized_motifs_3 < 0.967741935483871 ~ "Leaf 3",
+
+```{r plot-leaf-1}
+set.seed(29919)
+
+leaf_1_plot <- st_intersection(hamilton_net$edges,
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density < 0.00591859212548868,
+                             normalized_motifs_3 < 0.86734693877551, 
+                             Type == "Suburban") |> 
+                      slice_sample(n=1) |> 
+                      pull(GeoUID))))
+
+leaf_1_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_1_plot$edge_index)
+
+#leaf_1_plot <- 
+  leaf_1_plot |>
+  ggplot() +
+  geom_sf() +
+  ggtitle("edge density < 0.006, motifs < 0.867") +
+  theme_void()
+```
+
+```{r plot-leaf-2}
+set.seed(84451)
+
+leaf_2_plot <- st_intersection(hamilton_net$edges,
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density >= 0.00591859212548868,
+                             normalized_motifs_3 < 0.967741935483871, 
+                             Type == "Suburban") |> 
+                      slice_sample(n=1) |> 
+                      pull(GeoUID))))
+
+leaf_2_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_2_plot$edge_index)
+
+leaf_2_plot |>
+  ggplot() +
+  geom_sf() +
+  ggtitle("edge density < 0.006, motifs >= 0.968") +
+  theme_void()
+```
+
+```{r display-suburban-networks}
+#| echo: false
+
+leaf_1_plot + plot_spacer() + leaf_3_plot
+```
+
+Typically urban networks contain isomorphic patterns that repeat (large number of motifs).
+
+```{r plot-leaf-3}
+set.seed(92348)
+
+leaf_1_plot <- st_intersection(hamilton_net$edges,
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density < 0.00591859212548868,
+                             normalized_motifs_3 >= 0.86734693877551, 
+                             Type == "Urban") |> 
+                      slice_sample(n=1) |> 
+                      pull(GeoUID))))
+
+leaf_1_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_1_plot$edge_index)
+
+#leaf_1_plot <- 
+  leaf_1_plot |>
+  ggplot() +
+  geom_sf() +
+  ggtitle("edge density < 0.006, motifs >= 0.867") +
+  theme_void()
+```
+
+```{r plot-leaf-4}
+set.seed(392847)
+
+leaf_4_plot <- st_intersection(hamilton_net$edges,
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density >= 0.00591859212548868,
+                             normalized_motifs_3 >= 0.967741935483871, 
+                             Type == "Urban") |> 
+                      slice_sample(n=1) |> 
+                      pull(GeoUID))))
+
+leaf_4_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_4_plot$edge_index)
+
+leaf_4_plot |>
+  ggplot() +
+  geom_sf() +
+  ggtitle("edge density < 0.006, motifs >= 0.968") +
+  theme_void()
+```
+
+```{r display-urban-networks}
+#| echo: false
+
+leaf_2_plot + plot_spacer() + leaf_4_plot
+```
+
+## Analysis of pedshed areas as a function of network attributes
+
+```{r estimate-model-area}
+#| include: false
+mdl <- d |>
+  mutate(area = log(area),
+         edge_density_2 = edge_density^2,
+         normalized_motifs_3_2 = normalized_motifs_3^2) |>
+  lm(area ~ edge_density + edge_density_2 + normalized_motifs_3, data = _)
+```
+
+```{r tidy-results-of-model-area}
+#| include: false
+
+mdl_tidy <- mdl |>
+  tidy() |>
+  mutate(term = c("intercept", "edge_density", "edge_density^2", "normalized_motifs_3"))
+```
+
+```{r plot-effects-transitivity}
+#| include: false
+
+b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
+b2 <- mdl_tidy |> filter(term == "edge_density") |> pull(estimate)
+b3 <- mdl_tidy |> filter(term == "edge_density^2") |> pull(estimate)
+f_transitivity <- function(x) exp(b1 + b2 * x +  b3 * x^2)
+
+effect_transitivity <- ggplot() +
+  geom_function(fun = f_transitivity,
+                xlim = c(0, 0.4)) +
+    ylim(c(0.05, 1.9)) +
+  ggtitle("Expected area as a function of transitivity") + 
+  ylab("area (sq.km)") +
+  xlab("transitivity") +
+  theme_minimal()
+```
+
+```{r plot-effect-motifs}
+#| include: false
+
+b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
+b4 <- mdl_tidy |> filter(term == "normalized_motifs_3") |> pull(estimate)
+f_motifs <- function(x) exp(b1 + b4 * x)
+
+effect_motifs <- ggplot() +
+  geom_function(fun = f_motifs,
+                xlim = c(0.45, 1.1)) +
+  ylim(c(0.05, 1.9)) +
+  ggtitle("Expected area as a function of normalized motifs (size = 3)") + 
+  ylab("area (sq.km)") +
+  xlab("normalized motifs (size = 3)") +
+  theme_minimal()
+```
+
+```{r display-model-effects}
+#| echo: false
+
+effect_transitivity + effect_motifs
+```
+
+<!--
 Check the break labels of `tree3` to round:

### 2023-06-21 presentations/2023-NECTAR-coimbra.qmd
Updated NECTAR presentation

--- 2023-06-14 presentations/2023-colloque-montreal.qmd
@@
 title: |
-    | Profiling 
-    | the 15-min *pedshed* 
-    | from the perspective of 
-    | network attributes
+    | Reality or wishful thinking:
+    | Network attributes of 15-minute neighbourhoods
 author: "**Antonio Páez**, Belén Martín, Josh Arbess, James R. Dunn"
@@
 data("urban_types")
-load(file = glue(here(), "/data-raw/walksheds.rda"))
-load(file = glue(here(), "/data-raw/walksheds_net_attributes.rda"))
+load(file = glue(here(), "/data-raw/walksheds_da.rda"))
+load(file = glue(here(), "/data-raw/walksheds_da_net_vars.rda"))
+hamilton_edges <- hamilton_net$edges
 ```
@@
 #| include: false
-d <- walksheds_net_attributes |>
-                  drop_na()
+d <- walksheds_da_net_vars |>
+  drop_na() |>
+  mutate(normalized_motifs_3 = motifs_3/n_edges,
+         normalized_motifs_4 = motifs_4/n_edges)
 ```
@@
 
-A pedestrian shed $\equiv$ a 15-min isochrone.
+A 15-min pedestrian shed $\equiv$ a 15-min walking isochrone.
 
@@
 
-global <- walksheds |>
+global <- walksheds_da |>
   mutate(area = st_area(geometry),
@@
 
-by_type <- walksheds |>
+by_type <- walksheds_da |>
   mutate(area = st_area(geometry),
@@
 
+##
+
+Intersect the network and the pedsheds to extract the subset of the network for each pedshed.
+
+```{r}
+#| include: false
+
+#seed <- sample.int(29873, 1)
+#seed
+set.seed(20512)
+
+sample_walkshed <- walksheds_da |>
+  filter(Type == "Urban") |>
+  slice_sample(n = 1)
+#  filter(GeoUID == "35250102")
+
+# Find the intersection between the edges and the walksheds. The edges will be truncated, but the indices can be used to retrieve the full edges.
+wsnet <- hamilton_edges |>
+  st_filter(
+    sample_walkshed,
+    .predicate = st_intersects)
+
+# Extract the full edges that were found to intersect the walkshed
+wsnet <- hamilton_edges |> 
+  filter(edge_index %in% wsnet$edge_index)
+
+sample_da <- urban_hammer_da |>
+  filter(GeoUID == sample_walkshed$GeoUID)
+
+urban_walkshed <- ggplot() +
+  geom_sf(data = sample_walkshed) +
+  geom_sf(data = sample_da,
+          fill = NA,
+          color = "#619CFF",
+          linewidth = 4) +
+  geom_sf(data = wsnet) +
+  theme_void()
+
+```
+
+```{r}
+#| include: false
+
+#seed <- sample.int(29873, 1)
+#seed
+#137
+
+set.seed(137)
+
+sample_walkshed <- walksheds_da |>
+  filter(Type == "Suburban") |>
+  slice_sample(n = 1)
+#  filter(GeoUID == "35250102")
+
+# Find the intersection between the edges and the walksheds. The edges will be truncated, but the indices can be used to retrieve the full edges.
+wsnet <- hamilton_edges |>
+  st_filter(
+    sample_walkshed,
+    .predicate = st_intersects)
+
+# Extract the full edges that were found to intersect the walkshed
+wsnet <- hamilton_edges |> 
+  filter(edge_index %in% wsnet$edge_index)
+
+sample_da <- urban_hammer_da |>
+  filter(GeoUID == sample_walkshed$GeoUID)
+
+suburban_walkshed <- ggplot() +
+  geom_sf(data = sample_walkshed) +
+  geom_sf(data = sample_da,
+          fill = NA,
+          color = "#F8766D",
+          linewidth = 4) +
+  coord_sf(xlim = c(-79.94532, -79.91076),
+           ylim = c(43.25742, 43.27836)) +
+  geom_sf(data = wsnet) +
+  theme_void()
+
+```
+
+```{r}
+#| echo: false
+
+urban_walkshed + suburban_walkshed
+```
+
 ## {.smaller .scrollable}
@@
 
-- Transitivity (probability that the friends of my friends are my friends: minimal circuits).
+- Edge density (ratio of actual to maximum edges: how connected is the network).
 - Motifs (recurrence of graph elements: repetition of patterns).
@@
 - Global efficiency (harmonic mean of distances between all nodes: how easy it is on average to reach others).
-- Edge density (ratio of actual to maximum edges: how connected is the network).
+- Transitivity (probability that the friends of my friends are my friends: minimal circuits).
 - Edge connectivity (How many edges need to be removed to eliminate all paths between two nodes).
@@
 
-# This is Tree 3 with seed 3 (46125) in 05-Walkshed-Profiles.Rmd
-seed3 <- 46125
-set.seed(seed3)
+#seed3 <- sample.int(1000000, 1)
+#seed3 <- 853251
+#seed3
+set.seed(853251)
 tree3 <- evtree(Type ~ ., 
-                data = d |> select(-GeoUID))
+                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
+#plot(tree3)
 ```
@@
   geom_edge_label(aes(label = str_replace_all(breaks_label, 
-                                          "0.0207311636201516", 
-                                          "0.021"))) +
+                                              c("0.00591859212548868" = "0.006",
+                                                "0.86734693877551" = "0.867",
+                                                "0.967741935483871" = "0.968")))) +
   geom_node_splitvar(aes(label = ifelse(kids == 0,
@@
 
-set.seed(69717)
+set.seed(289919)
 
 leaf_1_plot <- st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(transitivity < 0.0207311636201516,
-                             motifs < 378, 
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density < 0.00591859212548868,
+                             normalized_motifs_3 < 0.86734693877551, 
                              Type == "Suburban") |> 
                       slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
+                      pull(GeoUID))))
+
+leaf_1_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_1_plot$edge_index)
+
+leaf_1_plot <- leaf_1_plot |>
   ggplot() +
   geom_sf() +
-  ggtitle("Transitivity < 0.021, motifs < 378") +
+  ggtitle("edge density < 0.006, motifs < 0.867") +
   theme_void()
@@
 leaf_3_plot <- st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(transitivity >= 0.0207311636201516,
-                             motifs < 1024, 
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density >= 0.00591859212548868,
+                             normalized_motifs_3 < 0.967741935483871, 
                              Type == "Suburban") |> 
                       slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
+                      pull(GeoUID))))
+
+leaf_3_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_3_plot$edge_index)
+
+leaf_3_plot <- leaf_3_plot |>
   ggplot() +
   geom_sf() +
-  ggtitle("Transitivity >= 0.021, motifs < 1,024") +
+  ggtitle("edge density < 0.006, motifs >= 0.968") +
   theme_void()
@@
 
-set.seed(44785)
+set.seed(92348)
 
 leaf_2_plot <- st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(transitivity < 0.0207311636201516,
-                             motifs >= 378, 
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density < 0.00591859212548868,
+                             normalized_motifs_3 >= 0.86734693877551, 
                              Type == "Urban") |> 
                       slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
+                      pull(GeoUID))))
+
+leaf_2_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_2_plot$edge_index)
+
+leaf_2_plot <- leaf_2_plot |>
   ggplot() +
   geom_sf() +
-  ggtitle("Transitivity < 0.021, motifs >= 378") +
+  ggtitle("edge density < 0.006, motifs >= 0.867") +
   theme_void()
@@
 
-set.seed(81461)
+set.seed(392047)
 
 leaf_4_plot <- st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(transitivity >= 0.0207311636201516,
-                             motifs >= 1024, 
+             walksheds_da |> 
+  filter(GeoUID == (walksheds_da_net_vars |> 
+                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
+                      filter(edge_density >= 0.00591859212548868,
+                             normalized_motifs_3 >= 0.967741935483871, 
                              Type == "Urban") |> 
                       slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
+                      pull(GeoUID))))
+
+leaf_4_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_4_plot$edge_index)
+
+leaf_4_plot <- leaf_4_plot |>
   ggplot() +
   geom_sf() +
-  ggtitle("Transitivity >= 0.021, motifs >= 1,024") +
+  ggtitle("edge density < 0.006, motifs >= 0.968") +
   theme_void()
@@
 d <- d |>
-  mutate(leaf = case_when(transitivity < 0.0207311636201516 & motifs < 378 ~ "Leaf 1",
-                          transitivity < 0.0207311636201516 & motifs >= 378 ~ "Leaf 2",
-                          transitivity >= 0.0207311636201516 & motifs < 1024 ~ "Leaf 3",
-                          transitivity >= 0.0207311636201516 & motifs >= 1024 ~ "Leaf 4"),
+  mutate(leaf = case_when(edge_density < 0.00591859212548868 & normalized_motifs_3 < 0.86734693877551 ~ "Leaf 1",
+                          edge_density < 0.00591859212548868 & normalized_motifs_3 >= 0.86734693877551 ~ "Leaf 2",
+                          edge_density >= 0.00591859212548868 & normalized_motifs_3 < 0.967741935483871 ~ "Leaf 3",
+                          edge_density >= 0.00591859212548868 & normalized_motifs_3 >= 0.967741935483871 ~ "Leaf 4"),
          leaf = factor(leaf))
@@
 d <- d |>
-  left_join(walksheds |>
+  left_join(walksheds_da |>
               mutate(area = st_area(geometry),
@@
                 ~round(.x, 2)),
-         Type = c("Typically suburban", "Typically urban", "Typically suburban", "Typically urban")) |>
-  select(leaf, Type, starts_with("numeric")) |>
+         Type = c("Typically suburban", "Typically urban", "Typically suburban", "Typically urban"),
+         n = c(d |> filter(leaf == "Leaf 1") |> nrow(),
+               d |> filter(leaf == "Leaf 2") |> nrow(),
+               d |> filter(leaf == "Leaf 3") |> nrow(),
+               d |> filter(leaf == "Leaf 4") |> nrow())) |>
+  select(leaf, n, Type, starts_with("numeric")) |>
   rename_with(~ str_replace(.x, "numeric.", ""), starts_with("numeric")) |>
@@
   mutate(area = log(area),
-         transitivity_2 = transitivity^2,
-         motifs = motifs/1000,
-         motifs_2 = (motifs)^2) |>
-  lm(area ~ transitivity + transitivity_2 + motifs + motifs_2, data = _)
+         edge_density_2 = edge_density^2,
+         normalized_motifs_3_2 = normalized_motifs_3^2) |>
+  lm(area ~ edge_density + edge_density_2 + normalized_motifs_3, data = _)
 ```
@@
   tidy() |>
-  mutate(term = c("intercept", "transitivity", "transitivity^2", "motifs", "motifs^2"))
+  mutate(term = c("intercept", "edge_density", "edge_density^2", "normalized_motifs_3"))
 ```
@@
 
-b2 <- mdl_tidy |> filter(term == "transitivity") |> pull(estimate)
-b3 <- mdl_tidy |> filter(term == "transitivity^2") |> pull(estimate)
+b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
+b2 <- mdl_tidy |> filter(term == "edge_density") |> pull(estimate)
+b3 <- mdl_tidy |> filter(term == "edge_density^2") |> pull(estimate)
 f_transitivity <- function(x) exp(b1 + b2 * x +  b3 * x^2)
 
-effect_transitivity <- ggplot() +
+effect_edge_density <- ggplot() +
   geom_function(fun = f_transitivity,
-                xlim = c(0, 0.12)) +
-    ylim(c(0.6, 1.6)) +
-  ggtitle("Expected area as a function of transitivity") + 
+                xlim = c(0, 0.4)) +
+    ylim(c(0.05, 1.9)) +
+  ggtitle("Expected area as a function of edge density") + 
   ylab("area (sq.km)") +
-  xlab("transitivity") +
+  xlab("edge density") +
   theme_minimal()
@@
 b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
-b4 <- mdl_tidy |> filter(term == "motifs") |> pull(estimate)
-b5 <- mdl_tidy |> filter(term == "motifs^2") |> pull(estimate)
-f_motifs <- function(x) exp(b1 + b4 * x +  b5 * x^2)
+b4 <- mdl_tidy |> filter(term == "normalized_motifs_3") |> pull(estimate)
+f_motifs <- function(x) exp(b1 + b4 * x)
 
@@
   geom_function(fun = f_motifs,
-                xlim = c(0, 6.824)) +
-  ylim(c(0.6, 1.6)) +
-  ggtitle("Expected area as a function of motifs (in 1,000s)") + 
+                xlim = c(0.45, 1.1)) +
+  ylim(c(0.05, 1.9)) +
+  ggtitle("Expected area as a function of normalized motifs") + 
   ylab("area (sq.km)") +
-  xlab("motifs (1,000s)") +
+  xlab("normalized motifs (size = 3)") +
   theme_minimal()
@@
 
-effect_transitivity + effect_motifs
+effect_edge_density + effect_motifs
 ```
@@
 - The size of pedsheds varies significantly with the attributes of the network.
-- The effect is non-linear.
-- Results suggest an optimal combination of attributes to maximize pedshed areas.
+- Results suggest pedshed areas maximized when many connections built with repetitive patterns.
 
@@
 - Expand the analysis to include the environment, i.e., the opportunity landscape.
-- Try other cities.
+- Examine other cities.

### 2023-07-06 Fifteen-minute-neighborhoods/Fifteen-minute-neighborhoods-title-page.Rmd
Updated analysis and files for paper.

@@
 
-bibliography: bibliography.bib
+bibliography: 
+  - '`r system(kpsewhich bibliography.bib)`'
 biblio-style: spphys

### 2023-07-06 data-raw/03-Network-Attributes-by-Walkshed.Rmd
Updated analysis and files for paper.

@@
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
-library(igraph)
+library(igraph) # Network Analysis and Visualization
+library(patchwork) # The Composer of Plots
 library(skimr) # Compact and Flexible Summaries of Data
 library(sf) # Simple Features for R
-library(sfnetworks)
-library(tidyr)
+library(sfnetworks) # Tidy Geospatial Networks
+library(tidyr) # Tidy Messy Data
 ```
@@
 
-Motfis (from igraph help pages)
-```{r}
-g <- sample_pa(100)
-motifs(g, 3)
-count_motifs(g, 3)
-sample_motifs(g, 3)
-
-```
-
-```{r}
-pattern <- graph.full(3)
-my.graph <- grg.game(100, 0.2) 
-```
-
 Example of pedshed with network:
@@
 ```
+
+Explore the motifs. First, extract a walkshed and convert it to sfnetworks class:
+```{r}
+sample_walkshed <- sample.int(nrow(walksheds_da), 1)
+
+wsnet <- hamilton_edges |>
+  st_filter(
+    walksheds_da[sample_walkshed,],
+    .predicate = st_intersects)
+
+# Extract the full edges that were found to intersect the walkshed
+wsnet <- hamilton_edges |> 
+  filter(edge_index %in% wsnet$edge_index)
+
+# Convert to sfnetwork
+wsnet <- as_sfnetwork(wsnet)
+
+ggplot() +
+  geom_sf(data = walksheds_da[sample_walkshed,]) +
+  geom_sf(data = wsnet |>
+            activate(edges) |>
+            st_as_sf()
+  ) +
+  theme_void()
+
+```
+
+Note that there are 16 classes of directed isomorphic graphs of size 3. Here numbers 1 and 15 are obtained:
+```{r}
+# Size is the number of nodes, number is the isomorphic class, directed true
+g1 <- graph_from_isomorphism_class(size = 3, number = 1, directed = TRUE)
+g15 <- graph_from_isomorphism_class(size = 3, number = 15, directed = TRUE)
+```
+
+Plot isomorphism:
+```{r}
+ggraph(g1, layout = "linear", circular = TRUE) +
+  geom_edge_fan2(arrow = arrow(length = unit(5, 'mm')),
+                 end_cap = circle(5, 'mm'),
+                 lineend = "round", 
+                 width = 1) + 
+  geom_node_point(size = 5) +
+  coord_equal() +
+  theme_void()
+
+```
+
+Obtain all directed isomorphisms of size 3:
+```{r}
+size_3_iso <- purrr::map(1:15,
+                         ~graph_from_isomorphism_class(size = 3, 
+                                                       number = .x, 
+                                                       directed = TRUE))
+```
+
+```{r}
+ggraph(size_3_iso[[10]], layout = "linear", circular = TRUE) +
+  geom_edge_fan2(arrow = arrow(length = unit(5, 'mm')),
+                 end_cap = circle(5, 'mm'),
+                 lineend = "round", 
+                 width = 1) + 
+  geom_node_point(size = 5) +
+  coord_equal() +
+  theme_void()
+```
+
+Motifs of size 3 in wsnet:
+```{r}
+motifs(wsnet)
+```
+
+The most common by far is class 10.
+
+```{r}
+count_motifs(wsnet)
+```
+
+Initialize a data frame to save the motif counts:
+```{r}
+walksheds_da_motif_3_count <- data.frame(matrix(0, nrow = nrow(walksheds_da), ncol = 16))
+
+walksheds_da_motif_3_count <- walksheds_da_motif_3_count |>
+  rename_with(~ paste0("iso_class_", c(0:15)), starts_with("X"))
+
+DA <- numeric(length = nrow(walksheds_da))
+```
+
+Compute the number of motifs of each class in the walksheds:
+```{r warning=FALSE}
+# Time algorithm
+start.time <- Sys.time()
+
+for(i in 1:nrow(walksheds_da)){
+  wsnet <- hamilton_edges |>
+    st_filter(
+      walksheds_da[i,],
+      .predicate = st_intersects)
+
+  wsnet <- hamilton_edges |> 
+    filter(edge_index %in% wsnet$edge_index)
+
+  # Convert to sfnetwork
+  wsnet <- as_sfnetwork(wsnet)
+  walksheds_da_motif_3_count[i,] <- motifs(wsnet)
+  DA[i] <- walksheds_da$GeoUID
+}
+
+walksheds_da_motif_3_count <- cbind(DA,
+                                    walksheds_da_motif_3_count)
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
+```
+
+Time difference:
+```{r}
+time.taken
+```
+
+Joing urban fabric type:
+```{r}
+walksheds_da_motif_3_count <- walksheds_da_motif_3_count |>
+  rename(GeoUID = DA) |>
+  left_join(walksheds_da |>
+              st_drop_geometry(),
+            by = "GeoUID") |>
+  select(GeoUID,
+         Type,
+         starts_with("iso_class"))
+```
+
+Save the data object:
+```{r}
+save(walksheds_da_motif_3_count, file = "walksheds_da_motif_3_count.rda", compress = "xz")
+```
+
+Summary of motif counts:
+```{r}
+summary(walksheds_da_motif_3_count)
+```
+
+The only motifs present are class 10 and class 15:
+```{r}
+iso_3_10_plot <- ggraph(size_3_iso[[10]], layout = "linear", circular = TRUE) +
+  geom_edge_fan2(arrow = arrow(length = unit(5, 'mm')),
+                 end_cap = circle(5, 'mm'),
+                 lineend = "round", 
+                 width = 1) + 
+  geom_node_point(size = 5) +
+  coord_equal() +
+  theme_void()
+
+iso_3_15_plot <- ggraph(size_3_iso[[15]], layout = "linear", circular = TRUE) +
+  geom_edge_fan2(arrow = arrow(length = unit(5, 'mm')),
+                 end_cap = circle(5, 'mm'),
+                 lineend = "round", 
+                 width = 1) + 
+  geom_node_point(size = 5) +
+  coord_equal() +
+  theme_void()
+
+iso_3_10_plot + iso_3_15_plot
+```
+
+Check walksheds with low number of motifs (normalized):
+```{r}
+d <- walksheds_da_net_vars |>
+                  drop_na() |>
+  mutate(normalized_motifs_3 = motifs_3/n_edges,
+         normalized_motifs_4 = motifs_4/n_edges)
+
+d |> filter(normalized_motifs_3 < 0.7)
+```
+
+Choose a walkshed with low number of motifs:
+```{r}
+ID <- "35250947" #"35250992"# "35250317"
+
+wsnet <- hamilton_edges |>
+  st_filter(
+    walksheds_da |> filter(GeoUID == ID),
+    .predicate = st_intersects)
+
+# Extract the full edges that were found to intersect the walkshed
+wsnet <- hamilton_edges |> 
+  filter(edge_index %in% wsnet$edge_index)
+
+# Convert to sfnetwork
+wsnet <- as_sfnetwork(wsnet)
+```
+
+```{r}
+ggplot()  +
+  geom_sf(data = walksheds_da |>
+            filter(GeoUID == ID),
+          color = "blue",
+          fill = "blue") +
+  geom_sf(data = urban_hammer_da |> 
+            filter(GeoUID == ID), 
+          fill = "red") +
+  geom_sf(data = wsnet |> 
+            activate(edges) |> 
+            st_as_sf())
+```
+
+Find the motifs of class 10 and 15 in wsnet:
+```{r}
+iso_class_10_list <- subgraph_isomorphisms(size_3_iso[[10]], 
+                                           wsnet, 
+                                           method = "vf2")
+
+iso_class_15_list <- subgraph_isomorphisms(size_3_iso[[15]], 
+                                           wsnet, 
+                                           method = "vf2")
+```
+
+The elements in the list that results are triads of nodes. 
+```{r}
+subgraph(wsnet, vids = iso_class_10_list[[1]]) |>
+  as_sfnetwork()
+```
+
+Extract the isomorphs of class 10 from the network:
+```{r message=FALSE}
+# This is the first one in the list
+motifs_class_10 <- subgraph(wsnet, vids = iso_class_10_list[[1]]) |>
+  as_sfnetwork()
+
+# Join every other isomorph in the list
+for(i in 2:length(iso_class_10_list)){
+  motifs_class_10 <- st_network_join(motifs_class_10,
+                   subgraph(wsnet, vids = iso_class_10_list[[i]]) |>
+                     as_sfnetwork())
+}
+```
+
+Extract the isomorphs of class 15 from the network:
+```{r message=FALSE}
+# This is the first one in the list
+motifs_class_15 <- subgraph(wsnet, vids = iso_class_15_list[[1]]) |>
+  as_sfnetwork()
+
+# Join every other isomorph in the list
+for(i in 2:length(iso_class_15_list)){
+  motifs_class_15 <- st_network_join(motifs_class_15,
+                   subgraph(wsnet, vids = iso_class_15_list[[i]]) |>
+                     as_sfnetwork())
+}
+```
+
+Plot the motifs of class 10:
+```{r}
+ggplot() + 
+  geom_sf(data = walksheds_da |>
+            filter(GeoUID == ID)) +
+  geom_sf(data = wsnet |>
+            activate(edges) |>
+            st_as_sf()) +
+  geom_sf(data = motifs_class_10 |>
+            activate(edges) |>
+            st_as_sf(),
+          color = "red") +
+  geom_sf(data = motifs_class_15 |>
+            activate(edges) |>
+            st_as_sf(),
+          color = "blue") +
+  theme_void()
+```
+
+Almost all elements of the network are part of a motif. But the number of motifs is low, which suggests that a large number of dead ends tends to reduce the number of motifs.
+
+Degree distribution:
+```{r}
+ggplot(data = data.frame(degree = degree(wsnet)) |>
+         mutate(degree = factor(degree,
+                                ordered = TRUE))|>
+         group_by(degree) |>
+         summarize(count = n())) +
+  geom_col(aes(x = degree, y = count))
+```
+
+Check walksheds with high number of motifs (normalized):
+```{r}
+d |> filter(normalized_motifs_3 > 1.07)
+```
+
+Choose a walkshed with high number of motifs:
+```{r}
+ID <- "35250675" #
+
+wsnet <- hamilton_edges |>
+  st_filter(
+    walksheds_da |> filter(GeoUID == ID),
+    .predicate = st_intersects)
+
+# Extract the full edges that were found to intersect the walkshed
+wsnet <- hamilton_edges |> 
+  filter(edge_index %in% wsnet$edge_index)
+
+# Convert to sfnetwork
+wsnet <- as_sfnetwork(wsnet)
+```
+
+Find the motifs of class 10 and 15 in wsnet:
+```{r}
+iso_class_10_list <- subgraph_isomorphisms(size_3_iso[[10]], 
+                                           wsnet, 
+                                           method = "vf2")
+
+iso_class_15_list <- subgraph_isomorphisms(size_3_iso[[15]], 
+                                           wsnet, 
+                                           method = "vf2")
+```
+
+Extract the isomorphs of class 10 from the network:
+```{r message=FALSE}
+# This is the first one in the list
+motifs_class_10 <- subgraph(wsnet, vids = iso_class_10_list[[1]]) |>
+  as_sfnetwork()
+
+# Join every other isomorph in the list
+for(i in 2:length(iso_class_10_list)){
+  motifs_class_10 <- st_network_join(motifs_class_10,
+                   subgraph(wsnet, vids = iso_class_10_list[[i]]) |>
+                     as_sfnetwork())
+}
+```
+
+Extract the isomorphs of class 15 from the network:
+```{r message=FALSE}
+# This is the first one in the list
+motifs_class_15 <- subgraph(wsnet, vids = iso_class_15_list[[1]]) |>
+  as_sfnetwork()
+
+# Join every other isomorph in the list
+for(i in 2:length(iso_class_15_list)){
+  motifs_class_15 <- st_network_join(motifs_class_15,
+                   subgraph(wsnet, vids = iso_class_15_list[[i]]) |>
+                     as_sfnetwork())
+}
+```
+
+Plot the motifs of class both classes:
+```{r}
+ggplot() + 
+  geom_sf(data = walksheds_da |>
+            filter(GeoUID == ID)) +
+  geom_sf(data = wsnet |>
+            activate(edges) |>
+            st_as_sf()) +
+  geom_sf(data = motifs_class_10 |>
+            activate(edges) |>
+            st_as_sf(),
+          color = "red") +
+  geom_sf(data = motifs_class_15 |>
+            activate(edges) |>
+            st_as_sf(),
+          color = "blue") +
+  theme_void()
+```
+
+Almost all elements of the network are part of a motif. But the number of motifs is low, which suggests that a large number of dead ends tends to reduce the number of motifs.
+
+Degree distribution:
+```{r}
+ggplot(data = data.frame(degree = degree(wsnet)) |>
+         mutate(degree = factor(degree,
+                                ordered = TRUE))|>
+         group_by(degree) |>
+         summarize(count = n())) +
+  geom_col(aes(x = degree, y = count))
+```
+
+```{r}
+wsnet <- hamilton_edges |>
+  st_filter(
+    walksheds_da |> filter(GeoUID == ID),
+    .predicate = st_intersects)
+
+# Extract the full edges that were found to intersect the walkshed
+wsnet <- hamilton_edges |> 
+  filter(edge_index %in% wsnet$edge_index)
+
+# Convert to sfnetwork
+wsnet <- as_sfnetwork(wsnet)
+```
+
+Compute the degree distribution in the walksheds:
+```{r warning=FALSE}
+# Time algorithm
+start.time <- Sys.time()
+
+# Initialize a dataframe 
+walksheds_da_degree_dist <- data.frame(GeoUID = numeric(), degree = numeric())
+
+for(i in 1:nrow(walksheds_da)){
+  wsnet <- hamilton_edges |>
+    st_filter(
+      walksheds_da[i,],
+      .predicate = st_intersects)
+
+  wsnet <- hamilton_edges |> 
+    filter(edge_index %in% wsnet$edge_index)
+
+  # Convert to sfnetwork
+  wsnet <- as_sfnetwork(wsnet)
+  walksheds_da_degree_dist <- rbind(walksheds_da_degree_dist,
+                                    data.frame(GeoUID = walksheds_da$GeoUID[i],
+                                               degree = degree(wsnet)))
+}
+
+end.time <- Sys.time()
+time.taken <- end.time - start.time
+time.taken
+```
+
+Time difference:
+```{r}
+time.taken
+```
+
+Join urban fabric types:
+```{r}
+walksheds_da_degree_dist <- walksheds_da_degree_dist |> 
+  left_join(walksheds_da |>
+              st_drop_geometry(),
+            by = "GeoUID")
+```
+
+Save object:
+```{r}
+save(walksheds_da_degree_dist, 
+     file = "walksheds_da_degree_dist.rda", 
+     compress = "xz")
+```
+
+```{r}
+walksheds_da_degree_dist |>
+  mutate(degree = factor(degree, 
+                         ordered = TRUE)) |>
+         group_by(Type, degree) |>
+         summarize(count = n(),
+                   .groups = "drop") |>
+  ggplot() +
+  geom_col(aes(x = degree, y = count, fill = Type), position = "dodge")
+```

### 2023-07-06 data-raw/04-Accessibility-Calculations.Rmd
Updated analysis and files for paper.

@@
 
+Distribution of travel times by fabric type:
+```{r}
+ttm_walk_to_amenities |> 
+  ggplot() +
+  geom_histogram(aes(x = travel_time, fill = Type),
+                 position = "dodge")
+```
+
 Summarize travel times to amenities from DA centroids:
@@
 ```
+
+See Millward, Spinney, and Scott (2013) for some distance-decay functions for active transport.
+The decay curves are very steep, so the difference in minimum time and mean time to destinations by urban fabric can potentially have a large impact on the decision to walk.
+
+For all destinations this is the equation:
+$$
+\log(Y) = 1.282 - 0.048 \cdot t
+$$
+For all destinations this is the equation:
+$$
+\log(Y) = 1.035 - 0.032 \cdot t
+$$
+
+$Y$ is the percentage of trips and $t$ is the time in minutes.
+
+```{r}
+ggplot() +
+  geom_function(fun = function(x) exp(1.282)*exp(-0.048)^x,
+                color = "blue") +
+  geom_function(fun = function(x) exp(1.035) * exp(-0.032)^x,
+                color = "red") +
+  #geom_vline(xintercept =  4.37) +
+  #geom_vline(xintercept = 6.02) +
+  xlim(1, 15)
+```
+
+The median of the minimum distance to any amenity in urban walksheds is 4.00 and in suburban sheds is 6.00.
+If we use the impedance function for all destinations, this means that the probability of taking an active trip is about 40\% lower in suburban sheds:
+```{r}
+exp(1.282)*exp(-0.048)^6.0/exp(1.282)*exp(-0.048)^4.0
+```

### 2023-07-06 paper/paper.Rmd
Updated analysis and files for paper.

---
title: "Which fifteen-minutes neighborhoods are dead-ends? An analysis of the network attributes of fifteen-minute pedsheds"
author:
  - name: Author One
    email: a1@example.com
    affiliation: University One
    correspondingauthor: true
    footnote: 1
  - name: Author Two
    email: a2@example.com
    affiliation: University Two
  - name: Author Three
    email: a3@example.com
    affiliation: University One
    footnote: 2
  - name: Author Four
    email: a4@example.com
    affiliation: University One
    footnote: 2
address:
  - code: University One
    organization: Department
    addressline: 1 main street
    city: City
    state: State
    postcode: 123456
    country: Country
  - code: University Two
    organization: Department
    addressline: 2 main street
    postcode: 2054
    city: City
    country: Country
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  Fifteen-minutes neihborhoods, a form of normative chronourbanism based on cumulative opportunities, has gained recognized as a way to reduce the need for motorized travel, and increase the livability, convenience, and health of the public. At the core of this concept is a pedshed, an area defined by the walkable isochrone of the eponymous fifteen minutes. As the idea of fifteen-minutes neighborhoods develops traction in policy and planning circles, it seems timely to revisit the way street network design can support—or obstruct—the stated goal of preserving or creating walkable neighborhoods with essential amenities. In this paper we examine a sample ($n=834$) of fifteen-minutes pedsheds in Hamilton, a medium-sized city in Canada, and how their sizes relate to the attributes of the transportation network. The analysis reveals that network design in suburban Hamilton conspires against the creating of fifteen-minutes neighborhoods. Much of urban Hamilton, in contrast, already has the characteristics of fifteen-minutes neighborhoods. The research points to elements of network design that can help to discriminate between candidate neighborhoods and dead-ends, and that can provide parameters for the design of new developments.

keywords: 
  - Fifteen-minutes neighborhoods
  - Pedshed
  - Walkability
  - Accessibility
  - Network analysis
journal: "Sustainable Cities and Society"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography:
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

# Introduction

"Il faut oublier la traversée de Paris d'est en ouest en voiture"^[ "We must forget about crossing Paris from east to west by car"] [@alimi2020].

With this declaration during her reelection campaign in 2020, Anne Hidalgo brought international attention to the concept of walkable fifteen-minutes neighborhoods.

<!-- 
Automotive retrofits in traditional city centers:

@bass2019automotive
-->

@knight2018walkable
@liu2022toward
@pozoukidou2021fifteen
@weng2019fifteen

# References {-}

### 2023-07-06 paper/paper.tex
Updated analysis and files for paper.

\documentclass[preprint, 3p,
authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

  \journal{Sustainable Cities and Society} % Sets Journal name

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Which fifteen-minutes neighborhoods are dead-ends? An analysis of the network attributes of fifteen-minute pedsheds},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\begin{frontmatter}

  \title{Which fifteen-minutes neighborhoods are dead-ends? An analysis
of the network attributes of fifteen-minute pedsheds}
    \author[University One]{Author One%
  \corref{cor1}%
  \fnref{1}}
   \ead{a1@example.com} 
    \author[University Two]{Author Two%
  %
  }
   \ead{a2@example.com} 
    \author[University One]{Author Three%
  %
  \fnref{2}}
   \ead{a3@example.com} 
    \author[University One]{Author Four%
  %
  \fnref{2}}
   \ead{a4@example.com} 
      \affiliation[University One]{
    organization={Department},addressline={1 main
street},city={City},postcode={123456},state={State},country={Country},}
    \affiliation[University Two]{
    organization={Department},addressline={2 main
street},city={City},postcode={2054},country={Country},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

  \begin{abstract}
  Fifteen-minutes neihborhoods, a form of normative chronourbanism based
  on cumulative opportunities, has gained recognized as a way to reduce
  the need for motorized travel, and increase the livability,
  convenience, and health of the public. At the core of this concept is
  a pedshed, an area defined by the walkable isochrone of the eponymous
  fifteen minutes. As the idea of fifteen-minutes neighborhoods develops
  traction in policy and planning circles, it seems timely to revisit
  the way street network design can support---or obstruct---the stated
  goal of preserving or creating walkable neighborhoods with essential
  amenities. In this paper we examine a sample (\(n=834\)) of
  fifteen-minutes pedsheds in Hamilton, a medium-sized city in Canada,
  and how their sizes relate to the attributes of the transportation
  network. The analysis reveals that network design in suburban Hamilton
  conspires against the creating of fifteen-minutes neighborhoods. Much
  of urban Hamilton, in contrast, already has the characteristics of
  fifteen-minutes neighborhoods. The research points to elements of
  network design that can help to discriminate between candidate
  neighborhoods and dead-ends, and that can provide parameters for the
  design of new developments.
  \end{abstract}
    \begin{keyword}
    Fifteen-minutes
neighborhoods \sep Pedshed \sep Walkability \sep Accessibility \sep 
    Network analysis
  \end{keyword}

 \end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

``Il faut oublier la traversée de Paris d'est en ouest en
voiture''\footnote{ ``We must forget about crossing Paris from east to
  west by car''} \citep{alimi2020}.

\citet{knight2018walkable} \citet{liu2022toward}
\citet{pozoukidou2021fifteen} \citet{weng2019fifteen}

\renewcommand\refname{References}
\bibliography{../bibliography/bibliography.bib}

\end{document}

### 2023-07-06 Fifteen-minute-neighborhoods/Fifteen-minute-neighborhoods-title-page.Rmd
Repaired renv.lock

--- 2023-06-14 Fifteen-minute-neighborhoods/Fifteen-minute-neighborhoods-title-page.Rmd
@@
 
-bibliography: bibliography.bib
+bibliography: 
+  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
 biblio-style: spphys

### 2024-10-18 data-raw/00-test-r5r.qmd
Verified that the project is reproducible after updating renv

---
title: "Test r5r"
author: "AP"
format: html
---

## Test package

This notebook is used to test the package [{r5r}](), which is used for routing and generating isochrones.

## Running Code

It is important to allocate memory before launching {r5r}:
```{r}
# allocate RAM memory to Java
options(java.parameters = "-Xmx2G")
```

1) build transport network, pointing to the path where OSM and GTFS data are stored
```{r}
library(r5r)
path <- system.file("extdata/poa", package = "r5r")
r5r_core <- setup_r5(data_path = path, verbose = FALSE)
```

2) load origin/destination points and set arguments
```{r}
points <- read.csv(system.file("extdata/poa/poa_hexgrid.csv", package = "r5r"))
mode <- c("WALK", "TRANSIT")
max_walk_time <- 30   # minutes
max_trip_duration <- 60 # minutes
departure_datetime <- as.POSIXct("13-05-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")
```

3.1) calculate a travel time matrix
```{r}
ttm <- travel_time_matrix(r5r_core = r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_time = max_walk_time,
                          max_trip_duration = max_trip_duration)
```

3.2) or get detailed info on multiple alternative routes
```{r}
det <- detailed_itineraries(r5r_core = r5r_core,
                            origins = points[370, ],
                            destinations = points[200, ],
                            mode = mode,
                            departure_datetime = departure_datetime,
                            max_walk_time = max_walk_time,
                            max_trip_duration = max_trip_duration,
                            shortest_path = FALSE,
                            drop_geometry = FALSE)
```

4) Calculate number of schools accessible within 20 minutes 
```{r}
access <- accessibility(r5r_core = r5r_core,
                        origins = points,
                        destinations = points,
                        opportunities_colname = "schools",
                        decay_function = "step",
                        cutoffs = 21,
                        mode =  c("WALK", "TRANSIT"),
                        verbose = FALSE)
```

5} Calculate isochrones
```{r}
ttm <- travel_time_matrix(r5r_core = r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_time = max_walk_time,
                          max_trip_duration = max_trip_duration)

iso <- isochrone(
  r5r_core = r5r_core,
  origins = points,
  mode = "WALK",
  mode_egress = "WALK",
  cutoffs = c(0, 15),
  sample_size = 0.8,
  departure_datetime = departure_datetime,
  time_window = 10L,
  max_walk_time = 15,
  max_bike_time = 30,
  max_car_time = 120,
  max_trip_duration = 120L,
  walk_speed = 3.6,
  bike_speed = 12,
  max_rides = 3,
  max_lts = 2,
  draws_per_minute = 5L,
  n_threads = Inf,
  verbose = FALSE,
  progress = TRUE
)
```

### 2024-10-18 data-raw/01-OSM-Network-and-Routing.Rmd
NECTAR presentation
Verified that the project is reproducible after updating renv

@@
 ---
+
+**NOTE:** Notebook takes about 6 hours to run.
 
@@
 library(h3) # R Bindings for H3
+library(here)
 library(purrr) # Functional Programming Tools
@@
 data("data_da_2016")
+
 ```
@@
 
-Also, sample in urban and suburban zones to replicate the number of points for DAs. We will this as a validation dataset.
+Also, sample in urban and suburban zones to replicate the number of points for DAs. We will this as a validation dataset. These are non-centroid origins:
 ```{r}
@@
 ```{r}
-save(hamilton_net, file = "hamilton_net.rda", compress = "xz")
+save(hamilton_net, 
+     file = paste0(here::here(), 
+                   "/data/hamilton_net.rda"), 
+                   compress = "xz")
 ```
@@
 
-The origins are the coordinates of the DAs and the destinations the coordinates of the h3 geohashes:
+The origins are coordinates of centroids and non-centroids points in DAs and the destinations the coordinates of the h3 geohashes:
 ```{r prepare-inputs}
@@
 ```{r}
-save(urban_hammer_da, file = "urban_hammer_da.rda", compress = "xz")
-save(ttm_walk_da, file = "ttm_walk_da.rda", compress = "xz")
-save(walksheds_da, file = "walksheds_da.rda", compress = "xz")
-save(ttm_walk_validation, file = "ttm_walk_validation.rda", compress = "xz")
-save(walksheds_validation, file = "walksheds_validation.rda", compress = "xz")
+save(urban_hammer_da, 
+     file = paste0(here::here(), 
+                   "/data/urban_hammer_da.rda"), 
+     compress = "xz")
+save(ttm_walk_da, 
+     file = paste0(here::here(), 
+                   "/data/ttm_walk_da.rda"), 
+     compress = "xz")
+save(walksheds_da, 
+     file = paste0(here::here(), 
+                   "/data/walksheds_da.rda"), 
+     compress = "xz")
+save(ttm_walk_validation, 
+     file = paste0(here::here(), 
+                   "/data/ttm_walk_validation.rda"), 
+     compress = "xz")
+save(walksheds_validation, 
+     file = paste0(here::here(), 
+                   "/data/walksheds_validation.rda"), 
+     compress = "xz")
 ```
@@
 ```
+
+Bind the centroid and non-centroid walksheds (da and validations, but bound could be used in more imaginative ways for cross-validation, for example):
+```{r}
+walksheds <- bind_rows(walksheds_da,
+                       walksheds_validation) |>
+  mutate(origin_type = rep(c("centroid", "non-centroid"), each = nrow(walksheds_da)))
+```

### 2024-10-18 data-raw/02-OSM-Amenities.Rmd
Verified that the project is reproducible after updating renv

@@
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(glue) # Interpreted String Literals
 library(h3) # R Bindings for H3
+library(here) # A Simpler Way to Find Your Files
 library(osmdata) # Import 'OpenStreetMap' Data as Simple Features or Spatial Objects
@@
 ```{r}
-save(sustenance, file = "sustenance.rda", compress = "xz")
-save(education, file = "education.rda", compress = "xz")
-save(financial, file = "financial.rda", compress = "xz")
-save(healthcare, file = "healthcare.rda", compress = "xz")
-save(amenities_points, file = "amenities_points.rda", compress = "xz")
+save(sustenance, 
+     file = paste0(here::here(), 
+                   "/data/sustenance.rda"),
+     compress = "xz")
+save(education, file = paste0(here::here(), 
+                   "/data/education.rda"),
+     compress = "xz")
+save(financial, file = paste0(here::here(), 
+                   "/data/financial.rda"),
+     compress = "xz")
+save(healthcare, file = paste0(here::here(), 
+                   "/data/healthcare.rda"),
+     compress = "xz")
+save(amenities_points, file = paste0(here::here(), 
+                   "/data/amenities_points.rda"),
+     compress = "xz")
 ```

### 2024-10-18 data-raw/03-Network-Attributes-by-Walkshed.Rmd
Verified that the project is reproducible after updating renv

@@
 library(dplyr) # A Grammar of Data Manipulation
+library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(glue) # Interpreted String Literals
+library(here) # A Simpler Way to Find Your Files
 library(igraph) # Network Analysis and Visualization
@@
 ```{r}
-load(file = "hamilton_net.rda")
-load(file = "urban_hammer_da.rda")
-load(file = "walksheds_da.rda")
-load(file = "walksheds_validation.rda")
+load(file = glue::glue(here::here(), "/data/hamilton_net.rda"))
+load(file = glue::glue(here::here(), "/data/urban_hammer_da.rda"))
+load(file = glue::glue(here::here(), "/data/walksheds_da.rda"))
+load(file = glue::glue(here::here(), "/data/walksheds_validation.rda"))
 ```
@@
 ```{r}
-save(walksheds_da_net_vars, file = "walksheds_da_net_vars.rda", compress = "xz")
+save(walksheds_da_net_vars, 
+     file = glue::glue(here::here(), "/data/walksheds_da_net_vars.rda"), 
+     compress = "xz")
 ```
@@
 ```{r}
-save(walksheds_validation_net_vars, file = "walksheds_validation_net_vars.rda", compress = "xz")
+save(walksheds_validation_net_vars, 
+     file = glue::glue(here::here(), 
+                       "/data/walksheds_validation_net_vars.rda"),
+     compress = "xz")
 ```
 
+Summarize. Notice that `girth` is problematic for the skim, although it can be examined with `summary()`:
 ```{r}
 walksheds_da_net_vars |>
+  select(-girth) |>
   group_by(Type) |>
@@
 ```
-
+Summarize the walksheds for 
 ```{r}
 walksheds_validation_net_vars |>
+  select(-girth) |>
   group_by(Type) |>
@@
 ```{r}
-save(walksheds_da_motif_3_count, file = "walksheds_da_motif_3_count.rda", compress = "xz")
+save(walksheds_da_motif_3_count, 
+     file = glue::glue(here::here(), "/data/walksheds_da_motif_3_count.rda"), compress = "xz")
 ```
@@
 
+Plot chosen walkshed:
 ```{r}
 ggplot()  +
+  geom_sf(data = urban_hammer_da |> 
+            filter(GeoUID == ID), 
+          fill = "red") +
   geom_sf(data = walksheds_da |>
@@
           fill = "blue") +
-  geom_sf(data = urban_hammer_da |> 
-            filter(GeoUID == ID), 
-          fill = "red") +
   geom_sf(data = wsnet |> 
@@
 
-Plot the motifs of class 10:
+Plot the motifs of class 10 and 15:
 ```{r}
@@
 ```{r}
-ID <- "35250675" #
+ID <- "35250677" #
 
@@
 
+<!--
 Find the motifs of class 10 and 15 in wsnet:
@@
 Almost all elements of the network are part of a motif. But the number of motifs is low, which suggests that a large number of dead ends tends to reduce the number of motifs.
+-->
 
@@
 save(walksheds_da_degree_dist, 
-     file = "walksheds_da_degree_dist.rda", 
+     file = glue::glue(here::here(), "/data/walksheds_da_degree_dist.rda"), 
      compress = "xz")

### 2024-10-18 data-raw/04-Accessibility-Calculations.Rmd
Verified that the project is reproducible after updating renv

@@
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
-library(h3)
+library(glue) # Interpreted String Literals
+library(h3) # R Bindings for H3
+library(here) # A Simpler Way to Find Your Files
 library(sf) # Simple Features for R
@@
 #load(file = "hamilton_net.rda")
-load(file = "amenities_points.rda")
-load(file = "urban_hammer_da.rda")
+load(file = paste0(here::here(), 
+                   "/data/amenities_points.rda"))
+load(file = paste0(here::here(), 
+                   "/data/urban_hammer_da.rda"))
 data(urban_types)
-load(file = "walksheds_da.rda")
-load(file = "walksheds_validation.rda")
+load(file = paste0(here::here(), 
+                   "/data/walksheds_da.rda"))
+load(file = paste0(here::here(), 
+                   "/data/walksheds_validation.rda"))
 ```
@@
 ```{r}
-save(amenities_walksheds_da, file = "amenities_walkshed_da.rda", compress = "xz")
-save(amenities_walksheds_validation, file = "amenities_walkshed_validation.rda", compress = "xz")
+save(amenities_walksheds_da, 
+     file = paste0(here::here(), 
+                   "/data/amenities_walkshed_da.rda"),
+     compress = "xz")
+save(amenities_walksheds_validation, file = paste0(here::here(), 
+                   "/data/amenities_walkshed_validation.rda"),
+     compress = "xz")
 ```
@@
 ```{r}
-load("ttm_walk_da.rda")
+load(paste0(here::here(), 
+                   "/data/ttm_walk_da.rda"))
 ```
 
+Summary of amenities by h3 geometry:
 ```{r}
@@
 $$
+
 For all destinations this is the equation:
@@
 $Y$ is the percentage of trips and $t$ is the time in minutes.
-
 ```{r}

### 2024-10-18 data-raw/05-Walkshed-Profiles.Rmd
Verified that the project is reproducible after updating renv

@@
 Load packages:
+
 ```{r warning=FALSE, message=FALSE}
@@
 Load data:
+
 ```{r}
-load(file = "amenities_walkshed_da.rda")
-load(file = "hamilton_net.rda")
-load(file = "urban_hammer_da.rda")
+load(file = glue::glue(here::here(), "/data/amenities_walkshed_da.rda"))
+load(file = glue::glue(here::here(), "/data/hamilton_net.rda"))
+load(file = glue::glue(here::here(), "/data/urban_hammer_da.rda"))
 data("urban_types")
-load(file = "walksheds_da.rda")
-load(file = "walksheds_da_net_vars.rda")
+load(file = glue::glue(here::here(), "/data/walksheds_da.rda"))
+load(file = glue::glue(here::here(), "/data/walksheds_da_net_vars.rda"))
 ```
@@
 The aim is to train a classification tree to profile urban and suburban walksheds based on the attributes of the network. We need to drop 1 observation that has an NA in `transitivity`:
+
 ```{r}
 d <- walksheds_da_net_vars |>
+  select(-girth) |>
   drop_na() |>
@@
 
-We use a technique with evolutionary trees - an alternative to classification trees that is less affecte by the greediness of the algorithm.
+We use a technique with evolutionary trees - an alternative to classification trees that is less affected by the greediness of the algorithm.
 Since evolutionary trees have some randomness (for the initial values), we can try several to find whether there are patterns.
@@
 Tree 1 with seed 1 (912571):
+
 ```{r}
@@
 Tree 2 with seed 2 (94766):
+
 ```{r}
@@
 Tree 3 with seed 3 (46125):
+
 ```{r}
@@
 Tree 4 with seed 4 (556090)
+
 ```{r}
@@
 Create a function to estimate the missclassification rate and the evaluation function:
+
 ```{r}
@@
 Plot the trees:
+
 ```{r}
@@
 Check the break labels of `tree3` to round:
+
 ```{r}
@@
 
-0.00591859212548868 & normalized_motifs_3 < 0.86734693877551 ~ "Leaf 1",
-                          edge_density < 0.00591859212548868 & normalized_motifs_3 >= 0.86734693877551 ~ "Leaf 2",
-                          edge_density >= 0.00591859212548868 & normalized_motifs_3 < 0.967741935483871 ~ "Leaf 3",
+0.00591859212548868 & normalized_motifs_3 \< 0.86734693877551 \~ "Leaf 1", edge_density \< 0.00591859212548868 & normalized_motifs_3 \>= 0.86734693877551 \~ "Leaf 2", edge_density \>= 0.00591859212548868 & normalized_motifs_3 \< 0.967741935483871 \~ "Leaf 3",
 
@@
 
-<!--
-Check the break labels of `tree3` to round:
+\<!-- Check the break labels of `tree3` to round:
+
 ```{r}
@@
 Plot the tree with the rounded break labels:
+
 ```{r}
@@
 Leaf 1:
+
 ```{r}
@@
 Leaf 2 (urban):
+
 ```{r}
@@
 Leaf 3 (suburban):
+
 ```{r}
@@
 Leaf 4 (urban):
+
 ```{r}
@@
 Add leaf labels to data:
+
 ```{r}
@@
 Summary of amenity categories by DA and Type:
+
 ```{r}
@@
 To map accessibility by category of amenity we first categorize the number of amenities:
+
 ```{r}
@@
 Plot:
+
 ```{r}
@@
 Pivot wider:
+
 ```{r}

### 2024-10-18 paper/paper.Rmd
Verified that the project is reproducible after updating renv

@@
 abstract: |
-  Fifteen-minutes neihborhoods, a form of normative chronourbanism based on cumulative opportunities, has gained recognized as a way to reduce the need for motorized travel, and increase the livability, convenience, and health of the public.
+  Fifteen-minutes neihborhoods, a form of normative chronourbanism based on cumulative opportunities, has aroused interest as a way to reduce the need for motorized travel, and increase the livability, convenience, and health of the public.
 At the core of this concept is a pedshed, an area defined by the walkable isochrone of the eponymous fifteen minutes.
@@
 
+```{r chunk-setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,
+                      warning = FALSE,
+                      message = FALSE)
+```
+
 # Introduction

### 2024-10-18 paper/paper.Rmd
methodology writng

@@
 
+15-minute walking neighborhoods
+are studied, and their accessibility levels assessed (positive analysis).
+Optimal opportunity landscapes are then used to simulate equivalent opportunity
+landscapes throughout the region. Accessibility is then reanalyzed from
+the normative perspective of the provision of opportunities. The results of
+this analysis are finally correlated to neighborhood network attributes, including
+connectivity, centrality, and clustering
+
+# Data
+
+The data used to study 15-minute walking neighborhoods in the city of Hamilton was obtained by extracting relevant attributes of the street network within 15-minute pedestrian sheds.
+These sheds were calculated for each neighborhood as polygon representations of origin-destination walking time matrices, with a maximum travel time of 15 minutes.
+Walkshed calculations were generated using the travel_time_matrix function for walking from the {r5r} R package.
+This process involved defining the street network, as well as the origins and destinations of walking trips.
+The street network data for these calculations was sourced from OpenStreetMap @OpenStreetMap2023.
+
+<!--- 
+To see the details of the routing see notebook 01-OSM-Network-and-Routing.Rmd in folder data-raw
+--->
+
+The origins we established are proxies for the centroids of neighborhoods in the urban and peri-urban areas of the city of Hamilton.
+Specifically, these are the centroids of Hamilton's Dissemination Areas (DA).
+A DA is composed of one or more neighboring dissemination blocks and represents the smallest standard geographic area for which all census data are disseminated in Canada.
+Data on urban and sub-urban boundaries and DAs were retrieved from @HamiltonData2023.
+
+Destinations were established as the centroids of geohashed hexagons within the region of interest, where Hamilton's amenities are located, assuming that presence of amenities is a proxy to provision of opportunities.
+These destinations were chosen to avoid calculating travel times to each individual amenity.
+Instead, we used the geohashed polygons as destinations, identifying the amenities within each polygon and assuming that the centroid of each hexagon represents the location of the amenities within it.
+Specifically, we retrieved geohashing data for Hamilton using the H3 system (SOURCE) and generated child hexagons at resolution 13, with an average area of 43.87 $m^2$.
+Amenities location and attributes were downloaded from @OpenStreetMap2023.
+<!--- 
+To see the details of the destinations see notebook 02-OSM-Amenities.Rmd in folder data-raw
+--->
+
+# Methods
+
+The method proposed involves to study 15-minute walking neighborhoods in two stages.
+The first stage comprises assessing accessibility levels from the normative perspective of the provision of opportunities.
+In the second stage, the results of this analysis are correlated to neighborhood network attributes, including connectivity, centrality, and clustering.
+
+## Calculation of accessibility levels
+
+The indicator we use assesses levels of accessibility from a normative perspective, focusing on reachable opportunities as estimates of the potential for spatial interactions @Hansen1959.
+The advantage of using this form of accessibility analysis in this study is that it has been shown to be a valuable method for exploring the relationship between transportation infrastructure and urban structure @Handy1997.
+
+In this study, we calculate cumulative accessibility scores by summing the number of amenities within urban and suburban walksheds, categorized into four types: financial, sustenance, healthcare, and library.
+<!--- 
+To see the details of the accessibility calculations see notebook 04-Accessibility-Calculations.Rmd in folder data-raw
+--->
+
+## Walkshed network attributes by walkshed
+
+We characterize urban and suburban walksheds by means of their network attributes.
+We selected 11 of the network attributes from those proposed in R/igraph, a widely used network analysis R library @IgraphManual.
+
+- Number of vertices.
+- Number of edges.
+- Diameter.
+- Radius.
+- Global efficiency: the global efficiency of a network is defined as the average of inverse. distances between all pairs of vertices.´
+- Edge density: the density of a graph is the ratio of the actual number of edges and the largest possible number of edges in the graph.
+- Mean distance: mean length of all the shortest paths from or to the vertices in the network.
+- Girth: the girth of a graph is the length of the shortest circle in it.
+Minimum cut of a graph: the minimum total weight of the edges needed to remove to separate the graph into (at least) two components. 
+- Edge connectivity: it is a measure of network vulnerability, minimum number of edges to remove to disconnect two nodes (there is no longer a path between them.
+- Transitivity: it is the probability that the neighbors of a node are neighbors between them. In a Manhattan-style grid this number will be low because there will be lots and lots of squares.
+- Motifs (calculated for size 3 and 4): It indicates the presence of particular subgraphs that repeat themselves.
+In other words, a graph with a large number of motifs will tend to have few unique elements (some elements happen again, again, and again).
+A graph with many unique elements will have few motifs (repetitive patterns).
+<!--- 
+To see the details of the attribute calculations see notebook 03-Network-Attributes-by-Walkshed.Rmd in folder data-raw
+--->
+
+## Walkshed classification
+
+To profile urban and suburban walksheds based on the attributes of the network we use a classification tree.
+In particular We use a technique with evolutionary trees - an alternative to classification trees that is less affected by the greediness of the algorithm.
+Since evolutionary trees have some randomness (for the initial values), we try several to find whether there are patterns.
+In this case we use four different random seeds.
+
+<!--- 
+To see the details of walkshed profile calculations see notebook 06-Walkshed-Profiles.Rmd in folder data-raw
+--->
+
 # References {-}

### 2024-10-18 paper/paper.Rmd
Copying the analysis to paper

--- 2024-10-18 paper/paper.Rmd
@@
 
+```{r load-packages}
+library(broom) # Convert Statistical Objects into Tidy Tibbles
+library(dplyr) # A Grammar of Data Manipulation
+library(evtree) # Evolutionary Learning of Globally Optimal Trees
+library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(glue) # Interpreted String Literals
+library(gt) # Easily Create Presentation-Ready Display Tables
+library(rpart) # Recursive Partitioning and Regression Trees
+library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
+library(patchwork) # The Composer of Plots
+library(partykit)
+library(tidyr) # Tidy Messy Data
+library(tree) # Classification and Regression Trees
+library(sf) # Simple Features for R
+library(skimr) # Compact and Flexible Summaries of Data
+library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
+library(stringr) # Simple, Consistent Wrappers for Common String Operations
+library(units) # Measurement Units for R Vectors
+library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
+```
+
+```{r load-data}
+ # From package vaccHamilton
+data("urban_types")
+
+# Analysis-ready data. See notebooks in folder data-raw for details
+load(file = glue::glue(here::here(),
+                       "/data/amenities_walkshed_da.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/hamilton_net.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/urban_hammer_da.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/walksheds_da.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/walksheds_da_net_vars.rda"))
+```
+
+The aim is to train a classification tree to profile urban and suburban walksheds based on the attributes of the network. We need to drop 1 observation that has an NA in `transitivity`:
+
+```{r}
+d <- walksheds_da_net_vars |>
+  select(-girth) |>
+  drop_na() |>
+  mutate(normalized_motifs_3 = motifs_3/n_edges,
+         normalized_motifs_4 = motifs_4/n_edges)
+```
+
+```{r identify-network-characteristics-by-urban-type}
+#seed1 <- 908544
+set.seed(908544)
+ev.tree <- evtree::evtree(Type ~ ., 
+                data = d |> select(-c(GeoUID, 
+                                      motifs_3, 
+                                      motifs_4, 
+                                      n_edges, 
+                                      n_nodes)))
+
+ct.tree <- partykit::cttree.treeree(Type ~ ., 
+                data = d |> select(-c(GeoUID, 
+                                      motifs_3, 
+                                      motifs_4, 
+                                      n_edges, 
+                                      n_nodes)))
+
+rpart.tree <- rpart::rpart(Type ~ ., 
+                data = d |> select(-c(GeoUID, 
+                                      motifs_3, 
+                                      motifs_4, 
+                                      n_edges, 
+                                      n_nodes)))
+
+tree.tree <- tree::tree(Type ~ ., 
+                data = d |> select(-c(GeoUID, 
+                                      motifs_3, 
+                                      motifs_4, n_edges,
+                                      n_nodes)))
+
+```
+
+```{r}
+evtree.cm <- data.frame(obs = d$Type,
+                        pred = predict(ev.tree, newdata = d)) |>
+  mutate(obs = ifelse(obs == "Urban", 1, 0),
+         pred = ifelse(pred == "Urban", 1, 0))
+
+xtabs(obs ~ pred,
+      data = evtree.cm)
+```
+
+```{r}
+rpart.tree |>
+  as.party() |>
+ggparty() +
+  geom_edge() +
+  geom_edge_label(size = 2.5) +
+  geom_node_label(aes(label = splitvar ),
+                  id = "inner",
+                  size = 2, nudge_y = -0.022)
+```
+
 # Introduction
 
-"Il faut oublier la traversée de Paris d'est en ouest en voiture"^[ "We must forget about crossing Paris from east to west by car"] [@alimi2020].
+> "Il faut oublier la traversée de Paris d'est en ouest en voiture"^[ "We must forget about crossing Paris from east to west by car"]. 
+> --Anne Hidalgo
 
-With this declaration during her reelection campaign in 2020, Anne Hidalgo brought international attention to the concept of walkable fifteen-minutes neighborhoods.
+With this declaration during her reelection campaign in 2020, Anne Hidalgo brought international attention to the concept of walkable fifteen-minutes neighborhoods [@alimi2020].
 

### 2024-10-18 paper/paper.Rmd
small edits in methodology

--- 2024-10-18 paper/paper.Rmd
@@
 
-The method proposed involves to study 15-minute walking neighborhoods in two stages.
-The first stage comprises assessing accessibility levels from the normative perspective of the provision of opportunities.
-In the second stage, the results of this analysis are correlated to neighborhood network attributes, including connectivity, centrality, and clustering.
+The proposed method involves studying 15-minute walking neighborhoods in two stages.
+The first stage assesses accessibility levels from a normative perspective, focusing on the provision of opportunities.
+In the second stage, the results are correlated with neighborhood network attributes, including connectivity, centrality, and clustering.
 

### 2024-10-18 paper/paper.Rmd
Classification tree for walksheds by urban types

@@
 
+```{r load-packages}
+library(broom) # Convert Statistical Objects into Tidy Tibbles
+library(cowplot)
+library(dplyr) # A Grammar of Data Manipulation
+library(evtree) # Evolutionary Learning of Globally Optimal Trees
+library(ggparty) # 'ggplot' Visualizations for the 'partykit' Package
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(glue) # Interpreted String Literals
+library(gt) # Easily Create Presentation-Ready Display Tables
+library(rpart) # Recursive Partitioning and Regression Trees
+library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
+library(patchwork) # The Composer of Plots
+library(partykit)
+library(tidyr) # Tidy Messy Data
+library(tree) # Classification and Regression Trees
+library(sf) # Simple Features for R
+library(skimr) # Compact and Flexible Summaries of Data
+library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
+library(stringr) # Simple, Consistent Wrappers for Common String Operations
+library(units) # Measurement Units for R Vectors
+library(vaccHamilton) # A Data Package to Estimate Accessibility of Vaccination Sites in Hamilton, ON
+```
+
+```{r load-data}
+ # From package vaccHamilton
+data("urban_types")
+
+# Analysis-ready data. See notebooks in folder data-raw for details
+load(file = glue::glue(here::here(),
+                       "/data/amenities_walkshed_da.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/hamilton_net.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/urban_hammer_da.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/walksheds_da.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/walksheds_da_net_vars.rda"))
+```
+
+The aim is to train a classification tree to profile urban and suburban walksheds based on the attributes of the network. We need to drop 1 observation that has an NA in `transitivity`:
+
+```{r prepare-data-frame-for-analysis}
+# Use normalized motifs because the number of motifs will depend on the number of edges. Divide by edges to obtain a comparable number
+# Remove girth which is problematic for modeling
+d <- walksheds_da_net_vars |>
+  select(-girth) |>
+  drop_na() |>
+  mutate(`Normalized Motifs (3)` = motifs_3/n_edges,
+         `Normalized Motifs (4)` = motifs_4/n_edges) |>
+  rename(`Edge Density` = edge_density)
+```
+
+```{r train-evolutionary-tree}
+#seed1 <- 908544
+set.seed(908544)
+ev.tree <- evtree::evtree(Type ~ ., 
+                data = d |> select(-c(GeoUID, 
+                                      motifs_3, 
+                                      motifs_4, 
+                                      n_edges, 
+                                      n_nodes)))
+```
+
+```{r create-tree-plot}
+party_plot <- ev.tree |>
+  ggparty() +
+  geom_edge() +
+  #geom_edge_label(size = 2.5) +
+  geom_node_label(aes(label = splitvar ),
+                  id = "inner",
+                  size = 2, nudge_y = -0.022) +
+  geom_node_splitvar(id = "inner", size = 4) +
+  geom_edge_label(mapping = aes(label = prettyNum(breaks_label, digits = 2)), size = 2.5) +
+  # # pass list to gglist containing all ggplot components we want to plot for each
+  # # (default: terminal) node
+  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = Type),
+                                        position = position_fill()),
+                               #xlab(""),
+                               #ylab(""),
+                               theme_void(),
+                               scale_fill_manual(name = "Urban fabric type",
+                    values = c("#FDAE61", "#1A9850")),
+                              theme(axis.text = element_blank())))
+```
+
+```{r legend-for-tree-plot}
+#creating a legend
+types_legend <- ggplot(d) +
+  geom_bar(aes(x = Type, 
+               fill = Type)) +
+  scale_fill_manual(name = "Urban fabric type",
+                    values = c("#FDAE61", "#1A9850"),
+              labels = c("Suburban", 
+                         "Urban") +
+                theme(legend.position = "bottom")) 
+types_legend <- cowplot::get_legend(types_legend)
+```
+
+```{r tree-plot, fig.cap="\\label{fig:classification-tree}Classification tree of walksheds by type of urban fabric"}
+cowplot::plot_grid(party_plot, legd_renamed, 
+                   nrow = 2,
+                   align = "v", axis = "l", rel_heights = c(6, 1), rel_widths = c(1, 1))
+```
+
 # Introduction
 
-"Il faut oublier la traversée de Paris d'est en ouest en voiture"^[ "We must forget about crossing Paris from east to west by car"] [@alimi2020].
+> "Il faut oublier la traversée de Paris d'est en ouest en voiture"^[ "We must forget about crossing Paris from east to west by car"]. 
+> --Anne Hidalgo
 
-With this declaration during her reelection campaign in 2020, Anne Hidalgo brought international attention to the concept of walkable fifteen-minutes neighborhoods.
+With this declaration during her reelection campaign in 2020, Anne Hidalgo brought international attention to the concept of walkable fifteen-minutes neighborhoods [@alimi2020].
 
@@
 
-The proposed method involves studying 15-minute walking neighborhoods in two stages.
-The first stage assesses accessibility levels from a normative perspective, focusing on the provision of opportunities.
-In the second stage, the results are correlated with neighborhood network attributes, including connectivity, centrality, and clustering.
+The method proposed involves to study 15-minute walking neighborhoods in two stages.
+The first stage comprises assessing accessibility levels from the normative perspective of the provision of opportunities.
+In the second stage, the results of this analysis are correlated to neighborhood network attributes, including connectivity, centrality, and clustering.
 

### 2024-10-18 paper/paper.tex
Verified that the project is reproducible after updating renv
Classification tree for walksheds by urban types

@@
 
+\usepackage{graphicx}
 \ifxetex
@@
   Fifteen-minutes neihborhoods, a form of normative chronourbanism based
-  on cumulative opportunities, has gained recognized as a way to reduce
+  on cumulative opportunities, has aroused interest as a way to reduce
   the need for motorized travel, and increase the livability,
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+The aim is to train a classification tree to profile urban and suburban
+walksheds based on the attributes of the network. We need to drop 1
+observation that has an NA in \texttt{transitivity}:
 
+\begin{verbatim}
+## pred
+##   0   1 
+## 114 228
+\end{verbatim}
+
+\includegraphics{paper_files/figure-latex/unnamed-chunk-3-1.pdf}
+
+\section{Introduction}\label{introduction}
+
+\begin{quote}
 ``Il faut oublier la traversée de Paris d'est en ouest en
 voiture''\footnote{ ``We must forget about crossing Paris from east to
-  west by car''} \citep{alimi2020}.
+  west by car''}. --Anne Hidalgo
+\end{quote}
+
+With this declaration during her reelection campaign in 2020, Anne
+Hidalgo brought international attention to the concept of walkable
+fifteen-minutes neighborhoods \citep{alimi2020}.
 
@@
 \citet{pozoukidou2021fifteen} \citet{weng2019fifteen}
+
+15-minute walking neighborhoods are studied, and their accessibility
+levels assessed (positive analysis). Optimal opportunity landscapes are
+then used to simulate equivalent opportunity landscapes throughout the
+region. Accessibility is then reanalyzed from the normative perspective
+of the provision of opportunities. The results of this analysis are
+finally correlated to neighborhood network attributes, including
+connectivity, centrality, and clustering
+
+\section{Data}\label{data}
+
+The data used to study 15-minute walking neighborhoods in the city of
+Hamilton was obtained by extracting relevant attributes of the street
+network within 15-minute pedestrian sheds. These sheds were calculated
+for each neighborhood as polygon representations of origin-destination
+walking time matrices, with a maximum travel time of 15 minutes.
+Walkshed calculations were generated using the travel\_time\_matrix
+function for walking from the \{r5r\} R package. This process involved
+defining the street network, as well as the origins and destinations of
+walking trips. The street network data for these calculations was
+sourced from OpenStreetMap \citet{OpenStreetMap2023}.
+
+The origins we established are proxies for the centroids of
+neighborhoods in the urban and peri-urban areas of the city of Hamilton.
+Specifically, these are the centroids of Hamilton's Dissemination Areas
+(DA). A DA is composed of one or more neighboring dissemination blocks
+and represents the smallest standard geographic area for which all
+census data are disseminated in Canada. Data on urban and sub-urban
+boundaries and DAs were retrieved from \citet{HamiltonData2023}.
+
+Destinations were established as the centroids of geohashed hexagons
+within the region of interest, where Hamilton's amenities are located,
+assuming that presence of amenities is a proxy to provision of
+opportunities. These destinations were chosen to avoid calculating
+travel times to each individual amenity. Instead, we used the geohashed
+polygons as destinations, identifying the amenities within each polygon
+and assuming that the centroid of each hexagon represents the location
+of the amenities within it. Specifically, we retrieved geohashing data
+for Hamilton using the H3 system (SOURCE) and generated child hexagons
+at resolution 13, with an average area of 43.87 \(m^2\). Amenities
+location and attributes were downloaded from \citet{OpenStreetMap2023}.
+
+\section{Methods}\label{methods}
+
+The method proposed involves to study 15-minute walking neighborhoods in
+two stages. The first stage comprises assessing accessibility levels
+from the normative perspective of the provision of opportunities. In the
+second stage, the results of this analysis are correlated to
+neighborhood network attributes, including connectivity, centrality, and
+clustering.
+
+\subsection{Calculation of accessibility
+levels}\label{calculation-of-accessibility-levels}
+
+The indicator we use assesses levels of accessibility from a normative
+perspective, focusing on reachable opportunities as estimates of the
+potential for spatial interactions \citet{Hansen1959}. The advantage of
+using this form of accessibility analysis in this study is that it has
+been shown to be a valuable method for exploring the relationship
+between transportation infrastructure and urban structure
+\citet{Handy1997}.
+
+In this study, we calculate cumulative accessibility scores by summing
+the number of amenities within urban and suburban walksheds, categorized
+into four types: financial, sustenance, healthcare, and library.
+
+\subsection{Walkshed network attributes by
+walkshed}\label{walkshed-network-attributes-by-walkshed}
+
+We characterize urban and suburban walksheds by means of their network
+attributes. We selected 11 of the network attributes from those proposed
+in R/igraph, a widely used network analysis R library
+\citet{IgraphManual}.
+
+\begin{itemize}
+\tightlist
+\item
+  Number of vertices.
+\item
+  Number of edges.
+\item
+  Diameter.
+\item
+  Radius.
+\item
+  Global efficiency: the global efficiency of a network is defined as
+  the average of inverse. distances between all pairs of vertices.´
+\item
+  Edge density: the density of a graph is the ratio of the actual number
+  of edges and the largest possible number of edges in the graph.
+\item
+  Mean distance: mean length of all the shortest paths from or to the
+  vertices in the network.
+\item
+  Girth: the girth of a graph is the length of the shortest circle in
+  it. Minimum cut of a graph: the minimum total weight of the edges
+  needed to remove to separate the graph into (at least) two components.
+\item
+  Edge connectivity: it is a measure of network vulnerability, minimum
+  number of edges to remove to disconnect two nodes (there is no longer
+  a path between them.
+\item
+  Transitivity: it is the probability that the neighbors of a node are
+  neighbors between them. In a Manhattan-style grid this number will be
+  low because there will be lots and lots of squares.
+\item
+  Motifs (calculated for size 3 and 4): It indicates the presence of
+  particular subgraphs that repeat themselves. In other words, a graph
+  with a large number of motifs will tend to have few unique elements
+  (some elements happen again, again, and again). A graph with many
+  unique elements will have few motifs (repetitive patterns).
+\end{itemize}
+
+\subsection{Walkshed classification}\label{walkshed-classification}
+
+To profile urban and suburban walksheds based on the attributes of the
+network we use a classification tree. In particular We use a technique
+with evolutionary trees - an alternative to classification trees that is
+less affected by the greediness of the algorithm. Since evolutionary
+trees have some randomness (for the initial values), we try several to
+find whether there are patterns. In this case we use four different
+random seeds.
 

### 2024-10-20 paper/paper.Rmd
methodology writing

--- 2024-10-18 paper/paper.Rmd
@@
 
-15-minute walking neighborhoods
-are studied, and their accessibility levels assessed (positive analysis).
-Optimal opportunity landscapes are then used to simulate equivalent opportunity
-landscapes throughout the region. Accessibility is then reanalyzed from
-the normative perspective of the provision of opportunities. The results of
-this analysis are finally correlated to neighborhood network attributes, including
-connectivity, centrality, and clustering
+We address the following research questions:
+
+1. Is there a difference in the transportation network topology between urban and suburban areas within 15-minute walking distances?
+2. If such a difference exists, what is its effect on the 15-minute walksheds in terms of accessibility, where accessibility is understood as the provision of opportunities?
 
@@
 
-The data used to study 15-minute walking neighborhoods in the city of Hamilton was obtained by extracting relevant attributes of the street network within 15-minute pedestrian sheds.
-These sheds were calculated for each neighborhood as polygon representations of origin-destination walking time matrices, with a maximum travel time of 15 minutes.
+Relevant data to develop this paper comprises to use 15-min pedestrian sheds (or walksehds) classified into urban and suburban types in Hamilton.
+
+15-minute walksheds in the city of Hamilton were obtained by extracting relevant structural attributes of the street network within 15-minute pedestrian sheds.
+These sheds were calculated for each shed as polygon representations of origin-destination walking time matrices, with a maximum travel time of 15 minutes.
 Walkshed calculations were generated using the travel_time_matrix function for walking from the {r5r} R package.
@@
 The street network data for these calculations was sourced from OpenStreetMap @OpenStreetMap2023.
-
 <!--- 
@@
 Specifically, we retrieved geohashing data for Hamilton using the H3 system (SOURCE) and generated child hexagons at resolution 13, with an average area of 43.87 $m^2$.
-Amenities location and attributes were downloaded from @OpenStreetMap2023.
+Amenities location and typologies were downloaded from @OpenStreetMap2023.
 <!--- 
@@
 The proposed method involves studying 15-minute walking neighborhoods in two stages.
-The first stage assesses accessibility levels from a normative perspective, focusing on the provision of opportunities.
-In the second stage, the results are correlated with neighborhood network attributes, including connectivity, centrality, and clustering.
+In the first stage, we assess the accessibility levels of Hamilton's watersheds from a normative perspective, focusing on the provision of opportunities.
+In the second stage, we characterize the topology urban and suburban walksheds based on a set of network structure attributes.
+Finally, we analyze the effect of walkshed size on urban and suburban networks, and consequently, on access to opportunities.
 
@@
 
-In this study, we calculate cumulative accessibility scores by summing the number of amenities within urban and suburban walksheds, categorized into four types: financial, sustenance, healthcare, and library.
+In this study, we calculate cumulative accessibility scores by summing the number of amenities within urban and suburban walksheds, categorized into the four types provided by the data source: financial, sustenance, healthcare, and library.
 <!--- 
@@
 
-## Walkshed network attributes by walkshed
+## Urban and suburban walkshed modelling
 
-We characterize urban and suburban walksheds by means of their network attributes.
-We selected 11 of the network attributes from those proposed in R/igraph, a widely used network analysis R library @IgraphManual.
+### Network attributes by walkshed
+
+In this work we characterize urban and suburban walksheds based on their network structure.
+As a way to capture and describe the structure of networks we calculate a selection of 12 network structural attributes using the network analysis R package igraph (@IgraphManual).
+The set of attributes is described as follows @IgraphManual:
 
@@
 - Number of edges.
-- Diameter.
-- Radius.
+- Diameter: it is the length of the shortest path between the most distanced nodes in a graph.
+- Radius: it is the minimum among all the maximum distances between a vertex to all other vertices.
 - Global efficiency: the global efficiency of a network is defined as the average of inverse. distances between all pairs of vertices.´
@@
 - Mean distance: mean length of all the shortest paths from or to the vertices in the network.
-- Girth: the girth of a graph is the length of the shortest circle in it.
-Minimum cut of a graph: the minimum total weight of the edges needed to remove to separate the graph into (at least) two components. 
+- Girth: the girth of a graph is the length of the shortest cycle in it, being a cycle a non-empty trail in which only the first and last vertices are equal.
+- Minimum cut of a graph: the minimum total weight of the edges needed to remove to separate the graph into (at least) two components. 
 - Edge connectivity: it is a measure of network vulnerability, minimum number of edges to remove to disconnect two nodes (there is no longer a path between them.
-- Transitivity: it is the probability that the neighbors of a node are neighbors between them. In a Manhattan-style grid this number will be low because there will be lots and lots of squares.
-- Motifs (calculated for size 3 and 4): It indicates the presence of particular subgraphs that repeat themselves.
+- Transitivity: it is the probability that the neighbors of a node are neighbors between them. 
+- Motifs (calculated for 3-node and 4-node sizes): it indicates the presence of particular sub-graphs that repeat themselves.
 In other words, a graph with a large number of motifs will tend to have few unique elements (some elements happen again, again, and again).
@@
 
-## Walkshed classification
+### Modelling
 
-To profile urban and suburban walksheds based on the attributes of the network we use a classification tree.
-In particular We use a technique with evolutionary trees - an alternative to classification trees that is less affected by the greediness of the algorithm.
-Since evolutionary trees have some randomness (for the initial values), we try several to find whether there are patterns.
-In this case we use four different random seeds.
+To model urban and suburban walksheds, we employ a classification tree approach using the set topology attributes calculated in the previous stage.
+Classification trees are machine learning techniques used for both regression and classification, widely appreciated in various disciplines for their ability to represent knowledge through decision rules, making them easy to interpret and analyze @Loh2011 @James2013 @Paez2019Trees.
+Specifically, we utilize evolutionary trees, a variant of decision trees that are less constrained by the limitations of greedy top-down induction algorithms @Cha2009 @Kretowski2007 @Grubinger2014.
+Since evolutionary trees use a genetic algorithm that introduces randomness in the initial conditions, we run the model from the R package evtree @Grubinger2014 with 20 random seeds to identify the network variables that consistently characterize urban and suburban walksheds.
+
+Finally we
 
@@
 
+## Walkshed size in urban and suburban areas
+
 # References {-}

### 2024-10-21 paper/paper.Rmd
Training of evolutionary trees complete; saved trained trees to ensure that results are consistent for paper.

--- 2024-10-18 paper/paper.Rmd
@@
 library(broom) # Convert Statistical Objects into Tidy Tibbles
-library(cowplot)
+library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'
 library(dplyr) # A Grammar of Data Manipulation
@@
 library(gt) # Easily Create Presentation-Ready Display Tables
-library(rpart) # Recursive Partitioning and Regression Trees
-library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
-library(patchwork) # The Composer of Plots
-library(partykit)
+#library(rpart) # Recursive Partitioning and Regression Trees
+#library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
+#library(patchwork) # The Composer of Plots
+#library(partykit) # A Toolkit for Recursive Partytioning
 library(tidyr) # Tidy Messy Data
-library(tree) # Classification and Regression Trees
+#library(tree) # Classification and Regression Trees
 library(sf) # Simple Features for R
 library(skimr) # Compact and Flexible Summaries of Data
-library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
+#library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
 library(stringr) # Simple, Consistent Wrappers for Common String Operations
@@
 data("urban_types")
+
+# Load trained evolutionary trees
+load(file = glue::glue(here::here(),
+                       "/data/trees.rda"))
 
@@
                        "/data/walksheds_da_net_vars.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/walksheds_validation.rda"))
+load(file = glue::glue(here::here(),
+                       "/data/walksheds_validation_net_vars.rda"))
 ```
 
-The aim is to train a classification tree to profile urban and suburban walksheds based on the attributes of the network. We need to drop 1 observation that has an NA in `transitivity`:
+```{r}
+# Bind the centroid and non-centroid walksheds:
+walksheds <- bind_rows(walksheds_da,
+                       walksheds_validation) |>
+  mutate(origin_type = rep(c("centroid", "non-centroid"), each = nrow(walksheds_da)))
+```
 
-```{r prepare-data-frame-for-analysis}
-# Use normalized motifs because the number of motifs will depend on the number of edges. Divide by edges to obtain a comparable number
-# Remove girth which is problematic for modeling
-d <- walksheds_da_net_vars |>
+```{r }
+# Obtain training and validation datasets by sampling from centroid and non-centroid walksheds:
+
+# Set seed for replicability
+seed <- 3626425
+set.seed(seed)
+
+# slice_sample() will select one of two rows for each DA at random
+walksheds_train <- walksheds |>
+  group_by(GeoUID) |>
+  slice_sample(prop = 0.5)
+
+# The antijoin will keep the rows in walksheds that do not appear in walksheds_train
+walksheds_test <- walksheds |>
+  anti_join(walksheds_train |>
+              st_drop_geometry() |>
+              select(GeoUID, origin_type),
+            by = c("GeoUID", "origin_type"))
+```
+
+```{r bind-centroid-non-centroid-network-variables}
+# Bind the centroid and non-centroid dataframes with walkshed network variables:
+walksheds_net_vars <- bind_rows(walksheds_da_net_vars,
+                                walksheds_validation_net_vars) |>
+  mutate(origin_type = rep(c("centroid",
+                             "non-centroid"), 
+                           each = nrow(walksheds_da)))
+```
+
+```{r obtain-training-testing-network-variables}
+# Now filter from the network variables to obtain training and test datasets:
+walksheds_train_net_vars <- walksheds_net_vars |>
+  semi_join(walksheds_train |>
+              st_drop_geometry() |>
+              select(GeoUID, origin_type),
+            by = c("GeoUID", "origin_type"))
+
+walksheds_test_net_vars <- walksheds_net_vars |>
+  semi_join(walksheds_test |>
+              st_drop_geometry() |>
+              select(GeoUID, origin_type),
+            by = c("GeoUID", "origin_type"))
+```
+
+```{r training-and-testing-dataframes-for-evolutionary-tree}
+d <- walksheds_train_net_vars |>
   select(-girth) |>
   drop_na() |>
-  mutate(`Normalized Motifs (3)` = motifs_3/n_edges,
-         `Normalized Motifs (4)` = motifs_4/n_edges) |>
-  rename(`Edge Density` = edge_density)
+  mutate(normalized_motifs_3 = motifs_3/n_edges,
+         normalized_motifs_4 = motifs_4/n_edges)
+
+d_test <- walksheds_test_net_vars |>
+  select(-girth) |>
+  drop_na() |>
+  mutate(normalized_motifs_3 = motifs_3/n_edges,
+         normalized_motifs_4 = motifs_4/n_edges)
 ```
 
-```{r train-evolutionary-tree}
-#seed1 <- 908544
-set.seed(908544)
-ev.tree <- evtree::evtree(Type ~ ., 
-                data = d |> select(-c(GeoUID, 
-                                      motifs_3, 
-                                      motifs_4, 
-                                      n_edges, 
-                                      n_nodes)))
+```{r best-tree}
+# Retrieve the best evolutionary tree.
+The algorithm of evolutionary trees has some randomness due to random initial values.
+We compared a number of trees and chose the one that gave the best fit in terms of missclassification and size of the model.
+See notebook `05-Walkshed-Profiles.Rmd in folder `data-raw` for details.
+
+best_evtree <- trees[[4]]
 ```
 
-```{r create-tree-plot}
-party_plot <- ev.tree |>
-  ggparty() +
+```{r generate-plot-best-tree}
+# Round the labels of the edges. The relevant information is in a dataframe produced by ggparty()
+ggparty_plot <- best_evtree |>
+  ggparty()
+
+# Round the number of digits and keep the inequality signs
+ggparty_data <- ggparty_plot$data |>
+  mutate(breaks_label = paste0(substr(breaks_label, 4, 5), 
+                               substr(breaks_label,10, 15)),
+         splitvar = case_when(splitvar == "edge_density" ~ "Edge Density",
+                              splitvar == "transitivity" ~ "Transitivity",
+                              splitvar == "normalized_motifs_3" ~ "Normalized Motifs (3)",
+                              is.na(splitvar) ~ NA))
+
+# Replace the dataframe in the ggplot object
+ggparty_plot$data <- ggparty_data
+
+evtree_plot <- ggparty_plot +
   geom_edge() +
-  #geom_edge_label(size = 2.5) +
-  geom_node_label(aes(label = splitvar ),
-                  id = "inner",
-                  size = 2, nudge_y = -0.022) +
-  geom_node_splitvar(id = "inner", size = 4) +
-  geom_edge_label(mapping = aes(label = prettyNum(breaks_label, digits = 2)), size = 2.5) +
+  geom_node_splitvar(id = "inner", size = 3) +
+  # parse should be set to FALSE so the labels do not try to parse the inequality signs
+  geom_edge_label(mapping = aes(label = breaks_label),
+                  parse = FALSE,
+                  size = 2.5) +
   # # pass list to gglist containing all ggplot components we want to plot for each
@@
                                         position = position_fill()),
-                               #xlab(""),
-                               #ylab(""),
-                               theme_void(),
+                               xlab(""),
+                               ylab(""),
                                scale_fill_manual(name = "Urban fabric type",
                                                  values = c("#FDAE61", "#1A9850")),
-                              theme(axis.text = element_blank())))
+                               theme(axis.text.y = element_blank(),
+                                     panel.background = element_rect(fill = NA),
+                                     axis.ticks = element_blank(),
+                                     plot.background =  element_rect(fill = NA)))) +
+  geom_node_label(line_list = list(aes(label = paste0("Leaf ", ifelse(id >= 6, id - 3, id - 2)))),
+                  line_gpar = list(list(size = 7, parse = F)),
+                  ids = "terminal",
+                  nudge_x = 0.02,
+                  nudge_y = -0.4)
 ```
 
-```{r legend-for-tree-plot}
-#creating a legend
-types_legend <- ggplot(d) +
-  geom_bar(aes(x = Type, 
-               fill = Type)) +
-  scale_fill_manual(name = "Urban fabric type",
-                    values = c("#FDAE61", "#1A9850"),
-              labels = c("Suburban", 
-                         "Urban") +
-                theme(legend.position = "bottom")) 
-types_legend <- cowplot::get_legend(types_legend)
-```
+```{r plot-evtree, fig.cap="\\label{fig:evolutionary-tree}Results of training evolutionary tree on the attributes of the network.
+The dependent variable is Urban Fabric, a binary variable with values "Urban" (green) and "Suburban" (orange)"}
+evtree_plot
 
-```{r tree-plot, fig.cap="\\label{fig:classification-tree}Classification tree of walksheds by type of urban fabric"}
-cowplot::plot_grid(party_plot, legd_renamed, 
-                   nrow = 2,
-                   align = "v", axis = "l", rel_heights = c(6, 1), rel_widths = c(1, 1))
+ggsave(evtree_plot,
+       file = glue::glue(here::here(),"/plots/evtree.png"),
+       width = 7,
+       units = "in")
 ```

### 2024-10-21 paper/paper.Rmd
methodology edits

--- 2024-10-20 paper/paper.Rmd
@@
 The proposed method involves studying 15-minute walking neighborhoods in two stages.
-In the first stage, we assess the accessibility levels of Hamilton's watersheds from a normative perspective, focusing on the provision of opportunities.
-In the second stage, we characterize the topology urban and suburban walksheds based on a set of network structure attributes.
-Finally, we analyze the effect of walkshed size on urban and suburban networks, and consequently, on access to opportunities.
+In the first stage, we assess the accessibility of Hamilton's walksheds from a normative perspective, focusing on the availability of opportunities.
+In the second stage, we analyze the topology of urban and suburban walksheds based on various network structure attributes.
+Finally, we examine whether there are differences in transportation network topology between urban and suburban areas, and analyze how these differences affect access to opportunities within walksheds.
 
@@
 
-Finally we
+Finally we study effects in accessibility
 

### 2024-10-21 paper/paper.Rmd
paper mehods edits

@@
 A DA is composed of one or more neighboring dissemination blocks and represents the smallest standard geographic area for which all census data are disseminated in Canada.
-Data on urban and sub-urban boundaries and DAs were retrieved from @HamiltonData2023.
+Data on urban and sub-urban boundaries, DAs,and population data in 2016 were retrieved from @HamiltonData2023.
 
@@
 Instead, we used the geohashed polygons as destinations, identifying the amenities within each polygon and assuming that the centroid of each hexagon represents the location of the amenities within it.
-Specifically, we retrieved geohashing data for Hamilton using the H3 system (SOURCE) and generated child hexagons at resolution 13, with an average area of 43.87 $m^2$.
-Amenities location and typologies were downloaded from @OpenStreetMap2023.
+Specifically, we retrieved geohashing data for Hamilton using the H3 system @h3Package and generated child hexagons at resolution 13, with an average area of 43.87 $m^2$.
+Amenities location and their typology were downloaded from @OpenStreetMap2023.
 <!--- 
@@
 In this study, we calculate cumulative accessibility scores by summing the number of amenities within urban and suburban walksheds, categorized into the four types provided by the data source: financial, sustenance, healthcare, and library.
+Additionally, we extract population data of each walkshed.
+Summary statistics of population are also included in the analysis.
+
+Additionally, the accessibility assessment of is completed by obtaining summary statistics of travel times in urban and suburban walksheds.
+To study the impact of urban and suburban walksheds on walking decisions, minimum and mean travel times to destinations are introduced into time-decay functions for walking decisions, as developed by Milward (2013).
+These are active travel impedance functions based on data from Canada.
+
 <!--- 
-To see the details of the accessibility calculations see notebook 04-Accessibility-Calculations.Rmd in folder data-raw
+To see the details of the accessibility calculations, summary statistics, and obtention of impedance functions see notebook 04-Accessibility-Calculations.Rmd in folder data-raw
 --->

### 2024-10-21 data-raw/05-Walkshed-Profiles.Rmd

@@
 In this document we profile the ped sheds based on their network attributes.
+
+WARNING: Rerunning this notebook can cause the training of the evolutionary trees to deviate slightly from the reported results.
 
@@
 Load data:
-
 ```{r}
@@
 load(file = glue::glue(here::here(), "/data/walksheds_da_net_vars.rda"))
+load(file = glue::glue(here::here(), "/data/walksheds_validation.rda"))
+load(file = glue::glue(here::here(), "/data/walksheds_validation_net_vars.rda"))
 ```
+
+Bind the centroid and non-centroid walksheds (da and validations, but bound could be used in more imaginative ways for cross-validation, for example):
+```{r}
+walksheds <- bind_rows(walksheds_da,
+                       walksheds_validation) |>
+  mutate(origin_type = rep(c("centroid", "non-centroid"), each = nrow(walksheds_da)))
+```
+
+Obtain training and validation datasets by sampling from centroid and non-centroid walksheds:
+```{r}
+seed <- 3626425
+
+# slice_sample() will select one of two rows for each DA at random
+walksheds_train <- walksheds |>
+  group_by(GeoUID) |>
+  slice_sample(prop = 0.5)
+
+# The antijoin will keep the rows in walksheds that do not appear in walksheds_train
+walksheds_test <- walksheds |>
+  anti_join(walksheds_train |>
+              st_drop_geometry() |>
+              select(GeoUID, origin_type),
+            by = c("GeoUID", "origin_type"))
+```
+
+Bind the centroid and non-centroid dataframes with walkshed network variables (da and validations):
+```{r}
+walksheds_net_vars <- bind_rows(walksheds_da_net_vars,
+                                walksheds_validation_net_vars) |>
+  mutate(origin_type = rep(c("centroid", "non-centroid"), each = nrow(walksheds_da)))
+```
+
+Now filter from the network variables to obtain training and test datasets:
+```{r}
+walksheds_train_net_vars <- walksheds_net_vars |>
+  semi_join(walksheds_train |>
+              st_drop_geometry() |>
+              select(GeoUID, origin_type),
+            by = c("GeoUID", "origin_type"))
+
+walksheds_test_net_vars <- walksheds_net_vars |>
+  semi_join(walksheds_test |>
+              st_drop_geometry() |>
+              select(GeoUID, origin_type),
+            by = c("GeoUID", "origin_type"))
+```
+
+Check these dataframes for NAs, inf, or other anomalies:
+```{r}
+summary(walksheds_train_net_vars)
+summary(walksheds_test_net_vars)
+```
+
+The girth is problematic. There is a single NA in `transitivity` in the test dataset.
 
@@
 ```{r}
-d <- walksheds_da_net_vars |>
+d <- walksheds_train_net_vars |>
+  select(-girth) |>
+  drop_na() |>
+  mutate(normalized_motifs_3 = motifs_3/n_edges,
+         normalized_motifs_4 = motifs_4/n_edges)
+
+d_test <- walksheds_test_net_vars |>
   select(-girth) |>
@@
 Since evolutionary trees have some randomness (for the initial values), we can try several to find whether there are patterns.
-In this case we try four different random seed.
 
-Tree 1 with seed 1 (912571):
-
+Bunch of trees with random seeds. First, make sure that the seeds are reproducible:
 ```{r}
-seed1 <- sample.int(1000000, 1)
-#seed1 <- 908544
-set.seed(908544)
-tree1 <- evtree(Type ~ ., 
-                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
-plot(tree1)
+n_trees <- 30 # select number of trees
+set.seed(98732475)
+seeds <- sample.int(1000000, n_trees)
 ```
 
-Tree 2 with seed 2 (94766):
-
+Traaaain!
 ```{r}
-seed2 <- sample.int(1000000, 1)
-#seed2 <- 669402
-seed2
-set.seed(seed2)
-tree2 <- evtree(Type ~ ., 
-                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
-plot(tree2)
+counter <- 0 
+trees <- list()
+for(i in seeds){
+  counter <- counter + 1
+  #set.seed(seeds[counter])
+  #tree_name <- glue::glue("tree{counter}")
+  trees[[counter]] <- evtree(Type ~ ., 
+                             data = d |> 
+                               select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes, origin_type)),
+                             control = evtree.control(seed = seeds[counter]))
+}
 ```
 
-Tree 3 with seed 3 (46125):
+Name the trees in the list above:
+```{r}
+named_trees <- data.frame(t = "evtree ", count = c(1:n_trees)) |>
+  mutate(t = glue::glue("{t}{count}"))
 
-```{r}
-seed3 <- sample.int(1000000, 1)
-#seed3 <- 853251
-seed3
-set.seed(seed3)
-tree3 <- evtree(Type ~ ., 
-                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
-plot(tree3)
+names(trees) <- named_trees$t
 ```
 
-Tree 4 with seed 4 (556090)
-
+Check the attributes of the trees:
 ```{r}
-seed4 <- sample.int(1000000, 1)
-#seed4 <- 707633
-seed4
-set.seed(seed4)
-tree4 <- evtree(Type ~ ., 
-                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
-plot(tree4)
+trees
 ```
 
-Create a function to estimate the missclassification rate and the evaluation function:
+The frequency of variables in this set of tree is as follows:
 
+Variable            | Number of trees where it appears
+--------------------| --------------------------------
+edge_density        | IIIII IIIII IIIII IIIII IIIII IIIII 
+normalized_motifs_3 | IIIII IIIII IIIII IIIII IIIII IIIII 
+diameter            | IIIII II
+transitivity        | II
+mean_distance       | II
+normalized_motifs_4 | I
+
+IIIII IIIII IIIII IIIII IIIII IIIII
+
+The pattern is dominated by edge_density and normalized_motifs_3, and this has been the case after repeating this procedure several times. By now I must have trained over 150 trees.
+
+There is something about reproducing the trained trees; after restarting the computer, using the same seed does not replicate the trees; after doing this process several times, the results are very similar, but to avoid problems with slightly different trees being produced every time, save the results of this run):
+```{r eval=FALSE}
+# WARNING: Run only if you wish to replace the file with the results used in the report
+# Save to `data` folder
+save(trees,
+     file = glue::glue(here::here(), "/data/trees.rda"),
+     compress = "xz")
+```
+
+Examine missclassification in training dataset.
+The missclassification is the proportion of times that the predicted class is different from the observed class.
+The evaluation function uses the missclassification rate but penalizes by complexity (number of nodes in the tree):
 ```{r}
@@
 #trees <- list("evtree" = tree1, "rpart" = rp, "ctree" = ct, "rpart2" = rp2, "ctree2" = ct2)
-trees <- list("evtree 1" = tree1, "evtree 2" = tree2, "evtree 3" = tree3, "evtree 4" = tree4)
+#trees <- list("evtree 1" = tree1, "evtree 2" = tree2, "evtree 3" = tree3, "evtree 4" = tree4)
 
-round(sapply(trees, function(obj) c("misclassification" = mc(obj), "evaluation function" = evalfun(obj))), digits = 7)
+mc_train <- round(sapply(trees, function(obj) c("misclassification" = mc(obj), "evaluation function" = evalfun(obj))), digits = 7)
+mc_train
+```
+
+Examine missclassificationon the test data:
+```{r}
+mc <- function(obj) 1 - mean(predict(obj, newdata = d_test) == d_test$Type)
+
+evalfun <- function(obj) 2 * nrow(d_test) * mc(obj) + width(obj) * log(nrow(d_test))
+
+#trees <- list("evtree" = tree1, "rpart" = rp, "ctree" = ct, "rpart2" = rp2, "ctree2" = ct2)
+#trees <- list("evtree 1" = tree1, "evtree 2" = tree2, "evtree 3" = tree3, "evtree 4" = tree4)
+
+mc_test <- round(sapply(trees, function(obj) c("misclassification" = mc(obj), "evaluation function" = evalfun(obj))), digits = 7)
+mc_test
+```
+
+The tree that best balances the evaluation function in the train and test datasets is `tree26`:
+```{r}
+(mc_train[2,] + mc_test[2,])/2
+```
+
+There are several ties, but are actually variations of the same model. Choose Tree 4:
+```{r}
+best_evtree <- trees[[4]]
+```
+
+Plot the tree:
+```{r}
+# Round the labels of the edges. The relevant information is in a dataframe produced by ggparty()
+ggparty_plot <- best_evtree |>
+  ggparty()
+
+# Round the number of digits and keep the inequality signs
+ggparty_data <- ggparty_plot$data |>
+  mutate(breaks_label = paste0(substr(breaks_label, 4, 5), 
+                               substr(breaks_label,10, 15)),
+         splitvar = case_when(splitvar == "edge_density" ~ "Edge Density",
+                              splitvar == "transitivity" ~ "Transitivity",
+                              splitvar == "normalized_motifs_3" ~ "Normalized Motifs (3)",
+                              is.na(splitvar) ~ NA))
+
+# Replace the dataframe in the ggplot object
+ggparty_plot$data <- ggparty_data
+
+ggparty_plot +
+  geom_edge() +
+  geom_node_splitvar(id = "inner", size = 3) +
+  # parse should be set to FALSE so the labels do not try to parse the inequality signs
+  geom_edge_label(mapping = aes(label = breaks_label),
+                  parse = FALSE,
+                  size = 2.5) +
+  # # pass list to gglist containing all ggplot components we want to plot for each
+  # # (default: terminal) node
+  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = Type),
+                                        position = position_fill()),
+                               xlab(""),
+                               ylab(""),
+                               scale_fill_manual(name = "Urban fabric type",
+                                                 values = c("#FDAE61", "#1A9850")),
+                               theme(axis.text.y = element_blank(),
+                                     panel.background = element_rect(fill = NA),
+                                     axis.ticks = element_blank()))) +
+  geom_node_label(line_list = list(aes(label = paste0("Leaf ", ifelse(id >= 6, id - 3, id - 2)))),
+                  line_gpar = list(list(size = 7, parse = F)),
+                  ids = "terminal",
+                  nudge_x = 0.02,
+                  nudge_y = -0.4)
+```
+
+The leaves are as follows:
+
+Leaf        | Character
+------------|----------
+Leaf 1      | Suburban
+Leaf 2      | Urban
+Leaf 3      | Suburban
+Leaf 4      | Urban
+
+Check the break labels of `tree4` to round:
+
+```{r}
+ggparty(best_evtree)$data$breaks_label
+```
+
+Use the classification tree to add classes to train data:
+```{r add-classification-to-train-data}
+
+d <- d |>
+  mutate(leaf = case_when(edge_density < 0.00590599876314162 & normalized_motifs_3 < 0.865853658536585  ~ "Leaf 1",
+                          edge_density < 0.00590599876314162 & normalized_motifs_3 >= 0.865853658536585 ~ "Leaf 2",
+                          edge_density >= 0.00590599876314162 & normalized_motifs_3 < 0.967078189300412 ~ "Leaf 3",
+                          edge_density >= 0.00590599876314162 & normalized_motifs_3 >= 0.967078189300412 ~ "Leaf 4") ,
+         leaf = factor(leaf))
+```
+
+Use the classification tree to add classes to test data:
+```{r add-classification-to-test-data}
+d_test <- d_test |>
+  mutate(leaf = case_when(edge_density < 0.00590599876314162 & normalized_motifs_3 < 0.865853658536585  ~ "Leaf 1",
+                          edge_density < 0.00590599876314162 & normalized_motifs_3 >= 0.865853658536585 ~ "Leaf 2",
+                          edge_density >= 0.00590599876314162 & normalized_motifs_3 < 0.967078189300412 ~ "Leaf 3",
+                          edge_density >= 0.00590599876314162 & normalized_motifs_3 >= 0.967078189300412 ~ "Leaf 4") ,
+         leaf = factor(leaf))
+```
+
+Join the classes to walksheds:
+```{r}
+walksheds_train <- walksheds_train |> 
+  left_join(d |>
+              st_drop_geometry() |>
+              select(GeoUID,
+                     Type,
+                     leaf),
+            by = c("GeoUID", 
+                   "Type"))
+```
+
+Examples of networks and walksheds by leaf.
+
+Leaf 1 (Urban):
+```{r plot-leaf-1}
+set.seed(29919)
+
+walkshed_leaf_1 <- walksheds_train |> 
+  ungroup () |>
+  filter(leaf == "Leaf 1") |>
+  slice_sample(n = 1)
+
+leaf_1_plot <- hamilton_net$edges |>
+  st_intersection(walkshed_leaf_1)
+
+leaf_1_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_1_plot$edge_index) |>
+  st_transform(crs = st_crs(urban_types))
+
+walkshed_leaf_1 <- walkshed_leaf_1 |> 
+  st_transform(crs = st_crs(urban_types))
+
+# Find the center of the bounding box; we'll use it to plot on a false origin and fixed extent
+max_x_1 <- st_bbox(leaf_1_plot)[3] - st_bbox(leaf_1_plot)[1]
+max_y_1 <- st_bbox(leaf_1_plot)[4] - st_bbox(leaf_1_plot)[2]
+center_1 <- c(max_x_1/2 + st_bbox(leaf_1_plot)[1], max_y_1/2 + st_bbox(leaf_1_plot)[2])
+
+ggplot() +
+  geom_sf(data = walkshed_leaf_1,
+          color = "#FDAE61",
+          fill = "#FDAE61") +
+  geom_sf(data = leaf_1_plot) +
+  ggtitle("Walkshed in Leaf 1 (Suburban)") +
+  theme_void()
+```
+
+Leaf 2 (Urban):
+```{r plot-leaf-2}
+set.seed(29919)
+
+walkshed_leaf_2 <- walksheds_train |> 
+  ungroup () |>
+  filter(leaf == "Leaf 2") |>
+  slice_sample(n = 1)
+
+leaf_2_plot <- hamilton_net$edges |>
+  st_intersection(walkshed_leaf_2)
+
+leaf_2_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_2_plot$edge_index) |>
+  st_transform(crs = st_crs(urban_types))
+
+walkshed_leaf_2 <- walkshed_leaf_2 |> 
+  st_transform(crs = st_crs(urban_types))
+
+# Find the center of the bounding box; we'll use it to plot on a false origin and fixed extent
+max_x_2 <- st_bbox(leaf_2_plot)[3] - st_bbox(leaf_2_plot)[1]
+max_y_2 <- st_bbox(leaf_2_plot)[4] - st_bbox(leaf_2_plot)[2]
+center_2 <- c(max_x_2/2 + st_bbox(leaf_2_plot)[1], max_y_2/2 + st_bbox(leaf_2_plot)[2])
+
+ggplot() +
+  geom_sf(data = walkshed_leaf_2,
+          color = "#1A9850",
+          fill = "#1A9850") +
+  geom_sf(data = leaf_2_plot) +
+  ggtitle("Walkshed in Leaf 2 (Urban)") +
+  theme_void()
+```
+
+Leaf 3 (Urban):
+```{r plot-leaf-3}
+set.seed(29922)
+
+walkshed_leaf_3 <- walksheds_train |> 
+  ungroup () |>
+  filter(leaf == "Leaf 3") |>
+  slice_sample(n = 1)
+
+leaf_3_plot <- hamilton_net$edges |>
+  st_intersection(walkshed_leaf_3)
+
+leaf_3_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_3_plot$edge_index) |>
+  st_transform(crs = st_crs(urban_types))
+
+walkshed_leaf_3 <- walkshed_leaf_3 |> 
+  st_transform(crs = st_crs(urban_types))
+
+# Find the center of the bounding box; we'll use it to plot on a false origin and fixed extent
+max_x_3 <- st_bbox(leaf_3_plot)[3] - st_bbox(leaf_3_plot)[1]
+max_y_3 <- st_bbox(leaf_3_plot)[4] - st_bbox(leaf_3_plot)[2]
+center_3 <- c(max_x_3/2 + st_bbox(leaf_3_plot)[1], max_y_3/2 + st_bbox(leaf_3_plot)[2])
+
+ggplot() +
+  geom_sf(data = walkshed_leaf_3,
+          color = "#FDAE61",
+          fill = "#FDAE61") +
+  geom_sf(data = leaf_3_plot) +
+  ggtitle("Walkshed in Leaf 3 (Suburban)") +
+  theme_void()
+```
+
+Leaf 4 (Urban):
+```{r plot-leaf-4}
+set.seed(29946765)
+
+walkshed_leaf_4 <- walksheds_train |> 
+  ungroup () |>
+  filter(leaf == "Leaf 4") |>
+  slice_sample(n = 1)
+
+leaf_4_plot <- hamilton_net$edges |>
+  st_intersection(walkshed_leaf_4)
+
+leaf_4_plot <- hamilton_net$edges |>
+  filter(edge_index %in% leaf_4_plot$edge_index) |>
+  st_transform(crs = st_crs(urban_types))
+
+walkshed_leaf_4 <- walkshed_leaf_4 |> 
+  st_transform(crs = st_crs(urban_types))
+
+# Find the center of the bounding box; we'll use it to plot on a false origin and fixed extent
+max_x_4 <- st_bbox(leaf_4_plot)[3] - st_bbox(leaf_4_plot)[1]
+max_y_4 <- st_bbox(leaf_4_plot)[4] - st_bbox(leaf_4_plot)[2]
+center_4 <- c(max_x_4/2 + st_bbox(leaf_4_plot)[1], max_y_4/2 + st_bbox(leaf_4_plot)[2])
+
+ggplot() +
+  geom_sf(data = walkshed_leaf_4,
+          color = "#1A9850",
+          fill = "#1A9850") +
+  geom_sf(data =leaf_4_plot) +
+  ggtitle("Walkshed in Leaf 4 (Typically urban)") +
+  theme_void()
+```
+
+Find the max extent in x and y:
+```{r}
+max_x <- max(c(max_x_1, max_x_2, max_x_3, max_x_4))
+max_y <- max(c(max_y_1, max_y_2, max_y_3, max_y_4))
+```
+
+Function to translate the objects to center them at a false origin of zero:
+```{r}
+
+leaf_1_gg <- ggplot() +
+  geom_sf(data = walkshed_leaf_1,
+          color = "#FDAE61",
+          fill = "#FDAE61") +
+  geom_sf(data = leaf_1_plot) +
+  theme_void()
+
+leaf_2_gg <- ggplot() +
+  geom_sf(data = walkshed_leaf_2,
+          color = "#1A9850",
+          fill = "#1A9850") +
+  geom_sf(data = leaf_2_plot) +
+  theme_void()
+
+leaf_3_gg <- ggplot() +
+  geom_sf(data = walkshed_leaf_3,
+          color = "#FDAE61",
+          fill = "#FDAE61") +
+  geom_sf(data = leaf_3_plot) +
+  theme_void()
+
+leaf_4_gg <- ggplot() +
+  geom_sf(data = walkshed_leaf_4,
+          color = "#1A9850",
+          fill = "#1A9850") +
+  geom_sf(data =leaf_4_plot) +
+  theme_void()
 
@@
 
-All trees have similar fit.
+```{r}
+sample_sheds_by_urban_type <- cowplot::plot_grid(leaf_1_gg, leaf_2_gg, leaf_3_gg, leaf_4_gg,
+                   nrow = 2, 
+                   ncol = 2,
+                   labels = c("Leaf 1 (Subrban)", "Leaf 2 (Urban)", "Leaf 3 (Suburban)", "Leaf 4 (Urban)"))
 
-Plot the trees:
+cowplot::save_plot(sample_sheds_by_urban_type,
+                   filename = glue::glue(here::here(),"/plots/sample_sheds_by_urban_type.png"),
+                   nrow = 2,
+                   ncol = 3,
+                   base_width = 3.5)
 
-```{r}
-tree1_plot <- ggparty(tree1) +
-  geom_edge() +
-  geom_edge_label(aes(label = breaks_label)) +
-  geom_node_label(aes(label = splitvar), ids = "inner") +
-  # identical to  geom_node_splitvar() +
-  geom_node_label(aes(label = info), ids = "terminal")
-
-tree2_plot <- ggparty(tree2) +
-  geom_edge() +
-  geom_edge_label(aes(label = breaks_label)) +
-  geom_node_label(aes(label = splitvar), ids = "inner") +
-  # identical to  geom_node_splitvar() +
-  geom_node_label(aes(label = info), ids = "terminal")
-
-tree3_plot <- ggparty(tree3) +
-  geom_edge() +
-  geom_edge_label(aes(label = breaks_label)) +
-  geom_node_label(aes(label = splitvar), ids = "inner") +
-  # identical to  geom_node_splitvar() +
-  geom_node_label(aes(label = info), ids = "terminal")
-
-tree4_plot <- ggparty(tree4) +
-  geom_edge() +
-  geom_edge_label(aes(label = breaks_label)) +
-  geom_node_label(aes(label = splitvar), ids = "inner") +
-  # identical to  geom_node_splitvar() +
-  geom_node_label(aes(label = info), ids = "terminal")
-
-(tree1_plot + tree2_plot) / (tree3_plot + tree4_plot)
+sample_sheds_by_urban_type
 ```
-
-Notice that edge_density and normalized_motifs_3 are the dominant attributes in the classification trees.
-Trees 1, 3, and 4 have the same general structure, and in fact trees 3 and 4 are identical.
-Based on this we choose to use `tree3`.
-
-Check the break labels of `tree3` to round:
-
-```{r}
-ggparty(tree3)$data$breaks_label
-```
-
-```{r add-classification-to-data}
-#| include: false
-
-d <- d |>
-  mutate(leaf = case_when(edge_density < 0.00591859212548868 & normalized_motifs_3 < 0.86734693877551 ~ "Leaf 1",
-                          edge_density < 0.00591859212548868 & normalized_motifs_3 >= 0.86734693877551 ~ "Leaf 2",
-                          edge_density >= 0.00591859212548868 & normalized_motifs_3 < 0.967741935483871 ~ "Leaf 3",
-                          edge_density >= 0.00591859212548868 & normalized_motifs_3 >= 0.967741935483871 ~ "Leaf 4"),
-         leaf = factor(leaf))
-```
-
-```{r join-area-of-ped-sheds}
-#| include: false
-
-d <- d |>
-  left_join(walksheds_da |>
-              mutate(area = st_area(geometry),
-                     area = units::set_units(area, km^2) |> 
-                       drop_units()) |>
-              st_drop_geometry(),
-            by = c("GeoUID", "Type")) 
-```
-
-```{r area-of-ped-sheds-by-classification}
-#| echo: false
-
-d |>
-  select(area, leaf) |>
-  group_by(leaf) |>
-  skim() |>
-  mutate(across(numeric.mean:numeric.p100, 
-                      ~round(.x, 2)),
-         Type = c("Typically suburban", "Typically urban", "Typically suburban", "Typically urban")) |>
-  select(leaf, Type, starts_with("numeric")) |>
-  rename_with(~ str_replace(.x, "numeric.", ""), starts_with("numeric")) |>
-  gt() |>
-  cols_width(everything() ~ px(100)) |>
-  tab_options(table.font.size = 30)
-```
-
-Typically suburban networks contain many unique patterns (few motifs).
-
-0.00591859212548868 & normalized_motifs_3 \< 0.86734693877551 \~ "Leaf 1", edge_density \< 0.00591859212548868 & normalized_motifs_3 \>= 0.86734693877551 \~ "Leaf 2", edge_density \>= 0.00591859212548868 & normalized_motifs_3 \< 0.967741935483871 \~ "Leaf 3",
-
-```{r plot-leaf-1}
-set.seed(29919)
-
-leaf_1_plot <- st_intersection(hamilton_net$edges,
-             walksheds_da |> 
-  filter(GeoUID == (walksheds_da_net_vars |> 
-                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
-                      filter(edge_density < 0.00591859212548868,
-                             normalized_motifs_3 < 0.86734693877551, 
-                             Type == "Suburban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID))))
-
-leaf_1_plot <- hamilton_net$edges |>
-  filter(edge_index %in% leaf_1_plot$edge_index)
-
-#leaf_1_plot <- 
-  leaf_1_plot |>
-  ggplot() +
-  geom_sf() +
-  ggtitle("edge density < 0.006, motifs < 0.867") +
-  theme_void()
-```
-
-```{r plot-leaf-2}
-set.seed(84451)
-
-leaf_2_plot <- st_intersection(hamilton_net$edges,
-             walksheds_da |> 
-  filter(GeoUID == (walksheds_da_net_vars |> 
-                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
-                      filter(edge_density >= 0.00591859212548868,
-                             normalized_motifs_3 < 0.967741935483871, 
-                             Type == "Suburban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID))))
-
-leaf_2_plot <- hamilton_net$edges |>
-  filter(edge_index %in% leaf_2_plot$edge_index)
-
-leaf_2_plot |>
-  ggplot() +
-  geom_sf() +
-  ggtitle("edge density < 0.006, motifs >= 0.968") +
-  theme_void()
-```
-
-```{r display-suburban-networks}
-#| echo: false
-
-leaf_1_plot + plot_spacer() + leaf_3_plot
-```
-
-Typically urban networks contain isomorphic patterns that repeat (large number of motifs).
-
-```{r plot-leaf-3}
-set.seed(92348)
-
-leaf_1_plot <- st_intersection(hamilton_net$edges,
-             walksheds_da |> 
-  filter(GeoUID == (walksheds_da_net_vars |> 
-                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
-                      filter(edge_density < 0.00591859212548868,
-                             normalized_motifs_3 >= 0.86734693877551, 
-                             Type == "Urban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID))))
-
-leaf_1_plot <- hamilton_net$edges |>
-  filter(edge_index %in% leaf_1_plot$edge_index)
-
-#leaf_1_plot <- 
-  leaf_1_plot |>
-  ggplot() +
-  geom_sf() +
-  ggtitle("edge density < 0.006, motifs >= 0.867") +
-  theme_void()
-```
-
-```{r plot-leaf-4}
-set.seed(392847)
-
-leaf_4_plot <- st_intersection(hamilton_net$edges,
-             walksheds_da |> 
-  filter(GeoUID == (walksheds_da_net_vars |> 
-                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
-                      filter(edge_density >= 0.00591859212548868,
-                             normalized_motifs_3 >= 0.967741935483871, 
-                             Type == "Urban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID))))
-
-leaf_4_plot <- hamilton_net$edges |>
-  filter(edge_index %in% leaf_4_plot$edge_index)
-
-leaf_4_plot |>
-  ggplot() +
-  geom_sf() +
-  ggtitle("edge density < 0.006, motifs >= 0.968") +
-  theme_void()
-```
-
-```{r display-urban-networks}
-#| echo: false
-
-leaf_2_plot + plot_spacer() + leaf_4_plot
-```
-
-## Analysis of pedshed areas as a function of network attributes
-
-```{r estimate-model-area}
-#| include: false
-mdl <- d |>
-  mutate(area = log(area),
-         edge_density_2 = edge_density^2,
-         normalized_motifs_3_2 = normalized_motifs_3^2) |>
-  lm(area ~ edge_density + edge_density_2 + normalized_motifs_3, data = _)
-```
-
-```{r tidy-results-of-model-area}
-#| include: false
-
-mdl_tidy <- mdl |>
-  tidy() |>
-  mutate(term = c("intercept", "edge_density", "edge_density^2", "normalized_motifs_3"))
-```
-
-```{r plot-effects-transitivity}
-#| include: false
-
-b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
-b2 <- mdl_tidy |> filter(term == "edge_density") |> pull(estimate)
-b3 <- mdl_tidy |> filter(term == "edge_density^2") |> pull(estimate)
-f_transitivity <- function(x) exp(b1 + b2 * x +  b3 * x^2)
-
-effect_transitivity <- ggplot() +
-  geom_function(fun = f_transitivity,
-                xlim = c(0, 0.4)) +
-    ylim(c(0.05, 1.9)) +
-  ggtitle("Expected area as a function of transitivity") + 
-  ylab("area (sq.km)") +
-  xlab("transitivity") +
-  theme_minimal()
-```
-
-```{r plot-effect-motifs}
-#| include: false
-
-b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
-b4 <- mdl_tidy |> filter(term == "normalized_motifs_3") |> pull(estimate)
-f_motifs <- function(x) exp(b1 + b4 * x)
-
-effect_motifs <- ggplot() +
-  geom_function(fun = f_motifs,
-                xlim = c(0.45, 1.1)) +
-  ylim(c(0.05, 1.9)) +
-  ggtitle("Expected area as a function of normalized motifs (size = 3)") + 
-  ylab("area (sq.km)") +
-  xlab("normalized motifs (size = 3)") +
-  theme_minimal()
-```
-
-```{r display-model-effects}
-#| echo: false
-
-effect_transitivity + effect_motifs
-```
-
-\<!-- Check the break labels of `tree3` to round:
-
-```{r}
-ggparty(tree3)$data$breaks_label
-```
-
-Plot the tree with the rounded break labels:
-
-```{r}
-ggparty(tree3) +
-  geom_edge() +
-  geom_edge_label(aes(label = str_replace(breaks_label, 
-                                          "0.0207311636201516", 
-                                          "0.021"))) +
-  geom_node_splitvar(aes(label = ifelse(kids == 0,
-                                        glue(" n = {nodesize}"), 
-                                        splitvar)),
-                     nudge_x = 0.0,
-                     ids = c(1, 2, 5)) +
-  geom_node_splitvar(aes(label = ifelse(kids == 0,
-                                        glue(" n = {nodesize}"), 
-                                        splitvar)),
-                     nudge_x = 0.02,
-                     ids = c(3, 4, 6, 7)) +
-  geom_node_plot(gglist = list(geom_bar(aes(x = "", 
-                                            fill = Type),
-                                        position = position_fill()),
-                               xlab("Type"),
-                               theme_minimal()),
-                 # draw only one label for each axis
-                 shared_axis_labels = TRUE,
-                 # draw line between tree and legend
-                 legend_separator = FALSE
-                 )
-```
-
-Find examples of ped sheds in each of the leafs and plot the network for illustration.
-
-Leaf 1:
-
-```{r}
-
-seed <- sample.int(100000, 1)
-set.seed(seed)
-seed
-#69717
-
-st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(transitivity < 0.0207311636201516,
-                             motifs < 378, 
-                             Type == "Suburban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
-  ggplot() +
-  geom_sf()
-```
-
-Leaf 2 (urban):
-
-```{r}
-seed <- sample.int(100000, 1)
-set.seed(seed)
-seed
-#84451, 44785
-
-st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(motifs >= 378, 
-                             transitivity < 0.0207311636201516,
-                             Type == "Urban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
-  ggplot() +
-  geom_sf()
-```
-
-Leaf 3 (suburban):
-
-```{r}
-seed <- sample.int(100000, 1)
-set.seed(seed)
-seed
-#81461, 84451
-
-st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(motifs < 1024, 
-                             transitivity >= 0.0207311636201516,
-                             Type == "Suburban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
-  ggplot() +
-  geom_sf()
-```
-
-Leaf 4 (urban):
-
-```{r}
-seed <- sample.int(100000, 1)
-set.seed(seed)
-seed
-#52534
-
-st_intersection(hamilton_net$edges,
-             walksheds |> 
-  filter(GeoUID == (walksheds_net_attributes |> 
-                      filter(motifs >= 1024, 
-                             transitivity >= 0.0207311636201516,
-                             Type == "Urban") |> 
-                      slice_sample(n=1) |> 
-                      pull(GeoUID)))) |>
-  ggplot() +
-  geom_sf()
-```
-
-Add leaf labels to data:
-
-```{r}
-d <- d |>
-  mutate(leaf = case_when(transitivity < 0.0207311636201516 & motifs < 378 ~ "Leaf 1",
-                          transitivity < 0.0207311636201516 & motifs >= 378 ~ "Leaf 2",
-                          transitivity >= 0.0207311636201516 & motifs < 1024 ~ "Leaf 3",
-                          transitivity >= 0.0207311636201516 & motifs >= 1024 ~ "Leaf 4"),
-         leaf = factor(leaf))
-```
-
-```{r}
-skim(d)
-```
-
-```{r}
-d |>
-  left_join(walksheds |>
-              select(-Type) |>
-              mutate(area = st_area(geometry),
-                     area = units::set_units(area, km^2) |> 
-                       drop_units(),
-                     area = log(area)) |>
-              st_drop_geometry(),
-            by = "GeoUID") |>
-  lm(area ~ leaf, data = _) |>
-  summary()
-```
-
-But what about accessibility??
-
-Summary of amenity categories by DA and Type:
-
-```{r}
-amenities_walksheds_summary <- amenities_walksheds |>
-  drop_na() |>
-  st_drop_geometry() |>
-  group_by(GeoUID,
-           Type,
-           Category) |>
-  summarize(n = n(),
-            .groups = "drop") |>
-  pivot_wider(id_cols = c(GeoUID, Type),
-              names_from = Category,
-              values_from = n,
-              values_fill = 0)
-```
-
-To map accessibility by category of amenity we first categorize the number of amenities:
-
-```{r}
-amentities_da <-  urban_hammer_da |>
-  select(GeoUID, Type.1) |>
-  rename(Type = Type.1) |>
-  left_join(amenities_walksheds_summary,
-            by = c("GeoUID", "Type")) |>
-  mutate(across(Sustenance:Library, ~ replace_na(.x, 0))) |>
-  pivot_longer(cols = -c(GeoUID, Type, geometry),
-               names_to = "Amenity_Category",
-               values_to = "Number_of_Amenities") |>
-  #mutate(Number_of_Amenities = ifelse(Number_of_Amenities == 0, NA, Number_of_Amenities)) |>
-  mutate(Number_of_Amenities = case_when(Number_of_Amenities == 0 ~ "0",
-                                         Number_of_Amenities > 0 & Number_of_Amenities <= 3 ~ "1-3",
-                                         Number_of_Amenities > 3 & Number_of_Amenities <= 6 ~ "4-6",
-                                         Number_of_Amenities > 6 & Number_of_Amenities <= 9 ~ "7-9",
-                                         Number_of_Amenities > 9 ~ "10 or more"),
-         Number_of_Amenities = factor(Number_of_Amenities,
-                                      levels = c("0", "1-3", "4-6", "7-9", "10 or more"),
-                                      ordered = TRUE))
-```
-
-Plot:
-
-```{r}
-amentities_da |>
-  ggplot() +
-  geom_sf(aes(fill = Number_of_Amenities)) +
-  geom_sf(data = urban_types |>
-            filter(Type != "Rural"),
-          aes(color = Type),
-          fill = NA) +
-  #scale_fill_fermenter(direction = 1, palette = "Greens") +
-  facet_wrap(~ Amenity_Category) + 
-  theme_void()
-```
-
-Pivot wider:
-
-```{r}
-junk <- amentities_da |>
-  pivot_wider(names_from = "Amenity_Category",
-              values_from = "Number_of_Amenities")
-```
-
-```{r}
-junk2 <- d |>
-  left_join(junk |>
-              #select(-Type) |>
-              st_drop_geometry(),
-            by = c("GeoUID", "Type")) |>
-  mutate(across(Sustenance:Library, ~forcats::fct_drop(.x)))
-```
-
-```{r}
-summary(junk2 |> select(Sustenance:Library))
-```
-
-## Clusters: try self-organizing maps (SOM)
-
-```{r}
-wsn.som <- trainSOM(x.data = walksheds_net_attributes |>
-                       select(-c(GeoUID, Type)),
-                     dimension = c(5,5), 
-                     verbose = TRUE, 
-                     nb.save = 5, 
-                     topo = "hexagonal")
-```
-
-```{r}
-wsn.som
-```
-
-```{r}
-plot(wsn.som, what="energy")
-```
-
-```{r}
-wsn.som$clustering
-```
-
-```{r}
-wsn.som$clustering |> table()
-```
-
-```{r}
-plot(wsn.som, what="obs", type="hitmap")
-```
-
-```{r}
-summary(wsn.som)
-```
-
-```{r}
-par(mfrow = c(2,2))
-plot(wsn.som, what = "obs", type = "color", variable = 1)
-plot(wsn.som, what = "obs", type = "color", variable = 2)
-```
-
-```{r}
-plot(wsn.som, what = "add", type = "pie", variable = walksheds_net_attributes$Type) +
-  scale_fill_brewer(type = "qual") + 
-  guides(fill = guide_legend(title = "Type"))
-```
-
-```{r}
-quality(wsn.som)
-```
-
-```{r}
-plot(superClass(wsn.som))
-```
-
-```{r}
-wsn.som <- trainSOM(x.data = walksheds_net_attributes |>
-                       select(-c(GeoUID, Type)),
-                     dimension = c(4, 4), #2,3
-                     verbose = TRUE, 
-                     nb.save = 5,
-                     topo = "square")
-```
-
-```{r}
-plot(wsn.som, what="energy")
-```
-
-```{r}
-quality(wsn.som)
-```
